---
title: "R Notebook"
output: html_notebook
---

## set-up

```{r}
rm(list = ls())

library(ggplot2)

library(scico)

library(reshape2)

library(tidyverse)

library(vroom)

library(data.table)

today <- Sys.Date()

# global ggplot2 theme
theme_set(
  theme_minimal() +
    theme(
      text = element_text(
        size = 16,
        family = "Helvetica",
        colour = "black"
      ),
      axis.text = element_text(colour = "black"),
      plot.title = element_text(hjust = 0.5),
      strip.text = element_text(colour = "black")
    )
)
```

## functions

```{r}
# visualize var-cov matrix
get_upper_tri <- function(cormat){
    cormat[lower.tri(cormat)]<- NA
    diag(cormat) <- NA
    return(cormat)
}

plot_var_cov_matrix <- function(covmat, output_name){
  # get just upper triangle
  upper_tri <- get_upper_tri(covmat)
  melted_cormat <- melt(upper_tri, na.rm = TRUE)
  # Heatmap
  plot_cor <- ggplot(data = melted_cormat, aes(Var2, Var1, fill = value)) +
    geom_tile(color = "white") +
    #geom_text(aes(label = round(value, 4)), size = 5, color = "black") +
    scale_fill_gradient2(
      low = "#65014B",
      mid = "#F5F0F0",
      high = "#0C4C00",
      midpoint = 0,
      space = "Lab",
      name = "(\u0394pi, \u0394pj)"
    ) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(
        angle = 45,
        vjust = 1,
        hjust = 1
      ),
      text = element_text(size = 14, family = "Helvetica")
    ) +
    coord_fixed() +
    labs(x = "", y = "")
  
  ggsave(output_name,
         plot_cor,
         bg = "white",
         device = cairo_pdf)
}

# buffalo coop gt curves
gt_from_pmat <- function(dps, calcabs = FALSE, output_name) {
  # sanity checks
  
  # loop over replicates
  # create matrix of allele frequncy increments
  # calculate cov matrix for increments
  results = NULL
  col_names <- colnames(dps[[1]])
  for (i in 1:ncol(dps[[1]])) {
    print(i)
    dp = NULL
    for (k in 1:length(dps)) {
      print(k)
      dp <- cbind(dp, (dps[[k]])[,i])
    }
    
    # covariance matrix
    print("Calculating covariance...")
    covmat <- cov(dp)
    covmat[lower.tri(covmat)] <- NA
    
    # plot variance covariance matrix
    plot_var_cov_matrix(covmat, paste(output_name, "_", i, ".pdf", sep = ""))
    
    # variances and covar
    print("Extracting covariances and variances...")
    covariances <- covmat[upper.tri(covmat)]
    variances <- diag(covmat)
    
    # cum sum of variances
    print("Cumulative sum of variances...")
    sums_var <- cumsum(variances)
    
    # cumulative sum of covariances
    print("Cumulative sum of covariances...")
    sums_cov = c(0)
    sums_abs_cov = c(0)
    sums_pos_cov = c(0)
    sums_neg_cov = c(0)
    for (j in 2:nrow(covmat)) {
      print(j)
      cov_sub <- covariances[1:(((j - 1) * (j)) / 2)]
      
      sum_abs_cov <- 2 * sum(abs(cov_sub), na.rm = T)
      sum_cov <- 2 * sum(cov_sub, na.rm = T)
      sum_neg_cov <- 2 * sum(abs(cov_sub[(cov_sub < 0)]), na.rm = T)
      sum_pos_cov <- 2 * sum(cov_sub[(cov_sub > 0)], na.rm = T)
      
      sums_cov = c(sums_cov, sum_cov)
      sums_abs_cov = c(sums_abs_cov, sum_abs_cov)
      sums_pos_cov = c(sums_pos_cov, sum_pos_cov)
      sums_neg_cov = c(sums_neg_cov, sum_neg_cov)
    }
    
    # calculate gt
    print("Calculating gt...")
    if (calcabs == TRUE) {
      gts = sums_abs_cov / (sums_var + sums_abs_cov)
    } else {
      gts = sums_cov / (sums_var + sums_cov)
    }
    
    print("Saving result...")
    result = data.frame(
      gen = 1:nrow(covmat),
      gt = gts,
      sum_cov = sums_cov,
      sum_abs_cov = sums_abs_cov,
      sum_pos_cov = sums_pos_cov,
      sum_neg_cov = sums_neg_cov,
      ratio = sums_neg_cov / (sums_pos_cov + sums_neg_cov),
      sum_var = sums_var,
      replicate = i
    )
    
    results = rbind(results, result)
  }
  return(results)
}

## buffalo gt curves
gt_from_covmat <- function(covmat) {
  covmat[lower.tri(covmat)] <- NA
  
  covariances <- covmat[upper.tri(covmat)]
  variances <- diag(covmat)
  
  # add up variances
  sums_var <- cumsum(variances)

  # cumulative sum of covariances
  sums_cov = c(0)
  sums_abs_cov = c(0)
  sums_pos_cov = c(0)
  sums_neg_cov = c(0)
  for(j in 2:nrow(covmat)){
    cov_sub <- covariances[1:(((j-1)*(j))/2)]
    
    sum_abs_cov <- 2*sum(abs(cov_sub), na.rm = T)
    sum_cov <- 2*sum(cov_sub, na.rm = T)
    sum_neg_cov <- 2*sum(abs(cov_sub[(cov_sub < 0)]), na.rm = T)
    sum_pos_cov <- 2*sum(cov_sub[(cov_sub > 0)], na.rm = T)
    
    sums_cov = c(sums_cov, sum_cov)
    sums_abs_cov = c(sums_abs_cov, sum_abs_cov)
    sums_pos_cov = c(sums_pos_cov, sum_pos_cov)
    sums_neg_cov = c(sums_neg_cov, sum_neg_cov)
  }
  
  # calculate gt
#   if (calcabs == TRUE) {
#     gts = sums_abs_cov/(sums_var + sums_abs_cov)
#   } else {
# 	  gts = sums_cov/(sums_var + sums_cov)
#   }
  
  result = data.frame(
    gen = 1:nrow(covmat),
    gt = sums_cov/(sums_var + sums_cov),
    gt_abs = sums_abs_cov/(sums_var + sums_abs_cov),
    sum_cov = sums_cov,
    sum_abs_cov = sums_abs_cov,
    sum_pos_cov = sums_pos_cov,
    sum_neg_cov = sums_neg_cov,
    sum_var = sums_var,
    ratio = sums_pos_cov/(sums_pos_cov + sums_neg_cov)
  )
  
  return(result)
}

## manhattan plots
# https://r-graph-gallery.com/101_Manhattan_plot.html
# https://danielroelfs.com/posts/how-i-create-manhattan-plots-using-ggplot/
plot_manhattan_by_site <- function(data, site, today, height, width, quick = T, prefix){
  
  print(site)
  
  print("Filter data to one site...")
  data <- data |>
    subset(SITE == site)
  
  if (quick){
    print("Subset insignificant data...") 
    data <- data |>
      subset(pvalue < 0.001)
  }
  
  print("Compute variables...")
  don <- data %>% 
    # Compute chromosome size
    group_by(CHROM) %>% 
    summarise(chr_len=max(POS)) %>%
    # Calculate cumulative position of each chromosome
    mutate(tot=cumsum(chr_len)-chr_len) %>%
    select(-chr_len) %>%
    # Add this info to the initial dataset
    left_join(data, ., by=c("CHROM"="CHROM")) %>%
    # Add a cumulative position of each SNP
    arrange(CHROM, POS) %>%
    mutate(BPcum=POS+tot)
  # Add highlight and annotation information
  #mutate(is_highlight=ifelse((CHROM == chrom) & (POS > start) & (POS < end), "yes", "no"))
  
  axisdf = don %>%
    group_by(CHROM) %>%
    summarize(center=( max(BPcum) + min(BPcum) ) / 2 )
  
  print("Plot...")
  ggplot(don, aes(x=BPcum, y=-log10(pvalue))) +
    # Show all points
    geom_point( aes(color=as.factor(CHROM)), alpha=0.8, size=1.3, shape = 16) +
    # Add highlighted points
    geom_point(data=subset(don, highlight=="yes"), color="orange", size=1.5) +
    scale_color_manual(values = rep(c("grey", "black"), 22 )) +
    # custom X axis:
    scale_x_continuous( label = axisdf$CHROM, breaks= axisdf$center ) +
    scale_y_continuous(expand = c(0, 0) ) +     # remove space between plot area and x axis
    # Custom the theme:
    #theme_bw() +
    theme( 
      legend.position="none",
      panel.border = element_blank(),
      panel.grid.major.x = element_blank(),
      panel.grid.minor.x = element_blank()
    ) +
    labs(x = "Chromosome", y = "-log10(p)", title = paste("Site #", site, sep = " "))
  
  # ggplot(subset(gwas_data, SITE == site), aes(x = POS, y = -log10(pvalue))) +
  #   geom_point(shape = 16) +
  #   facet_wrap(~CHROM, ncol = 5, labeller = "label_both", scales = "free_x") +
  #   theme(axis.text.x = element_text(angle = 45, hjust=1)) +
  #   labs(x = "Position", y = "-log10(p-value)")
  
  print("Save pdf...")
  ggsave(paste("../results/", today, "/", prefix, "_", site, ".pdf", sep = ""), height = height, width = width)
  print("Save png...")
  ggsave(paste("../results/", today, "/", prefix, "_", site, ".jpg", sep = ""), height = height, width = width)
}


```


## Conceptual images

```{r}
# functions from: https://si.biostat.washington.edu/sites/default/files/modules/Day2_AM_part1.pdf
fitfreq = function(q, h, s){
  p=1-q;
  return((q^2*(1+s) + p*q*(1+h*s))/( 1 + s*q*(2*h*p+q)))
}

WF.sel=function(N, q, h, s, G){
  t=array(,dim=G)
  t[1] = N*q
  for(i in 2:G){
    t[i] = rbinom(1,N,fitfreq(t[i-1]/N, h, s))
  }
  return(t)
}

gen = 100
N = 100
h = 0.5
s = 0.05
q0 = 0.5

selection = WF.sel(N, q0, h, s, gen)

drift = WF.sel(N, q0, h, 0, gen)

plotdata <- data.frame(
  gen = 0:gen,
  selection = c(q0*N, selection)/N,
  drift = c(q0*N, neutral)/N
)

plotdata <- melt(plotdata, id = "gen")

ggplot(plotdata, aes(x = gen, y = value)) +
  geom_abline(slope = 0, intercept = q0, lty = 2) +
  geom_point() +
  geom_line() +
  facet_wrap(~variable) +
  labs(x = "Generation", y = "Allele frequency")

ggsave(paste("../results/", today, "/allele_frequency_changes.pdf", sep = ""), height = 4, width = 7)

plotdata2 <- data.frame(
  gen = 0:gen,
  n1 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n2 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n3 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n4 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n5 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n6 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n7 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n8 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n9 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n10=  c(q0, WF.sel(N, q0, h, 0, gen)/N)
)

plotdata2 <- melt(plotdata2, id = "gen")

ggplot(plotdata2, aes(x = gen, y = value, group = variable)) +
  #geom_abline(slope = 0, intercept = q0, lty = 2) +
  geom_line(lwd = 2, color = "green") +
  theme_void() +
  labs(x = "Generation", y = "Allele frequency")

ggsave(paste("../results/", today, "/neutral_examples.pdf", sep = ""), height = 4, width = 7)

```

## Manhattan plots of sweep statistics in founders

```{r}

sweeps <- read.table("../data/sweeps.txt", header = T, sep = "\t")

sweeps <- melt(sweeps, id = c("Contig", "Var_Start", "Var_Stop", "Pos_Start", "Pos_Stop"))

sweeps <- sweeps[!(sweeps$variable %in% c("Theta_W", "Theta_Pi")),]

# manhattan plots of 
sweeps$variable <- gsub("_", " ", sweeps$variable)

sweeps$Contig <- as.factor(sweeps$Contig)

sweeps$Contig <- revalue(sweeps$Contig, c("0" = "1", "1" = "2", "2" = "3", "3" = "4", "4" = "5"))

ggplot(sweeps, aes(x = (Pos_Start + Pos_Stop)/2, y = value)) +
  geom_point(shape = 16) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_wrap(variable~Contig, ncol = 5, nrow = 5, scales = "free") +
  labs(x = "Position", y = "Statistic")

ggsave(paste("../results/", today, "/sweep_stats.pdf", sep = ""), height = 11, width = 13)
```

## seed mix: compare freqk estimated allele frequencies to vcf snp calls

```{r}
# freqk estimates
freqk_calls <- read.table("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/resources/calls.txt", sep = "|", header = F, col.names = c("RF_freqk", "AF_freqk"))

freqk_index <- read.table("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/resources/ref_index_greneNet.txt", sep = ",", header = F)

freqk_calls <- cbind(freqk_index, freqk_calls)

freqk_calls$V3 = freqk_calls$V3 + 2

# greneNet vcf
real_calls <- read.table("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/resources/greneNet_allele_counts.tsv", sep = "\t", header = F, col.names = c("CHROM", "POS", "REF", "ALT", "AN", "AC"))

# calculate reference allele frequency
real_calls$RF_real = 1 - real_calls$AC/real_calls$AN

# merge
names(freqk_calls)[2:3] = c("CHROM", "POS")

real_vs_freqk <- merge(real_calls, freqk_calls, by = c("CHROM", "POS"))

# plot
rho_estimate = signif(cor.test(real_vs_freqk$RF_real, real_vs_freqk$RF_freqk, method = "spearman")$estimate, 3)

ggplot(real_vs_freqk, aes(x = RF_real, y = RF_freqk)) +
  geom_bin2d() +
  #geom_smooth() +
  geom_abline(slope = 1, intercept = 0, color = "white", lwd = 2) +
  #scale_fill_continuous(trans = "log10") +
  scale_fill_scico(palette = "bilbao", trans = "log10") +
  theme_minimal() +
  theme(text = element_text(size = 16, family = "Helvetica", colour = "black"),
        axis.text.x = element_text(color = "black"),
        axis.text.y = element_text(color = "black"), 
        plot.title = element_text(hjust = 0.5)) +
  labs(x = "REF frequency in greneNet VCF", y = "REF frequency from\nk-mer counts", fill = "log10(count)", title = paste("\u03C1 = ", rho_estimate, sep = ""))

ggsave(paste("../results/", today, "/bin2d_greneNet_vs_freqk.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)
```

## randomly switch tracked allele frequencies

```{r}
set.seed(1111)

# LD-pruned data
#gen0 <- read.table("../data/average_seedmix_p0_LDpruned.txt", sep = "\t")
#hapfire <- read.table("../data/merged_hapFIRE_allele_frequency_LDpruned.txt", sep = "\t", header = T)

# full dataset
gen0 <- vroom("/global/scratch/users/milesroberts/tmp/average_seedmix_p0.txt", delim = "\t", col_names = F)

# randomly swith which allele is being tracked
# right now, the non-reference allele is always the one we're tracking
idx <- sample(c(T,F), size = nrow(gen0), replace = T)
gen0[idx,"X3"] <- 1 - gen0[idx,"X3"]

write.csv(gen0, "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/seedmix_hapfire_allele_frequency_rng_switched.csv", quote = F, row.names = F)

#hapfire_phase1 <- read_csv("/global/scratch/users/milesroberts/tmp/merged_hapFIRE_allele_frequency.csv")

# start <- Sys.time()
# hapfire_phase1 <- vroom("/global/scratch/users/milesroberts/tmp/merged_hapFIRE_allele_frequency.csv", delim = ",")
# end <- Sys.time()
# end - start
# 
# hapfire_phase2 <- vroom("/global/scratch/users/milesroberts/tmp/PHASE2_merged_hapFIRE_allele_frequency.csv", delim = ",")
# 
# hapfire <- cbind(hapfire_phase1, hapfire_phase2)
# 
# write.csv(hapfire, "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_2_hapfire_allele_frequency.csv", quote = F, row.names = F)

# randomly switch which allele is being tracked, same as gen0
# j = 0
# for(i in idx){
#   if(j %% 1000 == 0){
#     print(j)
#   }
#   hapfire[idx,] <- 1 - hapfire[idx,]
#   j = j + 1
# }

process_one_minus <- function(idx, chunk_size, output) {
  function(x, pos) {
    print(pos)
    print(Sys.time())
    # calculate end
    chunk_size <- nrow(x)
    end <- pos + chunk_size - 1
    # subtract 1 from some alleles
    print(paste("Start:", pos, "End:", end))
    idx_sub <- idx[pos:end]
    x[idx_sub,] <- 1 - x[idx_sub,]
    write.table(x, output, col.names=!file.exists(output), append = T, row.names = F, sep = ",", quote = F)
  }
}

# phase 1
read_csv_chunked(
  "/global/scratch/users/milesroberts/tmp/PHASE_1_2_COMBINED_hapFIRE_allele_frequency.csv",
  callback = SideEffectChunkCallback$new(
    process_one_minus(
      idx,
      5000,
      "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_2_COMBINED_hapFIRE_allele_frequency_switched.csv"
    )
  ),
  chunk_size = 5000
)

# phase 2
# read_csv_chunked(
#   "/global/scratch/users/milesroberts/tmp/PHASE2_merged_hapFIRE_allele_frequency.csv",
#   callback = SideEffectChunkCallback$new(
#     process_one_minus(
#       idx,
#       5000,
#       "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_2_hapfire_allele_frequency_rng_switched.csv"
#     )
#   ),
#   chunk_size = 5000
# )
```

## load allele frequencies

```{r}
#gen0 <- vroom("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/seedmix_hapfire_allele_frequency_rng_switched.csv", delim = ",", col_names = T)

gen0 <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/seedmix_hapfire_allele_frequency_rng_switched.csv", data.table = F)

#hapfire_phase1 <- vroom("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_hapfire_allele_frequency_rng_switched.csv", delim = ",")

#hapfire_phase1 <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_hapfire_allele_frequency_rng_switched.csv", data.table = F)

#hapfire_phase2 <- vroom("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_2_hapfire_allele_frequency_rng_switched.csv", delim = ",")

#hapfire_phase2 <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_2_hapfire_allele_frequency_rng_switched.csv", data.table = F)

#hapfire <- cbind(hapfire_phase1, hapfire_phase2)

#rm(hapfire_phase1, hapfire_phase2)
#gc()

hapfire <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_2_COMBINED_hapFIRE_allele_frequency_switched.csv", data.table = F)

# omit allele frequencies near abosrbing boundaries
non_abosrbing_freq <- (gen0$X3 >= 1/231) & (gen0$X3 <= 1 - 1/231)
gen0 <- gen0[non_abosrbing_freq,]
hapfire <- hapfire[non_abosrbing_freq,]
```

## feder test (LD pruned dataset)

```{r}
# subset different generations
gen1 <- hapfire[,grepl("_1_", names(hapfire))] 

gen2 <- hapfire[,grepl("_2_", names(hapfire))] 

gen3 <- hapfire[,grepl("_3_", names(hapfire))]

gen4 <- hapfire[,grepl("_4_", names(hapfire))] 

# subset to only plots present in all 3 years
shared_gen1 <- (gsub("_1_","_",names(gen1)) %in% gsub("_2_","_",names(gen2)))& (gsub("_1_","_",names(gen1)) %in% gsub("_3_","_",names(gen3))) & (gsub("_1_","_",names(gen1)) %in% gsub("_4_","_",names(gen4)))

shared_gen2 <- (gsub("_2_","_",names(gen2)) %in% gsub("_3_","_",names(gen3))) & (gsub("_2_","_",names(gen2)) %in% gsub("_1_","_",names(gen1))) & (gsub("_2_","_",names(gen2)) %in% gsub("_4_","_",names(gen4)))

shared_gen3 <- (gsub("_3_","_",names(gen3)) %in% gsub("_2_","_",names(gen2))) & (gsub("_3_","_",names(gen3)) %in% gsub("_1_","_",names(gen1))) & (gsub("_3_","_",names(gen3)) %in% gsub("_4_","_",names(gen4)))

shared_gen4 <- (gsub("_4_","_",names(gen4)) %in% gsub("_2_","_",names(gen2))) & (gsub("_4_","_",names(gen4)) %in% gsub("_1_","_",names(gen1))) & (gsub("_4_","_",names(gen4)) %in% gsub("_3_","_",names(gen3)))

gen1_sub <- gen1[,shared_gen1]

gen2_sub <- gen2[,shared_gen2]

gen3_sub <- gen3[,shared_gen3]

gen4_sub <- gen4[,shared_gen4]

# sort columns to match
gen4_sub <- gen4_sub[,paste("1_4", c(1,4,5,6,7,8,9,10,11,12), sep = "_")]

# calculate statistic
y1 <- (gen1_sub - gen0$V3)/sqrt(2*gen0$V3*(1-gen0$V3)*(1-0))
y1$gen <- 1
y1$CHROM <- gen0$V1
y1$POS <- gen0$V2
y1 <- melt(y1, id = c("gen", "CHROM", "POS"))

y2 <- (gen2_sub - gen1_sub)/sqrt(2*gen1_sub*(1-gen1_sub)*(2-1))
y2$gen <- 2
y2$CHROM <- gen0$V1
y2$POS <- gen0$V2
y2 <- melt(y2, id = c("gen", "CHROM", "POS"))

y3 <- (gen3_sub - gen2_sub)/sqrt(2*gen2_sub*(1-gen2_sub)*(3-2))
y3$gen <- 3
y3$CHROM <- gen0$V1
y3$POS <- gen0$V2
y3 <- melt(y3, id = c("gen", "CHROM", "POS"))

yvalues <- rbind(y1, y2, y3)
yvalues$SITE <- gsub("X", "", gsub("_.*", "", yvalues$variable))
yvalues$REP <- gsub(".*_", "", yvalues$variable)

t_test_stat <- yvalues %>% group_by(CHROM, POS, SITE) %>% summarize(ybar = mean(value, na.rm = T), s2 = var(value, na.rm = T), n = n())

t_test_stat$tvalue <- t_test_stat$ybar/sqrt(t_test_stat$s2/t_test_stat$n)

t_test_stat$pvalue <- dt(t_test_stat$tvalue, df = t_test_stat$n)

# Manhattan plot
map(unique(t_test_stat$SITE), ~plot_manhattan_by_site(data = t_test_stat, site = .x, today, height = 4, width = 10))

#load("../data/allele_allyears_frequencies_long_raw_climate.rda")
# calculate mean and average

# ybar <- (y1 + y2 + y3)/3
# 
# s2 <- (1/(3-1))*( (y1 - ybar)^2 + (y2 - ybar)^2 + (y3 - ybar)^2 )
# 
# t_test_stat <- ybar/sqrt(s2/3)
# 
# # 
# t_test_stat$CHROM <- gen0$V1
# 
# t_test_stat$POS <- gen0$V2
# 
# t_test_stat <- melt(t_test_stat, id = c("CHROM", "POS"))
# 
# t_test_stat$SITE <- gsub("X", "", gsub("_.*", "", t_test_stat$variable))
# 
# t_test_stat$REP <- gsub(".*_", "", t_test_stat$variable)
# 
# # average across reps of each pos at each site
# mean_t_stat <- t_test_stat %>% 
#   group_by(CHROM, POS, SITE) %>% 
#   summarize(MEANT = mean(value, na.rm = T), SDT = sd(value, na.rm = T))
# 
# # calculate p-values
# #
# mean_t_stat$PVALUE <- dt(mean_t_stat$MEANT, df = 2)
# 
# # adjust p-values
# mean_t_stat$ADJPVALUE <- p.adjust(mean_t_stat$PVALUE, method = "bonferroni")
# 
# # plot distribution of stats across positions and sites
# ggplot(mean_t_stat, aes(x = MEANT)) +
#   geom_density()
#   xlim(c(-10, 10))
#   
# # Manhattan plot
# map(unique(mean_t_stat$SITE), ~manhattan_by_site(data = mean_t_stat, site = .x, today, height = 4, width = 10))

#
#summary(unlist(as.vector(t_test_stat)))

```

## feder test (full frequency matrix)

```{r}

# Source - https://stackoverflow.com/a/33051935
# Posted by scoa
# Retrieved 2026-02-27, License - CC BY-SA 3.0
# con <- file("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/merged_hapFIRE_allele_frequency.csv","r")
# F_header <- readLines(con,n=1)
# close(con)
# 
# F_subset <- read_csv("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/merged_hapFIRE_allele_frequency.csv", col_select = "1_1_1")

gen0 <- read.table("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/average_seedmix_p0.txt", sep = "\t")

process_F <- function(gen0, chunk_size, output) {
  function(x, pos) {
    print(pos)
    print(Sys.time())
    gen0 <- gen0[pos:(pos + chunk_size - 1),]
    # generations
    header_names <- names(x)
    gen1 <- x[,grepl("_1_", header_names)] 
    gen2 <- x[,grepl("_2_", header_names)] 
    gen3 <- x[,grepl("_3_", header_names)]
    
    # subset to populations shared across generations
    names_gen1 <- names(gen1)
    names_gen2 <- names(gen2)
    names_gen3 <- names(gen3)
    
    shared_gen1 <- (gsub("_1_","_",names_gen1) %in% gsub("_2_","_",names_gen2)) & (gsub("_1_","_",names_gen1) %in% gsub("_3_","_",names(gen3)))
    shared_gen2 <- (gsub("_2_","_",names_gen2) %in% gsub("_3_","_",names_gen3)) & (gsub("_2_","_",names_gen2) %in% gsub("_1_","_",names_gen1))
    shared_gen3 <- (gsub("_3_","_",names_gen3) %in% gsub("_2_","_",names_gen2)) & (gsub("_3_","_",names_gen3) %in% gsub("_1_","_",names_gen1))
    
    gen1_sub <- gen1[,shared_gen1]
    gen2_sub <- gen2[,shared_gen2]
    gen3_sub <- gen3[,shared_gen3] 
    # calculate statistic
    y1 <- (gen1_sub - gen0$V3)/sqrt(2*gen0$V3*(1-gen0$V3)*(1-0))
    y1$gen <- 1
    y1$CHROM <- gen0$V1
    y1$POS <- gen0$V2
    y1 <- melt(y1, id = c("gen", "CHROM", "POS"))
    
    y2 <- (gen2_sub - gen1_sub)/sqrt(2*gen1_sub*(1-gen1_sub)*(2-1))
    y2$gen <- 2
    y2$CHROM <- gen0$V1
    y2$POS <- gen0$V2
    y2 <- melt(y2, id = c("gen", "CHROM", "POS"))
    
    y3 <- (gen3_sub - gen2_sub)/sqrt(2*gen2_sub*(1-gen2_sub)*(3-2))
    y3$gen <- 3
    y3$CHROM <- gen0$V1
    y3$POS <- gen0$V2
    y3 <- melt(y3, id = c("gen", "CHROM", "POS"))
    
    yvalues <- rbind(y1, y2, y3)
    yvalues$SITE <- gsub("X", "", gsub("_.*", "", yvalues$variable))
    yvalues$REP <- gsub(".*_", "", yvalues$variable)
    
    t_test_stat <- yvalues %>% group_by(CHROM, POS, SITE) %>% summarize(ybar = mean(value, na.rm = T), s2 = var(value, na.rm = T), n = n())
    
    t_test_stat$tvalue <- t_test_stat$ybar/sqrt(t_test_stat$s2/t_test_stat$n)
    
    t_test_stat$pvalue <- dt(t_test_stat$tvalue, df = t_test_stat$n)
    
    write.table(t_test_stat, output, col.names=!file.exists(output), append = T, row.names = F, sep = ",", quote = F)
  }
}

read_csv_chunked(
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/merged_hapFIRE_allele_frequency.csv",
  #callback = SideEffectChunkCallback$new(function(x, pos){print(pos)}),
  callback = SideEffectChunkCallback$new(process_F(gen0, 1000, "feder_t_test.txt")),
  chunk_size = 1000)
                

```

## plot full feder test

```{r}
# load results of test
# remove invalid points
feder <- vroom("feder_t_test.txt", delim= ",")

feder <- feder |>
  filter(is.finite(ybar))

# choose which points to highlight
# Chr1:1185512..1188696 - CRY2
# Chr1:24331373..24333999 - FT
# Chr2:11058944..11063324 - ELF3
# Chr2:11531967..11534358 - CAM5
# Chr2:18807505..18811125 - SOC1
# Chr4:9195247..9199840 - PHYD
# Chr4:11000771..11003010 - TSF
# Chr5:5170162..5173085 - CO
# Chr5:3173382..3179448 - FLC
# Chr5:18589438..18591506 - DOG1

feder$highlight <- "no"
buffer = 10000
chroms = c(1,1,2,2,2,4,4,5,5,5)
starts = c(1185512,24331373,11058944,11531967,18807505,9195247,11000771,5170162,3173382,18589438)
ends = c(1188696,24333999,11063324,11534358,18811125,9199840,11003010,5173085,3179448,18591506)
starts = starts - buffer
ends = ends + buffer

for(i in seq_along(chroms)){
  print(i)
  feder$highlight[(feder$CHROM == chroms[i]) & (feder$POS > starts[i]) & (feder$POS < ends[i])] <- "yes"
}

# test out a few plots
plot_manhattan_by_site(data = feder, site = 4, today, height = 3, width = 8, quick = F, prefix = "manhattan_flowering_time")

# plot them ALL!
map(unique(feder$SITE), ~plot_manhattan_by_site(data = feder, site = .x, today, height = 3, width = 8, quick = F, prefix = "manhattan_flowering_time"))

# drill down to certain peaks
# s10_c3 <- feder %>% subset(SITE == 10 & CHROM == 3)
# s10_c5 <- feder %>% subset(SITE == 10 & CHROM == 5)
# s32_c1 <- feder %>% subset(SITE == 32 & CHROM == 1)
```

## modify feder test

```{r}

p0 = 0.5
N0 = 500
L = 1:30
lambda = 100

truth = c(p0)
estims = c(p0)
covs = c(lambda)

# simulate 
for(i in L){
  # generate new generation
  x1 <- rbinom(1, N0, prob = p0)
  p1 <- x1/N0
  # sequencing
  coverage <- rpois(1, lambda = lambda)
  x1_hat <- rbinom(1, size = coverage, prob = p1)
  p1_hat <- x1_hat/coverage
  # save results
  truth = c(truth, p1)
  estims = c(estims, p1_hat)
  covs = c(covs, coverage)
}

results = data.frame(
  p = truth,
  p_hat = estims,
  cov = covs,
  i = c(0, L)
)

variance_table = data.frame(
  dt = diff(results$i),
  cov1 = results$cov[1:(nrow(results) - 1)],
  cov2 = results$cov[2:nrow(results)],
  dp = 2*results$p_hat[-1]*(1 - results$p_hat[-1])
)


variance_table$var = sqrt(variance_table$dt * variance_table$dp * (1/variance_table$cov1 + 1/variance_table$cov2))

z_stats <- diff(results$p_hat)/variance_table$var

hist(z_stats)

shapiro.test(z_stats)
```

## site climatic data

```{r}
#load("../data/sites.clim.rda")
load("../data/worldclim_sitesdata.rda")

#sites.clim <- sites.clim[,c("SITE_CODE", paste("bio", 1:19, sep = ""))]
```

## temporal autocovariance

```{r}

# change in allele frequency between t = 0 and t = 1
dp0 <- gen1_sub - gen0$V3
dp1 <- gen2_sub - gen1_sub
dp2 <- gen3_sub - gen2_sub
dp3 <- gen4_sub - gen3_sub

results = NULL

for(i in 1:ncol(dp0)){
  tmat <- cov(data.frame(
    dp0 = dp0[,i],
    dp1 = dp1[,i],
    dp2 = dp2[,i],
    dp3 = dp3[,i]
  ))
  
  sampleid = names(dp0)[i]
  site = gsub("X|_.*", "", sampleid)
  replicate = gsub(".*_", "", sampleid)
  
  # plot variance covariance matrix
  plot_var_cov_matrix(tmat, paste("../results/", today, "/", site, "_", replicate, "_var_cov_mat.pdf", sep = ""))
  
  result = gt_curves(tmat, calcabs = TRUE)
  
  result$site = site
  result$replicate = replicate
  
  results = rbind(results, result)
}

# average across reps in a site
mean_results = results %>% 
  group_by(site, gen) %>% 
  summarize(gt = mean(gt), 
            sum_cov = mean(sum_cov), 
            sum_var = mean(sum_var), 
            sum_abs_cov = mean(sum_abs_cov), 
            ratio = mean(ratio, na.rm = T))

# merge in climate data
mean_results <- merge(mean_results, worldclim_sitesdata, by = "site")

# plot mean of averages
ggplot(results, aes(
  x = gen,
  y = gt,
  group = interaction(site, replicate)
)) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(
    data = mean_results,
    aes(
      x = gen,
      y = gt,
      group = site,
      color = bio1
    ),
    lwd = 2,
    alpha = 1
  ) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name = "Generation", limits = c(1:4)) +
  labs(x = "Generation", y = "Proportion of allele frequency\nchange due to linked selection", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste("../results/", today, "/linked_selection.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)

# plot covariances
ggplot(results, aes(x = gen, y = sum_cov, group = interaction(site, replicate))) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(data = mean_results, aes(x = gen, y = sum_cov, group = site, color = bio1), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:3)) +
  labs(x = "Generation", y = "Sum of Cov(\u0394pi, \u0394pj)", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste("../results/", today, "/linked_selection_sum_cov.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)

# plot variances
ggplot(results, aes(x = gen, y = sum_var, group = interaction(site, replicate))) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(data = mean_results, aes(x = gen, y = sum_var, group = site, color = bio1), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:3)) +
  labs(x = "Generation", y = "Var(\u0394pi)", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste("../results/", today, "/linked_selection_sum_var.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)

# plot abs of covariances
ggplot(results, aes(x = gen, y = sum_abs_cov, group = interaction(site, replicate))) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(data = mean_results, aes(x = gen, y = sum_abs_cov, group = site, color = bio1), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:3)) +
  labs(x = "Generation", y = "Sum of |Cov(\u0394pi, \u0394pj)|", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste("../results/", today, "/linked_selection_sum_abs_cov.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)

# means of components
mean_results$gt2 <- mean_results$sum_abs_cov/(mean_results$sum_abs_cov + mean_results$sum_var)

ggplot(results, aes(x = gen, y = gt, group = interaction(site, replicate))) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(data = mean_results, aes(x = gen, y = gt2, group = site, color = bio1), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:3)) +
  labs(x = "Generation", y = "Proportion of allele frequency\nchange due to linked selection", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

# importance of fluctuating vs directional selection
ggplot(subset(results, gen == 3), aes(x = reorder(site, ratio, FUN = median), y = ratio)) +
  geom_boxplot() +
  geom_jitter() +
  geom_hline(yintercept = 0.5, lty = 2, lwd = 1.5) +
  labs(x = "Site", y = "Proportion of Cov(\u0394pi, \u0394pj) < 0", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste("../results/", today, "/linked_selection_directional_vs_fluctuating.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)
```

## analyze - simulations of gt curves

```{r}

params <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/config/gt_params.tsv", sep = "\t", header = T, data.table = F)

neutfreq_files <- list.files("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results", pattern = "*freqs.tsv", full.names = T)

#neutfreq_files <- sample(neutfreq_files, size = 1000, replace = F)

files_to_plot <- c(
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results/1_neutfreqs.tsv",
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results/10_neutfreqs.tsv",
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results/100_neutfreqs.tsv",
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results/1000_neutfreqs.tsv",
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results/1001_neutfreqs.tsv"
)

sim_curves <- NULL
output <- paste("simulated_gt_curves_" , today, ".txt", sep = "")

for(neutfreq_file in neutfreq_files){
  simid <- as.numeric(gsub(".*/", "", gsub("_neutfreqs.tsv", "", neutfreq_file)))
  
  print("Simulation ID:")
  print(simid)
  
  no_col <- max(count.fields(neutfreq_file, sep = "\t"))

  neutfreqs <- read.table(neutfreq_file, fill = T, col.names = 1:no_col, sep = "\t", skip = 5)

  neutfreqs <- reshape2::melt(neutfreqs, id = "X1")

  neutfreqs$pos <- str_extract(neutfreqs$value, ";.*;")
  neutfreqs$freq <- str_extract(neutfreqs$value, ";0.*$")

  neutfreqs$pos <- as.numeric(gsub(";", "", neutfreqs$pos))
  neutfreqs$freq <- as.numeric(gsub(";", "", neutfreqs$freq))

  neutfreqs <- neutfreqs %>%
    group_by(pos) %>%
    mutate(dp = freq - lag(freq))
  
  # ggplot(neutfreqs, aes(x = X1, y = dp, group = pos)) +
  #   geom_point() +
  #   geom_line()

  # Source - https://stackoverflow.com/a/5890689
  # Posted by Ista, modified by community. See post 'Timeline' for change history
  # Retrieved 2026-02-21, License - CC BY-SA 3.0
  neutfreqs <- neutfreqs[,c("X1", "pos", "dp")]
  neutfreqs <- neutfreqs[complete.cases(neutfreqs),]
  neutfreqs <- neutfreqs[!duplicated(neutfreqs[,c("X1", "pos")]),]
  dpmat <- reshape2::dcast(neutfreqs, pos ~ X1, value.var = "dp")

  #dpmat <- dpmat %>% filter(rowSums(is.na(.)) <= ncol(dpmat)*0.5)
  dpmat <- dpmat[,-2]
  
  #dpmat <- dpmat[complete.cases(dpmat),]

  dpcov_mat <- cov(dpmat[,-1], use = "pairwise.complete.obs")
  
  if(neutfreq_file %in% files_to_plot){
    print("Plotting variance covariance matrix...")
    plot_var_cov_matrix(dpcov_mat, paste(simid, "_var_cov_mat", sep = ""))
  }
  
  sim_curve <- gt_from_covmat(dpcov_mat)

  param_row <- params[(params$ID == simid),]
  
  sim_curve$alpha <- param_row[, "alpha"]
  sim_curve$sigma <- param_row[, "sigma"]
  sim_curve$tmu <- param_row[, "tmu"]
  sim_curve$adjust <- param_row[, "adjust"]
  
  write.table(sim_curve, output, col.names=!file.exists(output), append = T, row.names = F, sep = ",", quote = F)
  
}

```

# plot - simulations of gt curves
```{r}
sim_curves <- fread("simulated_gt_curves_2.txt")

# average across reps in a site
sim_curves$alpha2 = (sim_curves$alpha > 0)
sim_curves$alpha2[(sim_curves$alpha > 0)] = "Selection"
sim_curves$alpha2[(sim_curves$alpha == 0)] = "Neutral"

mean_sims = sim_curves %>% 
  group_by(gen, sigma, alpha, adjust) %>% 
  summarize(gts = mean(gt), 
            sum_cov = mean(sum_cov), 
            sum_abs_cov = mean(sum_abs_cov),
            sum_pos_cov = mean(sum_pos_cov),
            sum_neg_cov = mean(sum_neg_cov),
            sum_var = mean(sum_var), 
            ratio = mean(ratio, na.rm = T))

# gt over time
ggplot(mean_sims, aes(x = gen, y = gts, group = sigma, color = as.factor(sigma))) +
  geom_point(size = 1.5) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  #facet_grid(alpha2~adjust, labeller = as_labeller(c("TRUE" = "Ne constant", "FALSE" = "Ne lowered", "Selection" = "Selection", "Neutral" = "Neutral"))) +
  facet_grid(alpha~adjust) +
  geom_line() +
  ylim(0,1) +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "G(t)", color = "Self\nrate")

ggsave(paste("../results/", today, "/simulated_linked_selection_gt.pdf", sep = ""), height = 4.5, width = 7, device = cairo_pdf)

# sum of covariances over time
ggplot(mean_sims, aes(x = gen, y = sum_cov, group = sigma, color = as.factor(sigma))) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, lty = 2, lwd = 1.1) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  #facet_grid(alpha2~adjust, labeller = as_labeller(c("TRUE" = "Ne constant", "FALSE" = "Ne lowered", "Selection" = "Selection", "Neutral" = "Neutral"))) +
  facet_grid(alpha~adjust) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "Sum of Cov(\u0394pi, \u0394pj)", color = "Self\nrate")

ggsave(paste("../results/", today, "/simulated_linked_selection_sum_cov.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)

# sum of absolute value of covariances over time
ggplot(mean_sims, aes(x = gen, y = sum_abs_cov, group = sigma, color = as.factor(sigma))) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, lty = 2, lwd = 1.1) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  #facet_grid(alpha2~adjust, labeller = as_labeller(c("TRUE" = "Ne constant", "FALSE" = "Ne lowered", "Selection" = "Selection", "Neutral" = "Neutral"))) +
  facet_grid(alpha~adjust) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "Sum of |Cov(\u0394pi, \u0394pj)|", color = "Self\nrate")

ggsave(paste("../results/", today, "/simulated_linked_selection_sum_abs_cov.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)

# sum variances over time
ggplot(mean_sims, aes(x = gen, y = sum_var, group = interaction(sigma, alpha), color = as.factor(sigma))) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, lty = 2, lwd = 1.1) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  #facet_wrap(alpha2~adjust, labeller = as_labeller(c("TRUE" = "Ne constant", "FALSE" = "Ne lowered", "Selection" = "Selection", "Neutral" = "Neutral"))) +
  facet_grid(alpha~adjust) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "Sum of Var(\u0394pi)", color = "Self\nrate", linetype = "|s| > 0")

ggsave(paste("../results/", today, "/simulated_linked_selection_sum_var.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)

# importance of fluctuating vs directional selection
ggplot(mean_sims, aes(x = gen, y = ratio, group = sigma, color = as.factor(sigma))) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0.5, lty = 2, lwd = 1.1) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  #facet_grid(alpha2~adjust, labeller = as_labeller(c("TRUE" = "Ne constant", "FALSE" = "Ne lowered", "Selection" = "Selection", "Neutral" = "Neutral"))) +
  facet_grid(alpha~adjust) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "Proportion of Cov(\u0394pi, \u0394pj) < 0", color = "Self\nrate")

ggsave(paste("../results/", today, "/simulated_linked_selection_fluctuating.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)

# sum of negative covariances
ggplot(mean_sims, aes(x = gen, y = ratio, group = sigma, color = as.factor(sigma))) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0.5, lty = 2, lwd = 1.1) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  #facet_grid(alpha2~adjust, labeller = as_labeller(c("TRUE" = "Ne constant", "FALSE" = "Ne lowered", "Selection" = "Selection", "Neutral" = "Neutral"))) +
  facet_grid(alpha~adjust) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "Proportion of Cov(\u0394pi, \u0394pj) < 0", color = "Self\nrate")

ggsave(paste("../results/", today, "/simulated_linked_selection_fluctuating.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)

# try using averages of variances components
mean_sims$gt2 = mean_sims$sum_cov/(mean_sims$sum_cov + mean_sims$sum_var)

ggplot(mean_sims, aes(x = gen, y = gt2, group = interaction(sigma, alpha), color = as.factor(sigma))) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, lty = 2, lwd = 1.1) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  #facet_wrap(alpha2~adjust, labeller = as_labeller(c("TRUE" = "Ne constant", "FALSE" = "Ne lowered", "Selection" = "Selection", "Neutral" = "Neutral"))) +
  facet_grid(alpha~adjust) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "G(t)\nmean components", color = "Self\nrate")

ggsave(paste("../results/", today, "/simulated_linked_selection_gt_mean_components.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)

# absolute value of covariance
ggplot(mean_sims, aes(x = gen, y = sum_abs_cov/(sum_abs_cov + sum_var), group = interaction(sigma, alpha), color = as.factor(sigma))) +
  geom_point(size = 1.5) +
  geom_vline(xintercept = 5) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  #facet_wrap(alpha2~adjust, labeller = as_labeller(c("TRUE" = "Ne constant", "FALSE" = "Ne lowered", "Selection" = "Selection", "Neutral" = "Neutral"))) +
  facet_grid(alpha~adjust) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "G(t)\nabs. value", color = "Self\nrate")

ggsave(paste("../results/", today, "/simulated_linked_selection_gt_abs_value.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)
```

## temporal covariances - phase 2

```{r}
# # full dataset
# gen0 <- vroom("/global/scratch/users/milesroberts/tmp/average_seedmix_p0.txt", delim = "\t", col_names = FALSE)
# 
# start <- Sys.time()
# hapfire_phase1 <- vroom("/global/scratch/users/milesroberts/tmp/merged_hapFIRE_allele_frequency.csv", delim = ",")
# end <- Sys.time()
# end - start
# 
# hapfire_phase2 <- vroom("/global/scratch/users/milesroberts/tmp/PHASE2_merged_hapFIRE_allele_frequency.csv", delim = ",")
# 
# hapfire <- cbind(hapfire_phase1, hapfire_phase2)
# rm(hapfire_phase1, hapfire_phase2)
# gc()
# 
# # randomly swith which allele is being tracked
# # right now, the non-reference allele is always the one we're tracking
# idx <- sample(1:nrow(hapfire), size = round(nrow(hapfire)/2), replace = F)
# hapfire[idx,] <- 1 - hapfire[idx,]
# 

# make list of sites
sites <- unique(gsub("_.*", "", colnames(hapfire)))

# for now, ignore sites that lack reps with contiguous generations
sites <- sites[c(1,3,5:17,19:21,23,25:29,31)]

# loop over sites
all_gt_curves = NULL

for(site in sites){
  print(paste("Site:", site))
  
  # subset to one site
  print("Subsetting to one site...")
  hap_site <- hapfire %>% select(starts_with(paste(site, "_", sep = "")))
  
  # partition data by generation
  # should work for an arbitrary number of generations
  print("Paritioning data by generation...")
  num_gen <- length(unique(unlist(str_extract_all(colnames(hap_site), "_.*_"))))
  
  # get replicates that are in all generations
  reps <- gsub(".*_", "", colnames(hap_site))
  rep_tab <- table(reps)
  kept_reps <- names(rep_tab)[rep_tab == num_gen]
  
  gens <- list()
  gens[[1]] <- gen0$X3
  for(i in 1:num_gen){
    gens[[i+1]] <- hap_site %>% 
      select(matches(paste("_", i, "_", kept_reps, "$", sep = "", collapse = "|"))) %>%
      relocate(all_of(paste(site,"_", i, "_", kept_reps, sep = "")))
  }
  
  # calculate allele frequency differences between generations
  print("Calculating allele frequency increments...")
  dps <- list()
  for (i in 1:(length(gens) - 1)) {
    print(i)
    dps[[i]] <- gens[[i + 1]] - gens[[i]]
  }
  
  # calculate gt curves
  site_gt = gt_from_pmat(dps, calcabs = T, site)
  
  # save result
  site_gt$site = site
  all_gt_curves = rbind(all_gt_curves, site_gt)
}

#### PLOT #### 
# average across reps in a site
mean_gts = all_gt_curves %>% 
  group_by(site, gen) %>% 
  summarize(gt = mean(gt), 
            sum_cov = mean(sum_cov), 
            sum_var = mean(sum_var), 
            sum_abs_cov = mean(sum_abs_cov),
            sum_pos_cov = mean(sum_pos_cov),
            sum_neg_cov = mean(sum_neg_cov),
            ratio = mean(ratio, na.rm = T),
            n = n())

# remove sites with very few replicates or did not have at least 3 generations (including generation 0)
all_gt_curves <- all_gt_curves %>% filter(site != "10" & site != "43" & site != "24" & site != "26" & site != "23" & site != "37" & site != "60")
  
mean_gts <- mean_gts %>% filter(n > 2 & n <= 12)

# merge in climate data
mean_gts <- merge(mean_gts, worldclim_sitesdata, by = "site")

# plot mean of averages
ggplot(all_gt_curves, aes(
  x = gen,
  y = gt,
  group = interaction(site, replicate)
)) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(
    data = mean_gts,
    aes(
      x = gen,
      y = gt,
      group = site,
      color = bio1
    ),
    lwd = 2,
    alpha = 1
  ) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name = "Generation", limits = c(1:5)) +
  labs(x = "Generation", y = "Proportion of allele frequency\nchange due to linked selection", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste("../results/", today, "/linked_selection.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)

# plot covariances
ggplot(all_gt_curves, aes(x = gen, y = sum_cov, group = interaction(site, replicate))) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(data = mean_gts, aes(x = gen, y = sum_cov, group = site, color = bio1), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:5)) +
  labs(x = "Generation", y = "Sum of Cov(\u0394pi, \u0394pj)", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste("../results/", today, "/linked_selection_sum_cov.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)

# plot variances
ggplot(all_gt_curves, aes(x = gen, y = sum_var, group = interaction(site, replicate))) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(data = mean_gts, aes(x = gen, y = sum_var, group = site, color = bio1), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:5)) +
  labs(x = "Generation", y = "Var(\u0394pi)", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste("../results/", today, "/linked_selection_sum_var.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)

# plot positive and negative covariances
all_gt_curves_posneg <- all_gt_curves %>% pivot_longer(cols = c(sum_pos_cov, sum_neg_cov))

mean_gts_posneg <- mean_gts %>% pivot_longer(cols = c(sum_pos_cov, sum_neg_cov))

all_gt_curves_posneg[(all_gt_curves_posneg$name == "sum_neg_cov"),"value"] <- all_gt_curves_posneg[(all_gt_curves_posneg$name == "sum_neg_cov"),"value"]*(-1)

mean_gts_posneg[(mean_gts_posneg$name == "sum_neg_cov"),"value"] <- mean_gts_posneg[(mean_gts_posneg$name == "sum_neg_cov"),"value"]*(-1)
                                                                                                                                                          
ggplot(all_gt_curves_posneg, aes(x = gen, y = value, group = interaction(site,name, replicate))) +
  facet_wrap(.~name, nrow = 2, ncol = 1, scales = "free_y")+
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_hline(yintercept = 0, lwd = 1.2, lty = 2) +
  geom_line(data = mean_gts_posneg, aes(x = gen, y = value, group = interaction(site,name), color = bio1), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:5)) +
  labs(x = "Generation", y = "Sum of Cov(\u0394pi, \u0394pj)", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste("../results/", today, "/linked_selection_sum_cov_components.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)

# plot abs of covariances
ggplot(all_gt_curves, aes(x = gen, y = sum_abs_cov, group = interaction(site, replicate))) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(data = mean_gts, aes(x = gen, y = sum_abs_cov, group = site, color = bio1), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:5)) +
  labs(x = "Generation", y = "Sum of |Cov(\u0394pi, \u0394pj)|", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste("../results/", today, "/linked_selection_sum_abs_cov.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)

# means of components
mean_gts$gt2 <- mean_gts$sum_abs_cov/(mean_gts$sum_abs_cov + mean_gts$sum_var)

ggplot(all_gt_curves, aes(x = gen, y = gt, group = interaction(site, replicate))) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(data = mean_gts, aes(x = gen, y = gt2, group = site, color = bio1), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:5)) +
  labs(x = "Generation", y = "Proportion of allele frequency\nchange due to linked selection", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste("../results/", today, "/linked_selection_mean_components.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)

# importance of fluctuating vs directional selection
ggplot(subset(all_gt_curves, gen == 5), aes(x = reorder(site, ratio, FUN = median), y = ratio)) +
  geom_boxplot() +
  geom_jitter() +
  geom_hline(yintercept = 0.5, lty = 2, lwd = 1.5) +
  labs(x = "Site", y = "Proportion of Cov(\u0394pi, \u0394pj) < 0", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste("../results/", today, "/linked_selection_directional_vs_fluctuating.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)
```

## historical vs modern pairings

```{r}
pairings <- read.csv("../data/comb_hist+1001g.prs_geo+gen.csv")

historical <- pairings[, 1:2]

historical$type <- "historical"

colnames(historical) <- gsub("historical_", "", colnames(historical))

modern <- pairings[, 5:6]

modern$type <- "modern"

colnames(modern) <- gsub("modern_", "", colnames(modern))

slim_pairings <- rbind(historical, modern)

slim_pairings$time <- slim_pairings$year - min(slim_pairings$year) + 1

slim_pairings <- slim_pairings %>% group_by(type, time) %>% summarize(n = n()) %>% arrange(time, .by_group = T)

slim_pairings$ID = 1:nrow(slim_pairings)

write.table(slim_pairings, "../config/sampling_times.tsv", quote = F, row.names = F, sep = "\t")

sum(slim_pairings$n)

```


## Fst - historical vs modern

```{r}
cutoff = 0.05

msprime <- vroom(
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/msprime_fst.txt",
  delim = "\t",
  col_names = F
)

msprime$ID <- as.numeric(gsub(".*/|.txt", "", fst$X1))

names(msprime)[2] = "msprime_fst"

gcta <- vroom(
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/gcta_fst.txt",
  delim = "\t",
  col_names = F
)

gcta$ID <- as.numeric(gsub(".*/|.fst:1", "", gcta$X1))

gcta <- gcta %>% group_by(ID) %>% summarize(gcta_fst = mean(X7), S = n())

gcta_msprime <- merge(msprime, gcta, by = "ID")

# compare gcta and msprime
nei_weir_fst <- ggplot(gcta_msprime, aes(x = msprime_fst, y = gcta_fst)) +
  geom_point(shape = 16) +
  #geom_smooth(method = "lm") +
  geom_abline(slope = 1, intercept = 0, lwd = 1.2, color = "green") +
  labs(x = "Nei 1973 Fst (msprime)", y = "Weir 1996 Fst (gcta)")

# merge with window sizes and calculate p-values
params <- vroom(
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/config/fst_params.tsv",
  delim = "\t"
)

params_fst <- merge(params, gcta_msprime, by = "ID")

params_fst %>% 
  #filter(gcta_fst > 0) %>% 
  group_by(L) %>% 
  summarize(upb = quantile(gcta_fst, probs = 1 - cutoff/2, na.rm = T),
            lwb = quantile(gcta_fst, probs = cutoff / 2, na.rm = T),
            ost1_filter_test = 1 - ecdf(gcta_fst)(0.003892181),
            erecta_filter_test = 1 - ecdf(gcta_fst)(0.032451811),
            ost_unfilter_test = 1 - ecdf(gcta_fst)(0.011700622),
            erecta_unfilter_test = 1 - ecdf(gcta_fst)(0.014078437))

# tact distribution of fst from gcta
fst_density <- ggplot(params_fst, aes(x = gcta_fst)) +
  geom_density() +
  facet_wrap(L~., ncol = 1, labeller = "label_both") +
  geom_vline(data = filter(params_fst, L == 3081), aes(xintercept = 0.003892181), lty = 1) +
  geom_vline(data = filter(params_fst, L == 3081), aes(xintercept = 0.011700622), lty = 2) +
  geom_vline(data = filter(params_fst, L == 6583), aes(xintercept = 0.032451811), lty = 1) +
  geom_vline(data = filter(params_fst, L == 6583), aes(xintercept = 0.014078437), lty = 2) +
  labs(x = "Neutral Weir 1996 Fst\n(Sampling desing, Ne = 20,000)")


ggarrange(fst_density, nei_weir_fst, ncol = 2, labels = c("A", "B"), font.label = list(size = 16, family = "Helvetica"))

ggsave(paste("../results/", today, "/baseline_fst_historical_modern.pdf", sep = ""), height = 4, width = 8, device = cairo_pdf)
```


## load scRNA-seq data

```{r}
global <- readRDS("/global/scratch/users/milesroberts/tmp/GSE226097_global_integration_221009.rds")

meta <- global[[]]

#flower <- readRDS("../data/GSE226097_flower_250805.rds")

#flower[["data"]]
```

## check overlap between GEA and AEGIS

```{r}
lfmm_files <- list.files("lfmm.*", path = "../data", full.names = T)

lfmm <- lapply(lfmm_files, read.csv, header = T)

names(lfmm) <- lfmm_files

lfmm <- lfmm %>% 
  bind_rows(.id = "groups")

lfmm$chrom <- as.numeric(gsub("_.*", "", lfmm$snp_id))
lfmm$pos <- as.numeric(gsub(".*_", "", lfmm$snp_id))

lfmm$clim <- gsub(".*_", "", gsub("_k.*", "", lfmm$groups))

# Chr1:1185512..1188696 - CRY2
# Chr1:24331373..24333999 - FT
# Chr2:11058944..11063324 - ELF3
# Chr2:11531967..11534358 - CAM5
# Chr2:18807505..18811125 - SOC1
# Chr4:9195247..9199840 - PHYD
# Chr4:11000771..11003010 - TSF
# Chr5:3173382..3179448 - FLC
# Chr5:5170162..5173085 - CO
# Chr5:18589438..18591506 - DOG1
buffer = 10000
chroms = c(1,1,2,2,2,4,4,5,5,5)
starts = c(1185512,24331373,11058944,11531967,18807505,9195247,11000771,5170162,3173382,18589438)
ends = c(1188696,24333999,11063324,11534358,18811125,9199840,11003010,5173085,3179448,18591506)
starts = starts - buffer
ends = ends + buffer

lfmm_ft_only <- NULL
for(i in seq_along(chroms)){
  print(i)
  lfmm_ft <- lfmm %>% subset((chrom == chroms[i]) & (pos >= starts[i]) & (pos <= ends[i]) & (significant == "True"))
  lfmm_ft_only <- rbind(lfmm_ft_only, lfmm_ft)
}


head(lfmm_ft_only)
```
