---
title: "Linked selection in GrENE-net phase 2"
output: 
  html_notebook: 
    toc: true
editor_options: 
  chunk_output_type: inline
---

# set-up

```{r}
rm(list = ls())
gc()
.rs.restartR(clean = TRUE)
```

# packages

```{r}
library(ggplot2)
library(scico)
library(reshape2)
library(tidyverse)
library(broom)
library(data.table)
library(testthat)
library(ggpubr)
library(MareyMap)
library(ggforce)
library(arrow)
library(sf)
library(maps)
library(lme4)
library(scales)

box::use(./R/gt[...],
         ./R/plot[...],
         ./R/misc[...],
         )

# test modules
test_file("./R/tests.R")

# file paths
today <- Sys.Date()
results_path <- paste("../results/by-date/", today, "/", sep = "")

# global ggplot2 theme
theme_set(
  theme_minimal() +
    theme(
      text = element_text(
        size = 16,
        family = "Helvetica",
        colour = "black"
      ),
      axis.text = element_text(colour = "black"),
      plot.title = element_text(hjust = 0.5),
      strip.text = element_text(colour = "black"),
      panel.grid.minor = element_blank()
    )
)
```

# Prepare allele frequencies

## randomly switch tracked allele

```{r}
set.seed(1111)

# LD-pruned data
#gen0 <- read.table("../data/average_seedmix_p0_LDpruned.txt", sep = "\t")
#hapfire <- read.table("../data/merged_hapFIRE_allele_frequency_LDpruned.txt", sep = "\t", header = T)

# full dataset
gen0 <- vroom("/global/scratch/users/milesroberts/tmp/average_seedmix_p0.txt", delim = "\t", col_names = F)

# randomly swith which allele is being tracked
# right now, the non-reference allele is always the one we're tracking
idx <- sample(c(T,F), size = nrow(gen0), replace = T)
gen0[idx,"X3"] <- 1 - gen0[idx,"X3"]

write.csv(gen0, "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/seedmix_hapfire_allele_frequency_random_switch.csv", quote = F, row.names = F)

#hapfire_phase1 <- read_csv("/global/scratch/users/milesroberts/tmp/merged_hapFIRE_allele_frequency.csv")

# start <- Sys.time()
# hapfire_phase1 <- vroom("/global/scratch/users/milesroberts/tmp/merged_hapFIRE_allele_frequency.csv", delim = ",")
# end <- Sys.time()
# end - start
# 
# hapfire_phase2 <- vroom("/global/scratch/users/milesroberts/tmp/PHASE2_merged_hapFIRE_allele_frequency.csv", delim = ",")
# 
# hapfire <- cbind(hapfire_phase1, hapfire_phase2)
# 
# write.csv(hapfire, "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_2_hapfire_allele_frequency.csv", quote = F, row.names = F)

# randomly switch which allele is being tracked, same as gen0
# j = 0
# for(i in idx){
#   if(j %% 1000 == 0){
#     print(j)
#   }
#   hapfire[idx,] <- 1 - hapfire[idx,]
#   j = j + 1
# }

process_one_minus <- function(idx, chunk_size, output) {
  function(x, pos) {
    print(pos)
    print(Sys.time())
    # calculate end
    chunk_size <- nrow(x)
    end <- pos + chunk_size - 1
    # subtract 1 from some alleles
    print(paste("Start:", pos, "End:", end))
    idx_sub <- idx[pos:end]
    x[idx_sub,] <- 1 - x[idx_sub,]
    write.table(x, output, col.names=!file.exists(output), append = T, row.names = F, sep = ",", quote = F)
  }
}

# phase 1
read_csv_chunked(
  "/global/scratch/users/milesroberts/tmp/PHASE_1_2_COMBINED_hapFIRE_allele_frequency.csv",
  callback = SideEffectChunkCallback$new(
    process_one_minus(
      idx,
      5000,
      "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_2_COMBINED_hapFIRE_allele_frequency_random_switch.csv"
    )
  ),
  chunk_size = 5000
)

# phase 2
# read_csv_chunked(
#   "/global/scratch/users/milesroberts/tmp/PHASE2_merged_hapFIRE_allele_frequency.csv",
#   callback = SideEffectChunkCallback$new(
#     process_one_minus(
#       idx,
#       5000,
#       "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_2_hapfire_allele_frequency_rng_switched.csv"
#     )
#   ),
#   chunk_size = 5000
# )
```

## remove fixed and lost loci

```{r}
# initial generation
# tracked allele randomly switched
gen0 <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/seedmix_hapfire_allele_frequency_random_switch.csv", data.table = F)

# later gen
# tracked allele randomly switched
hapfire <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_2_COMBINED_hapFIRE_allele_frequency_random_switch.csv", data.table = F)

# omit allele frequencies that start near abosrbing boundaries
non_abosrbing_freq <- (gen0$X3 >= 1/231) & (gen0$X3 <= 1 - 1/231)
gen0 <- gen0[non_abosrbing_freq,]
hapfire <- hapfire[non_abosrbing_freq,]

rm(non_abosrbing_freq)
gc()

# output filtered seedmix frequencies
write.table(gen0, "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/seedmix_hapfire_allele_frequency_random_switch_fixrm.csv", row.names = F, sep = ",", quote = F)

# omit allele frequencies that later reach absorbing boundaries
# it is possible that measurement errors cause frequency to reach 1 and then come back down
hapfire[(hapfire <= 0) | (hapfire >= 1)] <- NA

write.table(hapfire, "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_2_COMBINED_hapFIRE_allele_frequency_random_switch_fixrm.csv", row.names = F, sep = ",", quote = F)

# loop over site - replicate combinations
# remove allele frequency trajectory after first fixation/loss 
sites <- unique(gsub("_.*", "", colnames(hapfire)))
hapfire_nas <- NULL
for(site in unique(sites)){
  print(paste("Site:", site))
  hap_site <- hapfire %>% dplyr::select(starts_with(paste(site, "_", sep = "")))
  for(exprep in 1:12){
    print(paste("Rep:", exprep))
    hap_rep <- hap_site %>% dplyr::select(ends_with(paste("_", exprep, sep = "")))
    
    # sort names to be in order of generation
    hap_rep <- hap_rep %>% relocate(all_of(sort(colnames(hap_rep))))
    
    if(ncol(hap_rep) > 1){
      hap_rep_na <- t(apply(hap_rep, MARGIN = 1, rm_na_after_na))
      colnames(hap_rep_na) <- colnames(hap_rep_na)
    } else {
      hap_rep_na <- hap_rep
    }
    
    hapfire_nas <- cbind(hapfire_nas, hap_rep_na)
  }
}

# summary of frequencies that were removed
print("Percentage of frequencies at loss/fixation:")
sum(is.na(hapfire_nas))/prod(dim(hapfire_nas)) * 100

# output filtered hapfire frequencies
write.table(hapfire_nas, "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_2_COMBINED_hapFIRE_allele_frequency_random_switch_fixrm_trajectory.csv", row.names = F, sep = ",", quote = F)
```

## arcsine square-root transformation

```{r}
# apply function per chunk and append to file
chunked_arcsin_sqrt <- function(output) {
  function(x, pos) {
    cat(paste(pos, Sys.time(), "\n"))
    append_table(arcsin_sqrt(x), output)
  }
}

# loop over chunks of csv
read_csv_chunked(
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_2_COMBINED_hapFIRE_allele_frequency_random_switch_fixrm_trajectory.csv",
  callback = SideEffectChunkCallback$new(
    chunked_arcsin_sqrt(
      "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_2_COMBINED_hapFIRE_allele_frequency_random_switch_fixrm_trajectory_arcsin.csv"
    )
  ),
  chunk_size = 10000
)

```

# loading datasets

## sample sizes

```{r}

###
#setDTthreads(0) # use all threads
SUBSET = F # for testing, just load small subset of data
ASIN = F

#gen0 <- vroom("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/seedmix_hapfire_allele_frequency_rng_switched.csv", delim = ",", col_names = T)

# initial generation
# tracked allele randomly switched
#gen0 <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/seedmix_hapfire_allele_frequency_random_switch_fixrm.csv", data.table = F)

if(SUBSET){
  if(ASIN){
    gen0 <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/seedmix_hapfire_allele_frequency_random_switch_fixrm.csv", data.table = F, nrows = 30000)
    gen0$X3 <- arcsin_sqrt(gen0$X3)
  } else {
    gen0 <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/seedmix_hapfire_allele_frequency_random_switch_fixrm.csv", data.table = F, nrows = 30000)
  }
} else {
  if(ASIN){
    gen0 <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/seedmix_hapfire_allele_frequency_random_switch_fixrm.csv", data.table = F)
    gen0$X3 <- arcsin_sqrt(gen0$X3)
  } else {
    gen0 <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/seedmix_hapfire_allele_frequency_random_switch_fixrm.csv", data.table = F, nThread = 8)
  }
}


# later gen
# tracked allele randomly switched
#hapfire <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_2_COMBINED_hapFIRE_allele_frequency_random_switch_fixrm.csv", data.table = F)

if(SUBSET){
  if(ASIN){
      hapfire <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_2_COMBINED_hapFIRE_allele_frequency_random_switch_fixrm_trajectory_arcsin.csv", data.table = F, nrows = 300000)
  } else {
      hapfire <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_2_COMBINED_hapFIRE_allele_frequency_random_switch_fixrm_trajectory.csv", data.table = F, nrows = 300000)
  }
} else {
  if(ASIN){
      hapfire <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_2_COMBINED_hapFIRE_allele_frequency_random_switch_fixrm_trajectory_arcsin.csv", data.table = F)
  } else {
      hapfire <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_2_COMBINED_hapFIRE_allele_frequency_random_switch_fixrm_trajectory.csv", data.table = F)
  }
}

# year 1-3 sample sizes
sample_sizes_13 <- fread("/global/home/users/milesroberts/moi_lab_projects/grenenet-phase2/data/samples_data_fix57.csv")

sample_sizes_13$site_gen_plot <- paste(sample_sizes_13$site, sample_sizes_13$fix_57_generation, sample_sizes_13$plot, sep = "_")

sample_sizes_13 <- sample_sizes_13 %>% group_by(site_gen_plot) %>% summarize(n = sum(flowerscollected))

# year 4-5 sample sizes
sample_sizes_45 <- fread("/global/home/users/milesroberts/moi_lab_projects/grenenet-phase2/data/grenenet_year4_5_rep_metadata_predicted_flower_final.csv")

sample_sizes_45$site_gen_plot <- paste(sample_sizes_45$site, sample_sizes_45$fixed_57_generation, sample_sizes_45$TrayID, sep = "_")

sample_sizes_45 <- sample_sizes_45 %>% group_by(site_gen_plot) %>% summarize(n = sum(predicted_flower_counts))

# combine all sizes
sample_sizes <- rbind(sample_sizes_13, sample_sizes_45)

# number of chromosomes = 2*flower count
sample_sizes$n <- 2*sample_sizes$n
all(sample_sizes$n > 1)

rm(sample_sizes_13, sample_sizes_45)
gc()
# load site climatic data
#load("../data/sites.clim.rda")

#sites.clim <- sites.clim[,c("SITE_CODE", paste("bio", 1:19, sep = ""))]
```

## allele frequency matrix

```{r}
# one time conversion
if(!dir.exists("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/hapfire_parquet")){
  open_dataset("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_2_COMBINED_hapFIRE_allele_frequency_random_switch_fixrm_trajectory.csv", format = "csv", read_options = csv_read_options(block_size = 16L * 1024L * 1024L)) |>
  write_dataset("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/hapfire_parquet", format = "parquet", max_rows_per_file = 500000L)
}

# now can just read it
hapfire <- open_dataset("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/hapfire_parquet")
```

## site data

```{r}
# from grene-net r package
load("../data/worldclim_sitesdata.rda")
load("../data/locations_data.rda")

# 1. Create a spatial object for your target point
target_points <- locations_data %>% st_as_sf(coords = c("longitude", "latitude"), crs = 4326)

# 2. Get a dataset of world or US cities (from the maps package)
data(world.cities, package = "maps")

cities_sf <- world.cities %>%
  st_as_sf(coords = c("long", "lat"), crs = 4326)

# 3. Find the index of the nearest city
nearest_index <- st_nearest_feature(target_points, cities_sf)

# 4. Pull the closest city details
closest_city <- world.cities[nearest_index, ]
print(closest_city)

locations_data$city_country <- paste(closest_city$name, ", ", closest_city$country.etc, sep = "")
```

## LD matrix

```{r}
# one time conversion
if(!dir.exists("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/plink_parquet")){
  open_dataset("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/plink.ld.tsv", format = "tsv", read_options = csv_read_options(block_size = 16L * 1024L * 1024L)) |>
  write_dataset("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/plink_parquet", format = "parquet", max_rows_per_file = 2000000L)
}

ld_matrix <- open_dataset("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/plink_parquet")
```

# Feder test

## LD pruned data

```{r}
# subset different generations
gen1 <- hapfire[,grepl("_1_", names(hapfire))] 

gen2 <- hapfire[,grepl("_2_", names(hapfire))] 

gen3 <- hapfire[,grepl("_3_", names(hapfire))]

gen4 <- hapfire[,grepl("_4_", names(hapfire))] 

# subset to only plots present in all 3 years
shared_gen1 <- (gsub("_1_","_",names(gen1)) %in% gsub("_2_","_",names(gen2)))& (gsub("_1_","_",names(gen1)) %in% gsub("_3_","_",names(gen3))) & (gsub("_1_","_",names(gen1)) %in% gsub("_4_","_",names(gen4)))

shared_gen2 <- (gsub("_2_","_",names(gen2)) %in% gsub("_3_","_",names(gen3))) & (gsub("_2_","_",names(gen2)) %in% gsub("_1_","_",names(gen1))) & (gsub("_2_","_",names(gen2)) %in% gsub("_4_","_",names(gen4)))

shared_gen3 <- (gsub("_3_","_",names(gen3)) %in% gsub("_2_","_",names(gen2))) & (gsub("_3_","_",names(gen3)) %in% gsub("_1_","_",names(gen1))) & (gsub("_3_","_",names(gen3)) %in% gsub("_4_","_",names(gen4)))

shared_gen4 <- (gsub("_4_","_",names(gen4)) %in% gsub("_2_","_",names(gen2))) & (gsub("_4_","_",names(gen4)) %in% gsub("_1_","_",names(gen1))) & (gsub("_4_","_",names(gen4)) %in% gsub("_3_","_",names(gen3)))

gen1_sub <- gen1[,shared_gen1]

gen2_sub <- gen2[,shared_gen2]

gen3_sub <- gen3[,shared_gen3]

gen4_sub <- gen4[,shared_gen4]

# sort columns to match
gen4_sub <- gen4_sub[,paste("1_4", c(1,4,5,6,7,8,9,10,11,12), sep = "_")]

# calculate statistic
y1 <- (gen1_sub - gen0$V3)/sqrt(2*gen0$V3*(1-gen0$V3)*(1-0))
y1$gen <- 1
y1$CHROM <- gen0$V1
y1$POS <- gen0$V2
y1 <- melt(y1, id = c("gen", "CHROM", "POS"))

y2 <- (gen2_sub - gen1_sub)/sqrt(2*gen1_sub*(1-gen1_sub)*(2-1))
y2$gen <- 2
y2$CHROM <- gen0$V1
y2$POS <- gen0$V2
y2 <- melt(y2, id = c("gen", "CHROM", "POS"))

y3 <- (gen3_sub - gen2_sub)/sqrt(2*gen2_sub*(1-gen2_sub)*(3-2))
y3$gen <- 3
y3$CHROM <- gen0$V1
y3$POS <- gen0$V2
y3 <- melt(y3, id = c("gen", "CHROM", "POS"))

yvalues <- rbind(y1, y2, y3)
yvalues$SITE <- gsub("X", "", gsub("_.*", "", yvalues$variable))
yvalues$REP <- gsub(".*_", "", yvalues$variable)

t_test_stat <- yvalues %>% group_by(CHROM, POS, SITE) %>% summarize(ybar = mean(value, na.rm = T), s2 = var(value, na.rm = T), n = n())

t_test_stat$tvalue <- t_test_stat$ybar/sqrt(t_test_stat$s2/t_test_stat$n)

t_test_stat$pvalue <- dt(t_test_stat$tvalue, df = t_test_stat$n)

# Manhattan plot
map(unique(t_test_stat$SITE), ~plot_manhattan_by_site(data = t_test_stat, site = .x, today, height = 4, width = 10))

#load("../data/allele_allyears_frequencies_long_raw_climate.rda")
# calculate mean and average

# ybar <- (y1 + y2 + y3)/3
# 
# s2 <- (1/(3-1))*( (y1 - ybar)^2 + (y2 - ybar)^2 + (y3 - ybar)^2 )
# 
# t_test_stat <- ybar/sqrt(s2/3)
# 
# # 
# t_test_stat$CHROM <- gen0$V1
# 
# t_test_stat$POS <- gen0$V2
# 
# t_test_stat <- melt(t_test_stat, id = c("CHROM", "POS"))
# 
# t_test_stat$SITE <- gsub("X", "", gsub("_.*", "", t_test_stat$variable))
# 
# t_test_stat$REP <- gsub(".*_", "", t_test_stat$variable)
# 
# # average across reps of each pos at each site
# mean_t_stat <- t_test_stat %>% 
#   group_by(CHROM, POS, SITE) %>% 
#   summarize(MEANT = mean(value, na.rm = T), SDT = sd(value, na.rm = T))
# 
# # calculate p-values
# #
# mean_t_stat$PVALUE <- dt(mean_t_stat$MEANT, df = 2)
# 
# # adjust p-values
# mean_t_stat$ADJPVALUE <- p.adjust(mean_t_stat$PVALUE, method = "bonferroni")
# 
# # plot distribution of stats across positions and sites
# ggplot(mean_t_stat, aes(x = MEANT)) +
#   geom_density()
#   xlim(c(-10, 10))
#   
# # Manhattan plot
# map(unique(mean_t_stat$SITE), ~manhattan_by_site(data = mean_t_stat, site = .x, today, height = 4, width = 10))

#
#summary(unlist(as.vector(t_test_stat)))

```

## Phase 1 data

```{r}

# Source - https://stackoverflow.com/a/33051935
# Posted by scoa
# Retrieved 2026-02-27, License - CC BY-SA 3.0
# con <- file("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/merged_hapFIRE_allele_frequency.csv","r")
# F_header <- readLines(con,n=1)
# close(con)
# 
# F_subset <- read_csv("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/merged_hapFIRE_allele_frequency.csv", col_select = "1_1_1")

gen0 <- read.table("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/average_seedmix_p0.txt", sep = "\t")

process_F <- function(gen0, chunk_size, output) {
  function(x, pos) {
    print(pos)
    print(Sys.time())
    gen0 <- gen0[pos:(pos + chunk_size - 1),]
    # generations
    header_names <- names(x)
    gen1 <- x[,grepl("_1_", header_names)] 
    gen2 <- x[,grepl("_2_", header_names)] 
    gen3 <- x[,grepl("_3_", header_names)]
    
    # subset to populations shared across generations
    names_gen1 <- names(gen1)
    names_gen2 <- names(gen2)
    names_gen3 <- names(gen3)
    
    shared_gen1 <- (gsub("_1_","_",names_gen1) %in% gsub("_2_","_",names_gen2)) & (gsub("_1_","_",names_gen1) %in% gsub("_3_","_",names(gen3)))
    shared_gen2 <- (gsub("_2_","_",names_gen2) %in% gsub("_3_","_",names_gen3)) & (gsub("_2_","_",names_gen2) %in% gsub("_1_","_",names_gen1))
    shared_gen3 <- (gsub("_3_","_",names_gen3) %in% gsub("_2_","_",names_gen2)) & (gsub("_3_","_",names_gen3) %in% gsub("_1_","_",names_gen1))
    
    gen1_sub <- gen1[,shared_gen1]
    gen2_sub <- gen2[,shared_gen2]
    gen3_sub <- gen3[,shared_gen3] 
    # calculate statistic
    y1 <- (gen1_sub - gen0$V3)/sqrt(2*gen0$V3*(1-gen0$V3)*(1-0))
    y1$gen <- 1
    y1$CHROM <- gen0$V1
    y1$POS <- gen0$V2
    y1 <- melt(y1, id = c("gen", "CHROM", "POS"))
    
    y2 <- (gen2_sub - gen1_sub)/sqrt(2*gen1_sub*(1-gen1_sub)*(2-1))
    y2$gen <- 2
    y2$CHROM <- gen0$V1
    y2$POS <- gen0$V2
    y2 <- melt(y2, id = c("gen", "CHROM", "POS"))
    
    y3 <- (gen3_sub - gen2_sub)/sqrt(2*gen2_sub*(1-gen2_sub)*(3-2))
    y3$gen <- 3
    y3$CHROM <- gen0$V1
    y3$POS <- gen0$V2
    y3 <- melt(y3, id = c("gen", "CHROM", "POS"))
    
    yvalues <- rbind(y1, y2, y3)
    yvalues$SITE <- gsub("X", "", gsub("_.*", "", yvalues$variable))
    yvalues$REP <- gsub(".*_", "", yvalues$variable)
    
    t_test_stat <- yvalues %>% group_by(CHROM, POS, SITE) %>% summarize(ybar = mean(value, na.rm = T), s2 = var(value, na.rm = T), n = n())
    
    t_test_stat$tvalue <- t_test_stat$ybar/sqrt(t_test_stat$s2/t_test_stat$n)
    
    t_test_stat$pvalue <- dt(t_test_stat$tvalue, df = t_test_stat$n)
    
    write.table(t_test_stat, output, col.names=!file.exists(output), append = T, row.names = F, sep = ",", quote = F)
  }
}

read_csv_chunked(
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/merged_hapFIRE_allele_frequency.csv",
  #callback = SideEffectChunkCallback$new(function(x, pos){print(pos)}),
  callback = SideEffectChunkCallback$new(process_F(gen0, 1000, "feder_t_test.txt")),
  chunk_size = 1000)
                

```

## Phase 2 data

```{r}
# make list of sites
sites <- unique(gsub("_.*", "", colnames(hapfire)))

# for now, ignore sites that lack reps with contiguous generations
#sites <- sites[c(1,3,5:17,19:21,23,25:29,31)]
sites <- sites[-30]

sites <- sites[-1:-9]

# loop over sites
output <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-03-31/", "feder_test_phase2.txt", sep = "")

for(site in sites){
  print(paste("Site:", site))
  yvalues <- NULL
  for(exprep in 1:12) {
    print(paste("Rep:", exprep))
    # subset to one site
    print("Subsetting to one replicate...")
    hap_rep <- hapfire %>% 
      select(starts_with(paste(site, "_", sep = "")) & ends_with(paste("_", exprep, sep = "")))
    
    # partition data by generation
    # should work for an arbitrary number of generations
    print("Paritioning data by generation...")
    colnames(hap_rep) <- gsub("_", "", sort(unlist(str_extract_all(colnames(hap_rep), "_.*_"))))
    
    # sort names to be in order of generation
    hap_rep$`0` <- gen0$X3
    hap_rep <- hap_rep %>% relocate(all_of(sort(colnames(hap_rep))))
    
    if(ncol(hap_rep) == 1){
      print("No columns in allele frequency matrix, skipping to next replicate")
      next
    }
    
    # calculate allele frequency differences between generations
    print("Calculating allele frequency increments...")

    years <- as.numeric(colnames(hap_rep))
    dts <- (years - lag(years))[-1]
    print("Time increments:")
    print(dts)
    # calculate y values for t test
    for (i in 1:(length(years) - 1)) {
      print(paste("Generation ", i, sep = ""))
      p0 <- hap_rep[,i]
      p1 <- hap_rep[,i + 1]
      dp <- p1 - p0
      
      yvalue <- data.frame(
        y = dp/sqrt(2*p0*(1-p0)*dts[i]),
        CHROM = gen0$X1,
        POS = gen0$X2,
        gen = i,
        SITE = site,
        rep = exprep
      )
      
      yvalues <- rbind(yvalues, yvalue)
    }
  }
  
  # calculate averages and variances for a snp at a site
  print("Averaging across replicates and generations at a site")
  yvalues <- yvalues %>% group_by(SITE, CHROM, POS) %>% 
    summarize(ybar = mean(y, na.rm = T), 
              s2 = var(y, na.rm = T),
              n = n())
  
  yvalues$tvalue <- yvalues$ybar/sqrt(yvalues$s2/yvalues$n)
  yvalues$pvalue <- dt(yvalues$tvalue, df = yvalues$n)
  
  write.table(yvalues, output, col.names=!file.exists(output), append = T, row.names = F, sep = ",", quote = F)
}
```

## Try modifying Feder test to account for sampling noise

```{r}

p0 = 0.5
N0 = 500
L = 1:30
lambda = 100

truth = c(p0)
estims = c(p0)
covs = c(lambda)

# simulate 
for(i in L){
  # generate new generation
  x1 <- rbinom(1, N0, prob = p0)
  p1 <- x1/N0
  # sequencing
  coverage <- rpois(1, lambda = lambda)
  x1_hat <- rbinom(1, size = coverage, prob = p1)
  p1_hat <- x1_hat/coverage
  # save results
  truth = c(truth, p1)
  estims = c(estims, p1_hat)
  covs = c(covs, coverage)
}

results = data.frame(
  p = truth,
  p_hat = estims,
  cov = covs,
  i = c(0, L)
)

variance_table = data.frame(
  dt = diff(results$i),
  cov1 = results$cov[1:(nrow(results) - 1)],
  cov2 = results$cov[2:nrow(results)],
  dp = 2*results$p_hat[-1]*(1 - results$p_hat[-1])
)


variance_table$var = sqrt(variance_table$dt * variance_table$dp * (1/variance_table$cov1 + 1/variance_table$cov2))

z_stats <- diff(results$p_hat)/variance_table$var

hist(z_stats)

shapiro.test(z_stats)
```

# GLM 

https://doi.org/10.1038/s41586-024-07834-x

## calculate GLM

```{r}
# make list of sites
sites <- unique(gsub("_.*", "", colnames(hapfire)))

# for now, ignore sites that lack reps with contiguous generations
#sites <- sites[c(1,3,5:17,19:21,23,25:29,31)]
sites <- sites[c(-30,-32)]

# number of snps per window
window_size <- 3000

# create windows, should not bleed between chromosomes
blocks <- gen0 %>%
  group_by(X1) %>%
  mutate(block = (row_number() - 1) %/% window_size + 1,
         block = paste(X1, block, sep = "_"))

length(unique(blocks$block))

hapfire$block <- blocks$block

#output_glm <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/", today, "/glm_quasibinomial_coef.txt", sep = "")
output_glm <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-05-04/glm_quasibinomial_coef.txt", sep = "")

unique_blocks <- unique(hapfire$block)

# need to wrap function to handle cases when the data for a snp is all NA
safe_glm <- possibly(.f = function(df) {
  glm(value ~ t, data = df, weights = n, family = quasibinomial(link = "logit"))
}, otherwise = NULL)

for(b in unique_blocks[320:418]){
  print(paste("Window:", b))
  hap_block <- hapfire %>% filter(block == b)
  for(site in unique(sites)){
    print(paste("Site:", site))
    hap_site <- hap_block %>% dplyr::select(starts_with(paste(site, "_", sep = "")))
    gen0_block <- blocks %>% filter(block == b) %>% dplyr::select(starts_with("X"))
    
    # get sample sizes
    rep_n <- sample_sizes[(sample_sizes$site_gen_plot %in% colnames(hap_site)),]
    rep_n <- rbind(c(0, 1e3), rep_n)
    rep_n <- rep_n %>% arrange(site_gen_plot)
    
    # sort names to be in order of generation
    hap_site <- hap_site %>% relocate(all_of(sort(colnames(hap_site))))
    
    hap_site <- cbind(hap_site, gen0_block)
    
    hap_site <- reshape2::melt(hap_site, id = c("X1", "X2"))
    
    names(hap_site)[3] <- "site_gen_plot"
    
    hap_site <- merge(hap_site, rep_n, by = "site_gen_plot", all.x = T)
    
    hap_site$t <- gsub("_", "", str_extract(hap_site$site_gen_plot, "_.*_"))
    
    hap_site$t[hap_site$site_gen_plot == "X3"] <- 0
    
    hap_site$t <- as.numeric(hap_site$t)
    
    print("Building GLMs...")
    mod_res <- hap_site %>% group_by(X1, X2) %>%  
      filter( (value <= 1) & (value >= 0) ) %>%
      nest() %>%
      mutate(
        # calculate intercept in same scale as logit
        #fixed_intercept = map_dbl(data, ~ as.numeric(.x[(.x$site_gen_plot == "X3"), "value"])),
        #fixed_intercept_logit = log(fixed_intercept/(1-fixed_intercept)),
        # Fit the model to each nested data frame (.x)
        #model = map(data, ~glm(value ~ t - 1 + offset(rep(fixed_intercept_logit, length(value))), data = .x, weights = n, family = quasibinomial(link = "logit"))),
        # fit glm, but don't force fixed intercept
        #model = map(data, ~glm(value ~ t, data = .x, weights = n, family = quasibinomial(link = "logit"))),
        model = map(data, ~safe_glm(.x)),
        # Extract coefficients in a tidy format
        tidied = map(model, tidy)
      ) %>%
      # Flatten the results back into a regular data frame
      unnest(tidied) %>% 
      filter(term == "t") %>%
      dplyr::select(c("X1", "X2", "estimate", "std.error", "statistic", "p.value"))
    
    print("Saving output...")
    mod_res$site = site # add site label
    mod_res$block = b # add site label
    write.table(mod_res, output_glm, col.names=!file.exists(output_glm), append = T, row.names = F, sep = "\t", quote = F)
    print(Sys.time())
  }
  gc()
}

```

# Simualte G(t) curves

## analyze simulations

```{r}
params <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/config/gt_params.tsv", sep = "\t", header = T, data.table = F)

# grab files of neutral allele freq trajectories
neutfreq_files <- list.files("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results", pattern = "*freqs.tsv", full.names = T)
#neutfreq_files <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results/", params %>% group_by(VA, L, SIGMA, adjust) %>% slice_sample(n = 5) %>% pull(ID), "_neutfreqs.tsv", sep = "")

#neutfreq_files <- sample(neutfreq_files, size = 1000, replace = F)

# files_to_plot <- c(
#   "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results/1_neutfreqs.tsv",
#   "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results/10_neutfreqs.tsv",
#   "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results/100_neutfreqs.tsv",
#   "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results/751_neutfreqs.tsv",
#   "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results/841_neutfreqs.tsv",
#   "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results/1000_neutfreqs.tsv",
#   "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results/1001_neutfreqs.tsv"
# )

sim_curves <- NULL

output_sim <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/", today, "/simulated_gt_curves_" , today, ".txt", sep = "")

check_file_exists(output_sim)

#P <- 300 # number of permutations to estimate drift absoluate covariances

for(neutfreq_file in neutfreq_files){
  simid <- as.numeric(gsub(".*/", "", gsub("_neutfreqs.tsv", "", neutfreq_file)))
  
  print(paste(Sys.time(),": Simulation ID: ", simid, sep = ""))
  
  no_col <- max(count.fields(neutfreq_file, sep = "\t"))

  neutfreqs <- fread(neutfreq_file, fill = T, col.names = paste("X",1:no_col, sep=""), sep = "\t", data.table = F)

  neutfreqs <- reshape2::melt(neutfreqs, id = "X1")

  neutfreqs$pos <- str_extract(neutfreqs$value, ";.*;")
  neutfreqs$freq <- str_extract(neutfreqs$value, ";0.*$")

  neutfreqs$pos <- as.numeric(gsub(";", "", neutfreqs$pos))
  neutfreqs$freq <- as.numeric(gsub(";", "", neutfreqs$freq))
  
  # calculate increments
  # neutfreqs <- neutfreqs %>%
  #   group_by(pos) %>%
  #   mutate(dp = freq - lag(freq))

  # Source - https://stackoverflow.com/a/5890689
  # Posted by Ista, modified by community. See post 'Timeline' for change history
  # Retrieved 2026-02-21, License - CC BY-SA 3.0
  # neutfreqs <- neutfreqs[,c("X1", "pos", "dp")]
  # neutfreqs <- neutfreqs[complete.cases(neutfreqs),]
  # neutfreqs <- neutfreqs[!duplicated(neutfreqs[,c("X1", "pos")]),]
  # dpmat <- reshape2::dcast(neutfreqs, pos ~ X1, value.var = "dp")
  # dpmat <- dpmat[,c(-1,-2)]
  # dpcov_mat <- cov(dpmat, use = "pairwise.complete.obs")
  # sim_curve <- gt_from_covmat(dpcov_mat)
  
  # pivot to longer, convert NULL list elements to NA, flatten lists
  # pmat_sim <- neutfreqs %>% 
  #   dplyr::select(pos, X1, freq) %>%
  #   pivot_wider(names_from = "X1", values_from = "freq") %>%
  # mutate(across(everything(), ~ {
  #   # 1. Replace empty or NULL list items with a standard NA
  #   cleaned_list <- map(., ~ if(is.null(.x) || length(.x) == 0) NA else .x[[1]])
  #   # 2. Flatten the list into a regular atomic vector
  #   flatten(cleaned_list)
  # })) %>%
  # unnest_longer(where(is.list), keep_empty = TRUE) %>%
  # arrange(pos)
  
  neutfreqs <- neutfreqs[,c("X1", "pos", "freq")]
  neutfreqs <- neutfreqs[complete.cases(neutfreqs),]
  neutfreqs <- neutfreqs[!duplicated(neutfreqs[,c("X1", "pos")]),]
  neutfreqs <- reshape2::dcast(neutfreqs, pos ~ X1, value.var = "freq")
  pmat_sim <- neutfreqs[,-1]

  covmat_sim <- covmat_from_pmat(pmat_sim, correct_for_n = F, input_asin_trans = F)
  
  sim_curve <- gt_from_covmat(covmat_sim)
  
  # create blocks
  # block_sim <- data.frame(
  #   chrom = 1,
  #   pos = 1:nrow(pmat_sim)
  # )
  # 
  # block_sim <- create_blocks(block_sim, chrom, pos, 1000, "_")
  # 
  # # block-based sign permutation
  # sim_perm_curves <- NULL
  # for(i in 1:P){
  #   perm_covmat_sim <- covmat_from_pmat(pmat_sim, n = NULL, correct_for_n = F, input_asin_trans = F, procedure = "window", windows = block_sim$window)
  # 
  #   sim_perm_curve <- gt_from_covmat(perm_covmat_sim)
  #   sim_perm_curve$perm <- i
  #   
  #   sim_perm_curves <- rbind(sim_perm_curves, sim_perm_curve)
  # }
  # 
  # sim_perm_curves <- sim_perm_curves %>%
  #   group_by(gen) %>%
  #   summarize(med_sum_abs_cov = median(sum_abs_cov),
  #             upr_sum_abs_cov = quantile(sum_abs_cov, 0.975))
  # 
  # sim_curve <- sim_curve %>%
  #   left_join(sim_perm_curves, by = "gen") %>%
  #   mutate(cor_sum_abs_cov = sum_abs_cov - upr_sum_abs_cov) 

  # if(neutfreq_file %in% files_to_plot){
  #   print("Plotting variance covariance matrix...")
  #   plot_var_cov_matrix(covmat_sim, paste("../results/", today, "/", simid, "_var_cov_mat.pdf", sep = ""), include_values = F)
  # }
  # 

  param_row <- params[(params$ID == simid),]
  
  sim_curve$ID <- simid
  sim_curve$L <- param_row[, "L"]
  sim_curve$SIGMA <- param_row[, "SIGMA"]
  sim_curve$VA <- param_row[, "VA"]
  sim_curve$N <- param_row[, "N"]
  sim_curve$adjust <- param_row[, "adjust"]
  
  #write.table(sim_curve, output, col.names=!file.exists(output), append = T, row.names = F, sep = ",", quote = F)
  
  append_table(sim_curve, output_sim)
}

```

# Temperature covariances

```{r}

load("../data/climate_sites_experiment.rda")
load("../data/climate_sites_yearly.rda")

# Generations 1-5 → years 2018-2022
# Transition 0->1 uses P0 (common seed stock, no site climate), skip it for climate
# Allele transitions: 0->1, 1->2, 2->3, 3->4, 4->5
# Climate transitions: 1->2, 2->3, 3->4, 4->5 (year-to-year at sites)
gen_year <- data.frame(generation = 1:5, year = 2018:2022)

# Get yearly TerraClimate temperature (bio1) per site
clim_annual <- climate_sites_yearly %>%
  mutate(site = as.numeric(site)) %>%
  dplyr::select(site, year, bio1) %>%
  filter(year >= 2018, year <= 2022) %>%
  left_join(gen_year, by = "year") %>%
  arrange(site, year)

cat(sprintf("TerraClimate data: %d sites, years %s\n",
            length(unique(clim_annual$site)),
            paste(sort(unique(clim_annual$year)), collapse = ", ")))

clim_increments <- clim_annual %>%
  group_by(site) %>%
  arrange(generation) %>%
  mutate(
    dT = bio1 - lag(bio1),
    interval = paste0(lag(generation), "->", generation)
  ) %>%
  filter(!is.na(dT)) %>%
  ungroup()

cat(sprintf("\nClimate increments: %d observations across %d sites\n",
            nrow(clim_increments), length(unique(clim_increments$site))))

# Build climate increment matrix: sites x intervals
clim_mat <- clim_increments %>%
  dplyr::select(site, interval, dT) %>%
  pivot_wider(id_cols = site, names_from = interval, values_from = dT)

clim_avail <- sort(setdiff(colnames(clim_mat), "site"))
clim_matrix <- as.matrix(clim_mat[, clim_avail])
clim_comp <- complete.cases(clim_matrix)
clim_matrix <- clim_matrix[clim_comp, ]

cat(sprintf("Sites with complete climate data: %d\n", nrow(clim_matrix)))

# Q_climate: covariance of climate increments across sites
Q_climate <- cov(clim_matrix)

# Correlation version
R_climate <- cor(clim_matrix)

plot_var_cov_matrix(R_climate, paste("../results/", today, "/climate_increment_correlations.pdf", sep = ""), legend_name = "Mean\ncor(\u0394Ti, \u0394Tj)", low_col = "#023198", mid_col = "white", high_col = "#7E1700")

# see if covariant allele frequency trajectories follow covariant temperate trajectories, by site
#mean_dp_cov_by_site <- as.data.frame(melted_cov_dps %>% group_by(site, Var1, Var2) %>% summarize(mean_cov = mean(value)) %>% filter(!str_detect(Var1, "0") & !str_detect(Var2, "0")))

mean_dp_cov_by_site <- as.data.frame(melted_cov_dps %>% filter(!str_detect(Var1, "0") & !str_detect(Var2, "0")))

mean_dp_cov_by_site <- mean_dp_cov_by_site[!duplicated(mean_dp_cov_by_site$value),]

# observation
obs_dp_dt_match <- check_dp_match_dt(mean_dp_cov_by_site, clim_increments)

# permutation test
permuted_dp <- permute(mean_dp_cov_by_site, n = 1000, value)

permuted_dt <- permute(clim_increments, n = 1000, dT)

list_of_dps <- permuted_dp$perm %>% 
  map(as.data.frame)

list_of_dts <-permuted_dt$perm %>% 
  map(as.data.frame)

null_dist <- mapply(check_dp_match_dt, x = list_of_dps, y = list_of_dts)

null_dist <- as.data.frame(t(null_dist))

ggplot(null_dist, aes(x = `TRUE`/(`TRUE`+`FALSE`))) +
  geom_density(lwd = 2) +
  geom_vline(xintercept = obs_dp_dt_match["TRUE"]/sum(obs_dp_dt_match), lwd = 2, color = "red",) +
  labs(x = "Proportion of sgn Cov(\u0394pi, \u0394pj) = sgn \u0394Ti\u0394Tj", title = "Permuted null")

ggsave(paste("../results/", today, "/linked_selection_cov_matches_bio1.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)

# Compute anomalies
# "anomaly" is bio1 - average bio1 (residual bio1), which means there will by definition be some positive and some negative anomalies at each site
clim_anom <- clim_annual %>%
  group_by(site) %>%
  mutate(site_mean = mean(bio1),
         anomaly = bio1 - site_mean) %>%
  ungroup() %>%
  left_join(climate_sites %>% mutate(site = as.numeric(site)) %>%
              dplyr::select(site, site_bio1), by = "site")

# Anomaly heatmap: sites (sorted by temperature) x years
anom_wide <- clim_anom %>%
  dplyr::select(site, year, anomaly, site_bio1) %>%
  arrange(site_bio1)

p_anom_heat <- anom_wide %>%
  mutate(site = factor(site, levels = unique(site[order(site_bio1)]))) %>%
  ggplot(aes(x = factor(year), y = site, fill = anomaly)) +
  geom_tile(color = "grey90") +
  geom_text(aes(label = sprintf("%.1f", anomaly)), size = 2.5) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, name = "Anomaly\n(°C)") +
  labs(title = "Temperature anomaly per site per year (TerraClimate)",
       subtitle = "Sites sorted by mean temperature (cold at bottom, warm at top)",
       x = "Year (generation)", y = "Site") +
  theme_minimal(base_size = 11) +
  theme(axis.text.y = element_text(size = 6))

# Mean anomaly across sites per year
mean_anom <- clim_anom %>%
  group_by(year) %>%
  summarise(mean_anomaly = mean(anomaly), se = sd(anomaly) / sqrt(n()), .groups = "drop")

p_anom_bar <- mean_anom %>%
  ggplot(aes(x = factor(year), y = mean_anomaly, fill = mean_anomaly > 0)) +
  geom_col(alpha = 0.7, width = 0.6) +
  geom_errorbar(aes(ymin = mean_anomaly - 1.96*se, ymax = mean_anomaly + 1.96*se), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "blue"), guide = "none") +
  labs(title = "Mean temperature anomaly across sites",
       subtitle = "Warm/cold years drive allele frequency reversals",
       x = "Year (generation)", y = "Mean anomaly (°C)") +
  theme_minimal(base_size = 12)

p_anom_heat

p_anom_bar

# compare anomalies to bioclim
anom_vs_bioclim <- clim_annual %>%
  left_join(climate_sites %>% mutate(site = as.numeric(site)) %>%
              dplyr::select(site, site_bio1), by = "site") %>%
  mutate(anomaly_bioclim = bio1 - site_bio1)

cat("Anomaly = TerraClimate bio1 - bioclim baseline (site_bio1)\n")

# Build anomaly matrix: sites x years
anom_wide_bioclim <- anom_vs_bioclim %>%
  dplyr::select(site, year, anomaly_bioclim) %>%
  pivot_wider(id_cols = site, names_from = year, values_from = anomaly_bioclim)

anom_matrix <- as.matrix(anom_wide_bioclim[, -1])
rownames(anom_matrix) <- anom_wide_bioclim$site

# Covariance and correlation of anomalies across sites
Cov_anom <- cov(anom_matrix)
Cor_anom <- cor(anom_matrix)

cat("\nAnomaly covariance matrix (TerraClim - bioclim baseline):\n")

plot_var_cov_matrix(Cor_anom, paste("../results/", today, "/climate_anomalies_correlations.pdf", sep = ""), legend_name = "Mean\ncor(Ai, Aj)", low_col = "#023198", mid_col = "#C0E9C2", high_col = "#7E1700")

# try cmip6 model
bbox <- get_1km_bbox(37.8715, -122.2730)

area <- c(
  bbox$xmin,
  bbox$xmax,
  bbox$ymin,
  bbox$ymax
)

download_cmip6_ecmwfr(
  scenario = "ssp126",
  area = area,
  model = "default",
  service = "cds",
  frequency = "monthly",
  variable = c("Tmin", "Tmax"),
  year_start = 2000,
  year_end = 2015,
  month = 1:12,
  sec_wait = 3600,
  n_try = 10,
  update_everything = FALSE,
  path_download = "cmip6_downloaded"
)


```


# G(t) and Gabs(t)

## arcsin sqrt transformed frequencies

### point estimates

```{r}

# make list of sites
sites <- unique(gsub("_.*", "", colnames(hapfire)))

# for now, ignore some sites
sites <- sites[c(-30)]

output_gt_points <-  paste(
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/",
  today,
  "/gt_point_estimates.tsv",
  sep = ""
)

for(site in unique(sites)){
  print(paste("Site:", site))
  hap_site <- hapfire %>% 
    dplyr::select(starts_with(paste(site, "_", sep = ""))) %>%
    collect()
  
  for(exprep in 1:12){
    print(paste("Rep:", exprep))
    hap_rep <- hap_site %>% 
      dplyr::select(ends_with(paste("_", exprep, sep = "")))
    
    # sort names to be in order of generation
    hap_rep <- hap_rep %>% relocate(all_of(sort(colnames(hap_rep))))
    
    if(ncol(hap_rep) <= 2){
      print("Need at least two increments. Skipping to next replicate...")
      next
    }
    
    rep_n <- grab_sample_sizes(sample_sizes, site_gen_plot, colnames(hap_rep), 10000)
    
    hap_rep <- cbind(gen0$X3, hap_rep)
    colnames(hap_rep)[1] = paste(site, "_0_", exprep, sep = "")
    
    covmat <- covmat_from_pmat(hap_rep, rep_n$n, correct_for_n = T, input_asin_trans = F, procedure = "none")
    
    gt_result <- gt_from_covmat(covmat)
    
    gt_result$site<-site
    gt_result$rep<-exprep
    
    append_table(gt_result, output_gt_points)
    #gt_results <- rbind(gt_results, gt_result)
  }
}

gt_abs_mean <- gt_results %>%
  group_by(site, gen) %>%
  summarize(gt = mean(sum_cov)/mean(sum_cov+sum_var),
            gt_avg_ratio = mean(sum_cov/(sum_cov+sum_var)),
            gt_abs = mean(sum_abs_cov)/mean(sum_abs_cov+sum_var),
            gt_abs_2 = mean(sum_abs_cov)/mean(sum_cov+sum_var),
            gt_pos_abs = mean(sum_pos_cov)/mean(sum_abs_cov+sum_var),
            gt_neg_abs = mean(sum_neg_cov)/mean(sum_abs_cov+sum_var),
            cov_ratio = mean(sum_pos_cov)/mean(sum_abs_cov))
  
gt_abs_mean <- merge(gt_abs_mean, worldclim_sitesdata, by = "site")

ggplot(gt_abs_mean, aes(x = gen, y = gt_abs, group = site, color = bio1)) +
  geom_point() +
  geom_line() +
  #geom_errorbar(aes(ymin=lwr, ymax=upr), width=.2) +
  #geom_ribbon(data = gabs_null, mapping = aes(x=gen, ymin=lwr, ymax=upr), alpha = 0.2, color = "grey", size = 0.01) + 
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:5)) +
  labs(x = "Generation", y = expression(G[abs]), color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggplot(gt_abs_mean, aes(x = gen, y = gt_pos_abs, group = site, color = bio1)) +
  geom_point() +
  geom_line() +
  #geom_errorbar(aes(ymin=lwr, ymax=upr), width=.2) +
  #geom_ribbon(data = gabs_null, mapping = aes(x=gen, ymin=lwr, ymax=upr), alpha = 0.2, color = "grey", size = 0.01) + 
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:5)) +
  labs(x = "Generation", y = expression(G[abs]), color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggplot(gt_abs_mean, aes(x = gen, y = gt_neg_abs, group = site, color = bio1)) +
  geom_point() +
  geom_line() +
  #geom_errorbar(aes(ymin=lwr, ymax=upr), width=.2) +
  #geom_ribbon(data = gabs_null, mapping = aes(x=gen, ymin=lwr, ymax=upr), alpha = 0.2, color = "grey", size = 0.01) + 
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:5)) +
  labs(x = "Generation", y = expression(G[abs]), color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

gt_abs_mean$cov_ratio[is.nan(gt_abs_mean$cov_ratio) & (gt_abs_mean$gen == 1)] <- 0.5
ggplot(gt_abs_mean, aes(x = gen, y = cov_ratio, group = site, color = bio1)) +
  geom_point() +
  geom_line() +
  geom_hline(linetype = 2, color = "black", lwd = 1.5, yintercept = 0.5) +
  #geom_errorbar(aes(ymin=lwr, ymax=upr), width=.2) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:5)) +
  labs(x = "Generation", y = "Propotion Cov > 0", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")
```

### bootstrap

```{r}
sign_permute_blocks <- create_blocks(gen0, X1, X2, 15000, "_")
#hapfire$window <- sign_permute_blocks$window

W <- length(unique(sign_permute_blocks$window))
B <- 200

# define output
output_gt_boot <-  paste(
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/",
  today,
  "/gt_bootstraps.tsv",
  sep = ""
)

#check_file_exists(output_gt_boot)

# loop over sites
# make list of sites, ignoring some
sites <- unique(gsub("_.*", "", colnames(hapfire)))
sites <- sites[c(-30)]

for(site in unique(sites)[26:30]){
  print(paste("Site:", site))
  hap_site <- hapfire %>% 
    dplyr::select(starts_with(paste(site, "_", sep = ""))) %>%
    collect()
  
  for(exprep in 1:12){
    print(paste("Rep:", exprep))
    hap_rep <- hap_site %>% 
      dplyr::select(ends_with(paste("_", exprep, sep = "")))
    
    # sort names to be in order of generation
    hap_rep <- hap_rep %>% relocate(all_of(sort(colnames(hap_rep))))
    
    if(ncol(hap_rep) <= 2){
      print("Need at least two increments. Skipping to next replicate...")
      next
    }
    
    rep_n <- grab_sample_sizes(sample_sizes, site_gen_plot, colnames(hap_rep), 10000)
    
    hap_rep <- cbind(gen0$X3, hap_rep)
    colnames(hap_rep)[1] = paste(site, "_0_", exprep, sep = "")
    
    hap_rep$window <- sign_permute_blocks$window
    # bootstrap sample allele frequencies by window
    nested <- hap_rep %>% nest_by(window) %>% ungroup()
    
    for (b in 1:B){
      gt_boot <- nested %>%
        slice_sample(n = W, replace = TRUE) %>%
        unnest(cols = c(data)) %>%
        dplyr::select(-window) %>%
        as.data.frame() %>%
        covmat_from_pmat(data, 
                         n = rep_n$n, 
                         correct_for_n = T, 
                         input_asin_trans = F, 
                         procedure = "none") %>%
        gt_from_covmat()
      
      gt_boot$site <- site
      gt_boot$rep <- exprep
      gt_boot$boot <- b
      append_table(gt_boot, output_gt_boot)
    }
  }
}

gt_boot_results <- fread(output_gt_boot) %>% 
  distinct(site, rep, gen, boot, .keep_all = T)

gt_boot_rep_cis <- gt_boot_results %>%
  group_by(site, rep, gen) %>% 
  summarize(
    lwr = quantile(sum_cov, 0.025, na.rm = T),
    med = median(sum_cov, na.rm = T),
    upr = quantile(sum_cov, 0.975, na.rm = T)
  )

gt_boot_mean_cis <- gt_boot_results %>%
  group_by(site, gen, boot) %>% # average across replicates within bootstrap
  summarize(sum_cov = mean(sum_cov)) %>%
  summarize(
    lwr = quantile(sum_cov, 0.025, na.rm = T),
    med = median(sum_cov, na.rm = T),
    upr = quantile(sum_cov, 0.975, na.rm = T)
  )

ggplot(gt_results, aes(x = gen, y = sum_cov, group = interaction(site, rep))) +
  geom_point() +
  geom_line() +
  geom_hline(yintercept = 0, lwd = 0.5, lty = 2, color = "black") +
  #facet_wrap(.~site) +
  geom_errorbar(mapping = aes(y = med, ymin = lwr, ymax = upr), data = gt_boot_cis, width = 0.2)
```

### sign permutation

```{r}
# make list of sites
# for now, ignore some sites
sites <- unique(gsub("_.*", "", colnames(hapfire)))
sites <- sites[c(-30)]

# permuation parameters
sign_permute_replicates <- 300
sign_permute_blocks <- create_blocks(gen0, X1, X2, 15000, "_")

# define output
output_gt_perm <- paste(
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/",
  today,
  "/gt_permutations.tsv",
  sep = ""
)

#check_file_exists(output_gt_perm)

# loop over sites
for (site in unique(sites)[29:30]) {
  print(paste("Site:", site))
  hap_site <- hapfire %>% 
    dplyr::select(starts_with(paste(site, "_", sep = ""))) %>%
    collect()
  
  for (exprep in 1:12) {
    print(paste("Rep:", exprep))
    hap_rep <- hap_site %>%
      dplyr::select(ends_with(paste("_", exprep, sep = "")))
    
    # sort names to be in order of generation
    hap_rep <- hap_rep %>%
      relocate(all_of(sort(colnames(hap_rep))))
    
    if (ncol(hap_rep) <= 2) {
      print("Need at least two increments. Skipping to next replicate...")
      next
    }
    
    rep_n <- grab_sample_sizes(sample_sizes, site_gen_plot, colnames(hap_rep), 10000)
    
    hap_rep <- cbind(gen0$X3, hap_rep)
    colnames(hap_rep)[1] = paste(site, "_0_", exprep, sep = "")
    
    for (p in 1:sign_permute_replicates) {
      covmat <- covmat_from_pmat(
        hap_rep,
        rep_n$n,
        correct_for_n = T,
        input_asin_trans = F,
        procedure = "window",
        windows = sign_permute_blocks$window
      )
      
      gt_perm_result <- gt_from_covmat(covmat)
      
      gt_perm_result$site = site
      gt_perm_result$rep = exprep
      gt_perm_result$perm = p
      
      append_table(gt_perm_result, output_gt_perm)
      #gt_perm_results <- rbind(gt_perm_results, gt_perm_result)
    }
  }
}

gt_perm_results <- fread(output_gt_perm)  %>% 
  distinct(site, rep, gen, perm, .keep_all = T)

gt_perm_rep <- gt_perm_results %>%
  group_by(site, gen, rep) %>%
  summarize(lwr = quantile(sum_cov, 0.025), # get quantiles across permutations
            med = median(sum_cov),
            upr = quantile(sum_cov, 0.975))

gt_perm_mean <- gt_perm_results %>%
  group_by(site, gen, perm) %>% # average across replicates within permutation
  summarize(sum_cov = mean(sum_cov)) %>%
  summarize(lwr = quantile(sum_cov, 0.025), # get quantiles across permutations
            med = median(sum_cov),
            upr = quantile(sum_cov, 0.975))





gt_perm_results %>%
  group_by(site, gen, perm) %>% # average across replicates within permutation
  summarize(cov_ratio = mean(sum_pos_cov) / mean(sum_abs_cov)) %>%
  summarize(
    lwr = quantile(cov_ratio, 0.025, na.rm = T),
    med = median(cov_ratio, na.rm = T),
    upr = quantile(cov_ratio, 0.975, na.rm = T)
  )

gt_perm_results %>%
  group_by(site, gen, perm) %>% # average across replicates within permutation
  summarize(gt_abs = mean(sum_abs_cov) / mean(sum_abs_cov + sum_var)) %>%
  summarize(
    lwr = quantile(gt_abs, 0.025),
    med = median(gt_abs),
    upr = quantile(gt_abs, 0.975)
  )

gt_perm_results %>%
  group_by(site, gen, perm) %>% # average across replicates within permutation
  summarize(gt = mean(sum_cov) / mean(sum_cov + sum_var)) %>%
  summarize(lwr = quantile(gt, 0.025),
            med = median(gt),
            upr = quantile(gt, 0.975))

gt_perm_results %>%
  group_by(site, gen, perm) %>% # average across replicates within permutation
  summarize(gt_pos_abs = mean(sum_pos_cov) / mean(sum_abs_cov + sum_var)) %>%
  summarize(lwr = quantile(gt_pos_abs, 0.025),
            med = median(gt_pos_abs),
            upr = quantile(gt_pos_abs, 0.975))

gt_perm_results %>%
  group_by(site, gen, perm) %>% # average across replicates within permutation
  summarize(gt_abs_2 = mean(sum_abs_cov) / mean(sum_cov + sum_var)) %>%
  summarize(lwr = quantile(gt_abs_2, 0.025),
            med = median(gt_abs_2),
            upr = quantile(gt_abs_2, 0.975))
```

### compare bootstrap with permutation null

```{r}

points <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-06-17/gt_point_estimates.tsv") %>%
  left_join(select(locations_data, site, city_country), by = "site")

perms <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-06-17/gt_permutations.tsv") %>% 
  distinct(site, rep, gen, perm, .keep_all = T)

boots <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-06-17/gt_bootstraps.tsv", colClasses = "numeric") %>% 
  distinct(site, rep, gen, boot, .keep_all = T) %>%
  mutate_if(is.character,as.numeric)

# average absolute cov sum under permutation
# perms <- perms %>%
#   group_by(site, rep, gen) %>%
#   summarize(perm_lwr_sum_abs_cov = quantile(sum_abs_cov, 0.025),
#             perm_mean_sum_abs_cov = mean(sum_abs_cov),
#             perm_upr_sum_abs_cov = quantile(sum_abs_cov, 0.975),
#             perm_lwr_gt_abs = quantile(sum_abs_cov/(), 0.025),
#             perm_mean_gt_abs = mean(sum_abs_cov),
#             perm_upr_gt_abs = quantile(sum_abs_cov, 0.975))

perms_summary <- perms %>%
  group_by(site, gen, rep) %>%
  summarize(sum_abs_cov_perm_lwr = quantile(sum_abs_cov, 0.025),
    sum_abs_cov_perm_upr = quantile(sum_abs_cov, 0.975),
    sum_abs_cov = NA,
    gt_abs_perm_lwr = quantile(gt_abs, 0.025),
    gt_abs_perm_upr = quantile(gt_abs, 0.975),
    gt_abs = NA) %>%
  left_join(select(locations_data, site, city_country), by = "site")

boots_summary <- boots %>%
  group_by(site, gen, rep) %>%
  summarize(sum_abs_cov_boot_lwr = quantile(sum_abs_cov, 0.025, na.rm = T),
    sum_abs_cov_boot_upr = quantile(sum_abs_cov, 0.975, na.rm = T),
    sum_abs_cov = NA,
    gt_abs_boot_lwr = quantile(gt_abs, 0.025, na.rm = T),
    gt_abs_boot_upr = quantile(gt_abs, 0.975, na.rm = T),
    gt_abs = median(gt_abs, na.rm = T)) %>%
  left_join(select(locations_data, site, city_country), by = "site") %>%
  left_join(worldclim_sitesdata, by = "site")

## plot sum of covariance through time
ggplot(points, aes(x = gen, y = sum_abs_cov, group = interaction(site,rep))) +
  geom_point(shape = 16) +
  geom_ribbon(data = perms_summary, mapping = aes(x=gen, ymin=sum_abs_cov_perm_lwr, ymax=sum_abs_cov_perm_upr), alpha = 0.2, color = "grey", size = 0.01) +
  facet_wrap(.~city_country+site, scales = "free_y") +
  geom_line() +
  geom_errorbar(data = boots_summary, mapping = aes(x = gen, ymin = sum_abs_cov_boot_lwr, ymax = sum_abs_cov_boot_upr), width = 0.1, color = "red")

ggsave(paste(results_path, "sum_abs_cov_perm_boot.pdf", sep = ""), width = 13, height = 9)

# plot G(t) per replicate through time
ggplot(boots_summary, aes(x = gen, y = gt_abs, group = interaction(site, rep), color = bio1)) +
  geom_point(shape = 16) +
  geom_line() +
  geom_errorbar(data = boots_summary, mapping = aes(x = gen, ymin = gt_abs_boot_lwr, ymax = gt_abs_boot_upr), width = 0.1, color = "red") +
  geom_ribbon(data = perms_summary, mapping = aes(x=gen, ymin=gt_abs_perm_lwr, ymax=gt_abs_perm_upr), alpha = 0.2, color = "grey", size = 0.01) +
  facet_wrap(.~city_country+site, scales = "free_y") +
  scale_color_scico(palette = "berlin") +
  labs(x = "Generation", y = expression(G[abs]), color = "Mean\nAnnual\nTemp.\n(\u00B0C)")


ggsave(paste(results_path, "gt_abs_perm_boot.pdf", sep = ""), width = 12, height = 9)



boots_summary <- boots %>%
  # group_by(site, gen, rep, boot) %>%
  # mutate(gt_abs = (sum_abs_cov)/(sum_abs_cov+sum_var)) %>%
  group_by(site, gen, rep) %>%
  summarize(gt_abs_boot_lwr = quantile(gt_abs, 0.025, na.rm = T),
    gt_abs_boot_upr = quantile(gt_abs, 0.975, na.rm = T),
    gt_abs = NA)

ggplot(points_summary, aes(x = gen, y = gt_abs, group = site)) +
  geom_point() +
  geom_errorbar(data = boots_summary, mapping = aes(x = gen, ymin = gt_abs_boot_lwr, ymax = gt_abs_boot_upr), width = 0.1) +
  geom_ribbon(data = perms_summary, mapping = aes(x=gen, ymin=gt_abs_perm_lwr, ymax=gt_abs_perm_upr), alpha = 0.2, color = "grey", size = 0.01) + 
  geom_line()



# subtract off permutation null
points_cor <- points %>% 
  left_join(perms, by = c("site", "rep", "gen")) %>%
  mutate(cor_sum_abs_cov = sum_abs_cov - perm_upr_sum_abs_cov) %>%
  group_by(site, gen) %>%
  summarize(gt_abs_cor = mean(cor_sum_abs_cov)/mean(sum_abs_cov + sum_var))

boots_cor <- boots %>%
  left_join(perms, by = c("site", "rep", "gen")) %>%
  mutate(cor_sum_abs_cov = sum_abs_cov - perm_upr_sum_abs_cov) %>%
  group_by(site, gen, boot) %>%
  summarize(gt_abs_cor = mean(cor_sum_abs_cov, na.rm = T)/mean(sum_abs_cov + sum_var, na.rm = T)) %>%
  group_by(site, gen) %>%
  summarize(gt_abs_cor_lwr = quantile(gt_abs_cor, 0.025, na.rm = T),
            gt_abs_cor_upr = quantile(gt_abs_cor, 0.975, na.rm = T))
  
points_cor %>% 
  left_join(boots_cor, by = c("site", "gen"))


```


## raw frequencies

### calculate frequency increments

```{r}
# make list of sites
sites <- unique(gsub("_.*", "", colnames(hapfire)))

# for now, ignore sites that do not have at least 2 incremenent
#sites <- sites[c(-11, -14, -18, -19, -30, -31)]
sites <- sites[-30]

# loop over sites
pdiffs <- NULL
gc()

for(site in sites) {
  print(paste("Site:", site))
  for (i in 1:12) {
    print(paste("Rep:", i))
    hap_rep_i <- hapfire %>% dplyr::select(starts_with(paste(site, "_", sep = "")) & ends_with(paste("_", i, sep = "")))
    # if (ncol(hap_rep_i) <= 1) {
    #   print("Need at least two increments. Skipping to next replicate...")
    #   n_valid_reps <- n_valid_reps - 1
    #   next
    # }
    
    # get sample sizes
    # n_i <- sample_sizes[(sample_sizes$site_gen_plot %in% colnames(hap_rep_i)), ]
    # n_i <- rbind(c(0, 1e5), n_i)
    # n_i <- n_i %>% arrange(site_gen_plot)
    
    # sort names to be in order of generation
    hap_rep_i <- hap_rep_i %>% relocate(all_of(sort(colnames(hap_rep_i))))
    # create matrix of allele frequencies
    # L X G
    pmat_i <- NULL
    pmat_i <- cbind(pmat_i, gen0$X3)
    colnames(pmat_i)[1] <- "0"
    pmat_i <- cbind(pmat_i, hap_rep_i)
    #colnames(pmat_i)[2:ncol(pmat_i)] <- colnames(hap_rep_i)
    
    # for (w in 1:ncol(hap_rep_i)) {
    #   pmat_i <- cbind(pmat_i, hap_rep_i[, w])
    #   colnames(pmat_i)[w + 1] <- colnames(hap_rep_i)[w]
    # }
    
    # calculate increments
    pdiff_i <- as.matrix(pmat_i[, -1] - pmat_i[, -ncol(pmat_i)])
    colnames(pdiff_i) <- colnames(hap_rep_i)
    pdiffs <- cbind(pdiffs, pdiff_i)
  }
}

# save increment matrix
write.csv(pdiffs, "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/allele_frequency_increments.csv", row.names = F, quote = F)
  
```

### calculate covariances

```{r}
if(!exists("pdiffs")){
  pdiffs <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/allele_frequency_increments.csv", header = T)
}

# save un-corrected covariance matrices
cov_pdiffs <- cov(pdiffs)

write.table(cov_pdiffs, "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/increment_covariances.tsv",row.names = T, col.names = T, quote = F, sep = "\t")
```

### correct covariances for sample size

```{r}
# load covariance matrix as matrix object
if(!exists("cov_pdiffs")){
  cov_pdiffs <- as.matrix(read.table("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/raw_increment_covariances.txt", header = TRUE, row.names = 1))
  colnames(cov_pdiffs) <- gsub("X", "", colnames(cov_pdiffs))
}

# make list of sites
sites <- unique(gsub("_.*", "", colnames(hapfire)))

# for now, ignore sites that lack reps with contiguous generations
#sites <- sites[c(1,3,5:17,19:21,23,25:29,31)]
sites <- sites[-30]

# rows and replicates
rows_cov_pdiffs <- rownames(cov_pdiffs)
cols_cov_pdiffs <- colnames(cov_pdiffs)

# create new matrix to store bias-corrected cov matrices
unbiased_cov_pdiffs <- cov_pdiffs

# loop over sites
for(site in sites) {
  print(paste("Site:", site))
  for(exprep in 1:12) {
    print(paste("Rep:", exprep))
    # subset to one site
    hap_rep <- hapfire %>% select(starts_with(paste(site, "_", sep = "")) & ends_with(paste("_", exprep, sep = "")))
    
    if(ncol(hap_rep) <= 1){
      print("Need at least two increments. Skipping to next replicate...")
      next
    }
    
    # get sample sizes
    rep_n <- sample_sizes[(sample_sizes$site_gen_plot %in% colnames(hap_rep)),]
    rep_n <- rbind(c(0, 1e5), rep_n)
    rep_n <- rep_n %>% arrange(site_gen_plot)
    
    # sort names to be in order of generation
    hap_rep <- hap_rep %>% relocate(all_of(sort(colnames(hap_rep))))
    
    # create matrix of allele frequencies
    # L X G
    pmat <- NULL
    pmat <- cbind(pmat, gen0$X3)
    colnames(pmat)[1] <- "0"
    for(i in 1:ncol(hap_rep)){
      pmat <- cbind(pmat, hap_rep[,i])
      colnames(pmat)[i+1] <- colnames(hap_rep)[i]
    }
    
    # subset covariance matrix
    covmat <- cov_pdiffs[grep(paste("^", site, "_.*_", exprep, "$", sep = ""), rows_cov_pdiffs), grep(paste("^", site, "_.*_", exprep, "$", sep = ""), cols_cov_pdiffs)]
    
    # correct covariance matrix for sample size
    correct_covmat <- correct_covmat_for_n(covmat, pmat, rep_n$n)
    
    unbiased_cov_pdiffs[grep(paste("^", site, "_.*_", exprep, "$", sep = ""), rows_cov_pdiffs), grep(paste("^", site, "_.*_", exprep, "$", sep = ""), cols_cov_pdiffs)] <- correct_covmat
  }
}

# check that correction behaved as expected
boolean_cov_pdiffs <- (unbiased_cov_pdiffs > cov_pdiffs)

# save corrected covariance matrix
write.table(cov_pdiffs, "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/unbiased_increment_covariances.tsv", row.names = T, col.names = T, quote = F, sep = "\t")
```

### block covariances across all replicates

```{r}
pdiffs <- as.data.frame(pdiffs)
if(!exists("pdiffs")){
  pdiffs <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/allele_frequency_increments.csv", header = T)
}

# if(!exists("cov_pdiffs")){
#   cov_pdiffs <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/unbiased_increment_covariances.tsv", header = T, row.names = 1))
# 
#   colnames(cov_pdiffs) <- gsub("X", "", colnames(cov_pdiffs))
# }

# number of snps per window
window_size <- 15000

# create windows, should not bleed between chromosomes
blocks <- gen0 %>%
  group_by(X1) %>%
  mutate(block = (row_number() - 1) %/% window_size + 1,
         window = paste(X1, block, sep = "_"))

length(unique(blocks$window))

pdiffs$window <- blocks$window

conv_gt_results <- NULL

for(w in unique(pdiffs$window)){
  print(paste("Window:", w))
  pdiff_w <- pdiffs %>% filter(window == w) %>% select(-window)
  
  # increment covariances
  cov_pdiffs <- cov(pdiff_w, use = "pairwise.complete.obs")
  
  # covariances within a rep
  row_rep <- gsub("_.*_", "", rownames(cov_pdiffs))
  col_rep <- gsub("_.*_", "", colnames(cov_pdiffs))
  
  row_site <- gsub("_.*", "", rownames(cov_pdiffs))
  col_site <- gsub("_.*", "", colnames(cov_pdiffs))
  
  row_gen <- gsub("_", "", str_extract(rownames(cov_pdiffs), "_.*_"))
  col_gen <- gsub("_", "", str_extract(colnames(cov_pdiffs), "_.*_"))
  
  total_var <- 0
  nreps <- 0
  for(k in unique(row_rep)){
    cov_k <- cov_pdiffs[k == row_rep, k == col_rep]
    # TO DO: correct covariances within reps by sample size
    total_var <- total_var + sum(cov_k)
    nreps <- nreps + 1
  }
  
  # covariances across reps
  total_covar <- 0
  for(k in unique(row_site)){
    for(g in unique(row_gen)){
      cov_k <- cov_pdiffs[(k == row_site) & (g == row_gen), (k != col_site) & (g != col_gen)]
      total_covar <- total_covar + sum(cov_k)
    }
  }
  
  # save results
  # npairs is actually number of pairs*2 because I'm double-counting covariances above
  npairs = ((length(unique(row_site))*length(unique(row_gen)))^2 - length(unique(row_site))*length(unique(row_gen)))
    
  conv_gt_result = data.frame(window = w,
                      S = nrow(pdiff_w),
                      nreps = nreps,
                      npairs = npairs,
                      total_covar = total_covar,
                      total_var = total_var,
                      mean_covar = total_covar/npairs,
                      mean_var = total_var/nreps)
  
  conv_gt_results <-  rbind(conv_gt_results, conv_gt_result)
}

# save results
write.table(conv_gt_results, paste(results_path, "block_conv_gt_for_bootstrapping.txt", sep = ""), row.names = F, quote = F, sep = "\t")

```

### make confidence intervals

```{r}
B = 1000 # number of bootstrap samples per site
W = 200 # number of windows in the genome

#cis <- NULL

boot_samples = NULL
for(b in 1:B){
  boot_sample <- conv_gt_results %>%
    slice_sample(n = W, replace = TRUE) %>%  # Sample groups with replacement
    summarize(
      gt_shared = weighted.mean(mean_covar, S)/weighted.mean(mean_var, S)
    )
  boot_samples <- rbind(boot_samples, boot_sample)
}

boot_samples %>% summarize(lwr = quantile(gt_shared, 0.025), med = median(gt_shared), upr = quantile(gt_shared, 0.975))

```

### covariances across pairs of environments

```{r}
if(!exists("cov_pdiffs")){
  cov_pdiffs <- as.matrix(read.table("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/unbiased_increment_covariances.tsv", header = TRUE, row.names = 1))
  colnames(cov_pdiffs) <- gsub("X", "", colnames(cov_pdiffs))
}

sites <- unique(gsub("_.*", "", colnames(cov_pdiffs)))

row_site <- gsub("_.*", "", rownames(cov_pdiffs))
col_site <- gsub("_.*", "", colnames(cov_pdiffs))

row_rep <- gsub("_.*_", "", rownames(cov_pdiffs))
col_rep <- gsub("_.*_", "", colnames(cov_pdiffs))

pair_gt_results = NULL

for(i in 1:(length(sites) - 1)){
  a = sites[i]
  for(j in (i+1):length(sites)){
    
    b = sites[j]
    
    print(paste("Sites:", i, j))
    
    # subset cov to site pair
    cov_pair <- cov_pdiffs[row_site == a, col_site == b]
    
    # remove overlapping increments
    sub_row_rep <- gsub("_.*_", "", rownames(cov_pair))
    sub_col_rep <- gsub("_.*_", "", colnames(cov_pair))
  
    sub_row_site <- gsub("_.*", "", rownames(cov_pair))
    sub_col_site <- gsub("_.*", "", colnames(cov_pair))
  
    sub_row_gen <- gsub("_", "", str_extract(rownames(cov_pair), "_.*_"))
    sub_col_gen <- gsub("_", "", str_extract(colnames(cov_pair), "_.*_"))
    
    # covariances across reps, non-overlapping increments
    total_covar <- 0
    for(k in colnames(cov_pair)){
      col_gen <- gsub("_", "", str_extract(k, "_.*_"))
      cov_k <- cov_pair[sub_row_gen != col_gen, k]
      total_covar <- total_covar + 2*sum(cov_k)
    }
    
    a_b_reps <- unique(c(sub_row_rep, sub_col_rep))
    total_var <- 0
    nreps <- 0
    for(k in a_b_reps){
      cov_k <- cov_pdiffs[k == row_rep, k == col_rep]
      total_var <- total_var + sum(cov_k)
      nreps <- nreps + 1
    }
    
    npairs = length(unique(sub_row_rep))*length(unique(sub_col_rep))
    
    pair_gt_result = data.frame(site_a = a,
                                site_b = b,
                      nreps = nreps,
                      npairs = npairs,
                      total_covar = total_covar,
                      total_var = total_var,
                      mean_covar = total_covar/npairs,
                      mean_var = total_var/nreps
                      )
    
    pair_gt_results = rbind(pair_gt_results, pair_gt_result)
    
  }
}

pair_gt_results$ge_t <- pair_gt_results$mean_covar/pair_gt_results$mean_var

pair_gt_results_clim <- merge(pair_gt_results, worldclim_sitesdata, by.x = "site_a", by.y = "site")

pair_gt_results_clim <- merge(pair_gt_results_clim, worldclim_sitesdata, by.x = "site_b", by.y = "site", suffixes = c("_a", "_b"))

pair_gt_results_clim$delta_bio1 <- (pair_gt_results_clim$bio1_a - pair_gt_results_clim$bio1_b)

ggplot(pair_gt_results_clim, aes(x = delta_bio1, y = ge_t)) +
  geom_point()
```

### sign permute allele frequency changes in blocks

```{r}

if(!exists("pdiffs")){
  pdiffs <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/allele_frequency_increments.csv", header = T)

  #pdiffs <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/allele_frequency_increments.csv", header = T, select = c("1_1_1", "1_2_1", "1_3_1", "1_1_2", "1_2_2", "1_3_2", "1_1_3", "1_2_3", "1_3_3", "1_1_4", "1_2_4", "1_3_4"))
}

sign_permute_replicates <- 200

sign_permute_blocks <- create_blocks(gen0, X1, X2, 15000, "_")

pdiffs$window = sign_permute_blocks$window

rm(sign_permute_blocks)
gc()

# make list of sites
sites <- unique(gsub("_.*", "", colnames(pdiffs)))
sites <- sites[c(-31)]

#
output_gt_perm <- paste(results_path, "gt_abs_permutations.txt", sep = "")

# create permutations of allele frequency changes by site
mean_gt_perms = NULL
# subset to one site
for (site in sites) {
  print(paste("Site:", site))
  pdiffs_site <- pdiffs %>%
    dplyr::select(starts_with(paste(site, "_", sep = "")), window)
  # now permute many times
  for (p in 1:sign_permute_replicates) {
    # subset to one rep
    print(paste("Replicate:", p))
    # choose a random replicate
    # within a block, randomly multiply allele frequency change by +1 or -1
    pdiffs_perm <- pdiffs_site %>%
      group_by(window) %>%
      mutate(across(where(is.numeric), ~ .x * sample(c(1, -1), size = 1))) %>%
      ungroup() %>%
      dplyr::select(-window)
    
    gt_perms = NULL
    # loop over each replicate
    for (exprep in 1:12) {
      pdiffs_perm_rep <- pdiffs_perm %>%
        dplyr::select(ends_with(paste("_", exprep, sep = "")))
      
      if (ncol(pdiffs_perm_rep) <= 1) {
        #print("Need at least two increments. Skipping to next replicate...")
        next
      }
      
      # un-corrected covariance matrices
      cov_pdiffs_perm <- cov(pdiffs_perm_rep)
      
      # TO-DO: correct covariance matrices
      
      # calculate covariance rolling sums
      gt_perm <- gt_from_covmat(cov_pdiffs_perm)
      
      # save result
      gt_perm$rep <- exprep
      gt_perms <- rbind(gt_perms, gt_perm)
    }
    
    # average across replicates to get one permutation sample
    if(is.null(gt_perms)){
      next
    }
    
    mean_gt_perm = gt_perms %>%
      group_by(gen) %>%
      summarize(
        gt = mean(sum_cov) / mean(sum_cov + sum_var),
        gt_abs = mean(sum_abs_cov) / mean(sum_abs_cov + sum_var),
        sum_cov = mean(sum_cov),
        sum_abs_cov = mean(sum_abs_cov),
        sum_pos_cov = mean(sum_pos_cov),
        sum_neg_cov = mean(sum_neg_cov),
        sum_var = mean(sum_var),
        ratio = mean(ratio, na.rm = T),
        n = n()
      )
    
    mean_gt_perm$perm = p
    mean_gt_perm$site = site
    
    write.table(mean_gt_perm, output_gt_perm, col.names=!file.exists(output_gt_perm), append = T, row.names = F, sep = ",", quote = F)
  }
}

```

### calculate permutation null distributions

```{r}

mean_gt_perms <- fread(paste(results_path, "gt_abs_permutations.txt", sep = ""), header = T)

# remove extra permutations
# sometimes happened, if R needed to be restarted from checkpoint
mean_gt_perms <- mean_gt_perms %>% 
  distinct(site, gen, perm, .keep_all = TRUE)

# calculate percentiles of permutation distribution
perm_int <- mean_gt_perms %>% 
  group_by(gen, site) %>% 
  summarize(lwr = quantile(gt_abs, 0.025), med = median(gt_abs), upr = quantile(gt_abs, 0.975)) %>%
  arrange(site)

write.table(perm_int, paste(results_path, "gt_abs_permutation_intervals_200.txt", sep = ""), row.names = F, quote = F, sep = "\t")
```


# Convergence correlations

## between environments, first generation

```{r}

# loop over pairs of sites
sites <- unique(gsub("_.*", "", colnames(hapfire)))
sites <- sites[c(-30,-32)]

conv_cor_blocks <- create_blocks(gen0, X1, X2, 15000, "_")

hapfire$window = conv_cor_blocks$window

output_conv_cor <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/", today, "/convergence_correlation_first_increment_by_block.tsv", sep = "")

for(w in unique(hapfire$window)){
  print(paste("Window:", w))
  hap_block <- hapfire %>% filter(window == w)
  gen0_block <- conv_cor_blocks %>% filter(window == w)
  for (i in seq_along(sites)) {
    a <- sites[i]
    hap_site_a <- hap_block %>% 
      dplyr::select(starts_with(paste(a, "_1_", sep = "")))
    print(paste("Site:", a))
    for (j in seq_len(length(sites))[-seq_len(i)]) {
      b <- sites[j]
      hap_site_b <- hap_block %>% 
        dplyr::select(starts_with(paste(b, "_1_", sep = "")))

      # calculate covariances across sites
      conv_cor_result <- covmat_pop_pair(hap_site_a, 
                      hap_site_b, 
                      paste("^", a, "_", sep = ""), 
                      paste("^", b, "_", sep = ""), 
                      gen0_block$X3)
      
      conv_cor_row <- data.frame(
        window = w,
        site_a = a,
        site_b = b,
        conv_cor = conv_cor_result
      )
      
      append_table(conv_cor_row, output_conv_cor)
    }
  }
}

```

### bootstrap intervals

```{r}

conv_cor_results <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-05-08/convergence_correlation_first_increment_by_block.tsv")

W = length(unique(conv_cor_results$window))
B = 1000
alpha = 0.05

boot_samples <- NULL

for(b in 1:B){
  
boot_sample <- conv_cor_results %>% 
  group_by(site_a, site_b) %>%
  slice_sample(n = W, replace = TRUE) %>% 
  summarize(
      conv_cor = mean(conv_cor)
    )
  
  boot_samples <- rbind(boot_samples, boot_sample)
}

conv_cor_cis <- boot_samples %>% 
  group_by(site_a, site_b) %>% 
  summarize(lwr = quantile(conv_cor, alpha/2), 
            med = median(conv_cor),
            upr = quantile(conv_cor, 1-alpha/2))

write.table(conv_cor_cis, paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/", today, "/convergence_correlation_bootstrap_intervals.txt", sep = ""), sep = "\t", quote = F, row.names = F)
```

## within environment, first generation

This should be nearly identical to the repeatability analysis from the Phase 1 science paper, except here I do a ratio of averages instead of an average of ratios.

```{r}
if(!exists("pdiffs")){
  pdiffs <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/allele_frequency_increments.csv", header = T)
}

# loop over pairs of sites
sites <- unique(gsub("_.*", "", colnames(pdiffs)))

conv_cor_blocks <- create_blocks(gen0, X1, X2, 15000, "_")

pdiffs$window = conv_cor_blocks$window

output_conv_cor <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/", today, "/convergence_correlation_first_increment_by_block_wn_env.tsv", sep = "")

for (i in seq_along(sites)) {
  a <- sites[i]
  pdiffs_site <- pdiffs %>% 
    dplyr::select(starts_with(paste(a, "_1_", sep = "")), window)
  
  print(paste("Site:", a))
  
  if(ncol(pdiffs_site) <= 2){
    print("Too few increments. Skipping to next site.")
    next
  }
  
  for(w in unique(pdiffs$window)){
    #print(paste("Window:", w))
    pdiffs_block <- pdiffs_site %>% filter(window == w) %>% dplyr::select(-window)
    gen0_block <- conv_cor_blocks %>% filter(window == w)
    
    conv_cor <- conv_cor_wn_env(pdiffs_block)
    
    conv_cor_row <- data.frame(
      window = w,
      site = a,
      S = nrow(pdiffs_block),
      conv_cor_numer = conv_cor[1],
      conv_cor_denom = conv_cor[2]
    )
    
    append_table(conv_cor_row, output_conv_cor)
  }
}


```

### bootstrap intervals

```{r}

conv_cor_results <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-05-28/convergence_correlation_first_increment_by_block_wn_env.tsv")

W = length(unique(conv_cor_results$window))
B = 1000
alpha = 0.05

boot_samples <- NULL

for(b in 1:B){
  
boot_sample <- conv_cor_results %>% 
  group_by(site) %>%
  slice_sample(n = W, replace = TRUE) %>% 
  summarize(
      conv_cor = weighted.mean(conv_cor_numer, S)/weighted.mean(conv_cor_denom, S)
    )
  
  boot_samples <- rbind(boot_samples, boot_sample)
}

conv_cor_cis <- boot_samples %>% 
  group_by(site) %>% 
  summarize(lwr = quantile(conv_cor, alpha/2), 
            med = median(conv_cor),
            upr = quantile(conv_cor, 1-alpha/2))

write.table(conv_cor_cis, paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/", today, "/convergence_correlation_wn_env_bootstrap_intervals.txt", sep = ""), sep = "\t", quote = F, row.names = F)
```

# Replicate G(t)

## calculate numerator and denominator in windows

```{r}

if(!exists("pdiffs")){
  pdiffs <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/allele_frequency_increments.csv", header = T)
}


# output file name
output_rep_gt <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/", today, "/replicate_gt_by_block_wn_env.tsv", sep = "")

# loop over pairs of sites
sites <- unique(gsub("_.*", "", colnames(pdiffs)))
rep_gt_blocks <- create_blocks(gen0, X1, X2, 15000, "_")
pdiffs$window = rep_gt_blocks$window

for (site in sites) {
    
  print(paste("Site:", site))
  
    pdiffs_site <- pdiffs %>% 
      dplyr::select(starts_with(paste(site, "_", sep = "")), window)
    
    rep_labels <- gsub(".*_", "", colnames(pdiffs_site))
    rep_labels <- rep_labels[-length(rep_labels)]
    
    time_labels <- gsub("_", "", str_extract(colnames(pdiffs_site), "_.*_"))
    time_labels <- time_labels[-length(time_labels)]
 
    if( (length(unique(time_labels)) <= 2) | (length(unique(rep_labels)) <= 2)){
      print("Too few time points or replicates. Skipping to next site...")
      next
    }
    
    for(w in unique(pdiffs$window)){
      #print(paste("Window:", w))
      pdiffs_block <- pdiffs_site %>% filter(window == w) %>% dplyr::select(-window)
      #pdiffs_site <- pdiffs_site %>% dplyr::select(-window)
      
      rep_gt <- replicate_gt(pdiffs_block, rep_labels, time_labels)
      #rep_gt <- replicate_gt(pdiffs_site, rep_labels, time_labels)
      rep_gt_row <- cbind(
        data.frame(
          window = w,
          site = site,
          S = nrow(pdiffs_block)
        ), 
        rep_gt)
      
      append_table(rep_gt_row, output_rep_gt)
    }
}

```

## bootstrap intervals

```{r}
rep_gt_results <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-06-01/replicate_gt_by_block_wn_env.tsv")

W = length(unique(rep_gt_results$window))
B = 1000
alpha = 0.05

boot_samples <- NULL

for(b in 1:B){
  
boot_sample <- rep_gt_results %>% 
  group_by(site) %>%
  slice_sample(n = W, replace = TRUE) %>% 
  summarize(
      rep_gt = weighted.mean(mean_covar, S)/weighted.mean(mean_var, S),
      rep_gt_pos = weighted.mean(mean_covar_pos, S)/weighted.mean(mean_var_pos, S),
      rep_gt_abs = weighted.mean(mean_covar_abs, S)/weighted.mean(mean_var_abs, S),
    )
  
  boot_samples <- rbind(boot_samples, boot_sample)
}

rep_gt_cis <- boot_samples %>% 
  group_by(site) %>% 
  summarize(lwr = quantile(rep_gt, alpha/2), 
            med = median(rep_gt),
            upr = quantile(rep_gt, 1-alpha/2))

rep_gt_abs_cis <- boot_samples %>% 
  group_by(site) %>% 
  summarize(lwr = quantile(rep_gt_abs, alpha/2), 
            med = median(rep_gt_abs),
            upr = quantile(rep_gt_abs, 1-alpha/2))

rep_gt_pos_cis <- boot_samples %>% 
  group_by(site) %>% 
  summarize(lwr = quantile(rep_gt_pos, alpha/2), 
            med = median(rep_gt_pos),
            upr = quantile(rep_gt_pos, 1-alpha/2))

write.table(rep_gt_abs_cis, paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/", today, "/replicate_gt_by_block_wn_env_bootstrap_intervals.txt", sep = ""), sep = "\t", quote = F, row.names = F)
```

## sign-permuted null distribution

```{r}

if(!exists("pdiffs")){
  pdiffs <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/allele_frequency_increments.csv", header = T)
}

# Number of sign permutations
perms <- 50

# loop over pairs of sites
sites <- unique(gsub("_.*", "", colnames(pdiffs)))
rep_gt_blocks <- create_blocks(gen0, X1, X2, 15000, "_")
pdiffs$window = rep_gt_blocks$window

output_rep_gt_perm <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/", today, "/replicate_gt_by_block_wn_env_permutations.tsv", sep = "")

for (site in sites) {
  
  print(paste("Site:", site))
  
  pdiffs_site <- pdiffs %>% 
    dplyr::select(starts_with(paste(site, "_", sep = "")), window)
  
  rep_labels <- gsub(".*_", "", colnames(pdiffs_site))
  rep_labels <- rep_labels[-length(rep_labels)]
  
  time_labels <- gsub("_", "", str_extract(colnames(pdiffs_site), "_.*_"))
  time_labels <- time_labels[-length(time_labels)]
  
  if( (length(unique(time_labels)) <= 2) | (length(unique(rep_labels)) <= 2)){
    print("Too few time points or replicates. Skipping to next site...")
    next
  }
  
  for(w in unique(pdiffs$window)){
    pdiffs_block <- pdiffs_site %>% filter(window == w) %>% dplyr::select(-window)
    
    for(perm in 1:perms){
      
      # draw whether an increment will be flipped or not 
      sign_permutations <- sample(c(-1,1), size = ncol(pdiffs_block), replace = T)
      
      # Create the diagonal matrix from vector
      P <- diag(sign_permutations)
      
      # Perform matrix multiplication to flip all increments at once
      pdiffs_block_perm <- as.matrix(pdiffs_block) %*% P
      
      # calculate replicate G(t)
      rep_gt <- replicate_gt(pdiffs_block_perm, rep_labels, time_labels)
      
      # save results
      rep_gt_row <- cbind(
        data.frame(
          window = w,
          site = site,
          perm = perm,
          S = nrow(pdiffs_block)
        ),
        rep_gt)
      
      append_table(rep_gt_row, output_rep_gt_perm)
    }
  }
}

rep_gt_perm <- read.table("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-06-13/replicate_gt_by_block_wn_env_permutations.tsv", sep = ",", header = T)

rep_gt_perm %>% 
  group_by(site) %>%
  summarize(
      rep_gt = weighted.mean(mean_covar, S)/weighted.mean(mean_var, S),
      rep_gt_pos = weighted.mean(mean_covar_pos, S)/weighted.mean(mean_var_pos, S),
      rep_gt_abs = weighted.mean(mean_covar_abs, S)/weighted.mean(mean_var_abs, S),
    )

```

# Recombination rate vs pi

## load, format, and filter recombination map

```{r}
#
data("Arabidopsis_thaliana")

recom_map_1 <- data.frame(
  chrom = 1,
  pos = Arabidopsis_thaliana@maps$`Chromosome 1`@physicalPositions,
  cm = Arabidopsis_thaliana@maps$`Chromosome 1`@geneticDistances
)

recom_map_1 <- recom_map_1 %>% filter(cm == cummax(cm))

recom_map_2 <- data.frame(
  chrom = 2,
  pos = Arabidopsis_thaliana@maps$`Chromosome 2`@physicalPositions,
  cm = Arabidopsis_thaliana@maps$`Chromosome 2`@geneticDistances
)

recom_map_2$cm <- recom_map_2$cm - min(recom_map_2$cm)

recom_map_2 <- recom_map_2 %>% filter(cm == cummax(cm))

recom_map_3 <- data.frame(
  chrom = 3,
  pos = Arabidopsis_thaliana@maps$`Chromosome 3`@physicalPositions,
  cm = Arabidopsis_thaliana@maps$`Chromosome 3`@geneticDistances
)

recom_map_3$cm <- recom_map_3$cm - min(recom_map_3$cm)

recom_map_3 <- recom_map_3 %>% filter(cm == cummax(cm))

recom_map_4 <- data.frame(
  chrom = 4,
  pos = Arabidopsis_thaliana@maps$`Chromosome 4`@physicalPositions,
  cm = Arabidopsis_thaliana@maps$`Chromosome 4`@geneticDistances
)

recom_map_4$cm <- recom_map_4$cm - min(recom_map_4$cm)

recom_map_4 <- recom_map_4 %>% filter(cm == cummax(cm))

# 5
recom_map_5 <- data.frame(
  chrom = 5,
  pos = Arabidopsis_thaliana@maps$`Chromosome 5`@physicalPositions,
  cm = Arabidopsis_thaliana@maps$`Chromosome 5`@geneticDistances
)

recom_map_5$cm <- recom_map_5$cm - min(recom_map_5$cm)

recom_map_5 <- recom_map_5 %>% filter(cm == cummax(cm))

# combine all chr
# convert pos to Mb
recom_map <- rbind(recom_map_1, recom_map_2, recom_map_3, recom_map_4, recom_map_5)

recom_map$pos <- recom_map$pos/1e6

ggplot(recom_map, aes(x = pos, y = cm)) +
  geom_point() +
  facet_wrap(chrom~., scales = "free") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Position (Mb)", y = "cM")

ggsave(paste(results_path, "marey_map.pdf", sep = ""), width = 7, height = 6)

```

## LD decay
```{r}

# LD decay threshold
threshold <- 0.1
  
# define blocks within which to calculate LD decay
# helps contextualize blocks for other analyses
ld_decay_blocks <- create_blocks(gen0, X1, X2, 15000, "_")

output_ld_decay <- paste0("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/", today, "/ld_decay_by_block.txt", sep = "")

check_file_exists(output_ld_decay)

# Loop over each block and calculate LD decay within the block
for(w in unique(ld_decay_blocks$window)) {
  cat(w, "\n")
  
  block_coords <- ld_decay_blocks %>%
    filter(window == w) %>%
    summarize(chrom = unique(X1),
              start = min(X2),
              end = max(X2))
  
  chrom = block_coords$chrom
  start = block_coords$start
  end   = block_coords$end
  
  # extract test region
  ld_region <- ld_matrix %>%
    filter(
      (CHR_A == chrom) &
        (BP_A >= start) &
        (BP_A <= end) &
        (CHR_B == chrom) &
        (BP_B >= start) &
        (BP_B <= end)
    ) %>%
    mutate(d = BP_B - BP_A, n = 231) %>%
    collect()
  
  # Fit Hill and Weir Non-Linear Least Squares (NLS) Model
  hw_model <- nls(
    R2 ~ hill_weir_r2(d, n, C),
    data = ld_region,
    start = list(C = 0.001),
    control = nls.control(maxiter = 1000, warnOnly = TRUE)
  )
  
  # Extract the fitted recombination parameter (C)
  fitted_C <- coef(hw_model)["C"]
  line1 <- paste("Estimated 4Nc:", signif(fitted_C,3), "\n")
  
  # Generate Predicted Trend Line Values
  pred_distances <- seq(0, max(ld_region$d), length.out = 1000)
  pred_r2 <- predict(hw_model, newdata = data.frame(d = pred_distances, n = 231))
  plot_data <- data.frame(d = pred_distances, R2 = pred_r2)
  
  # Calculate LD Decay Distance at r^2 threshold (e.g., r2 = 0.2)
  decay_distance <- pred_distances[which.min(abs(pred_r2 - threshold))]
  line2 <- paste("LD decays to r^2 =",
      threshold,
      "at:",
      round(decay_distance / 1000, 2),
      "kb\n")
  
  line3 <- paste("Block coordinates: ", chrom, ":", start, "..", end, "\n", sep = "")
  
  # Plot the Raw Data Bins and the Fitted Hill & Weir Curve
  ggplot(ld_region, aes(x = d, y = R2)) +
    geom_point(alpha = 0.3, color = "darkgray", shape = 16) +
    geom_line(
      data = plot_data,
      aes(x = d, y = R2),
      color = "skyblue",
      size = 1.2
    ) +
    geom_hline(yintercept = threshold,
               linetype = "dashed",
               color = "pink") +
    geom_vline(xintercept = decay_distance,
               linetype = "dashed",
               color = "pink") +
    labs(title = paste(line1,line2,line3, sep = ""), x = "Physical Distance (bp)", y = "LD (r^2)")
  
    ggsave(paste(results_path, "ld_decay_", chrom, "_", start, "_", end, ".png", sep = ""), width = 5.5, height = 6)
    gc()
    
    # save summary to table
    ld_decay_result <- data.frame(
      chrom = chrom,
      start = start,
      end = end,
      block = w,
      C = fitted_C,
      decay_threshold = threshold,
      decay_distance = decay_distance,
               )
    append_table(ld_decay_result, output_ld_decay)
}

```

## compile diversity values

```{r}
# now loop over diversity files
pi_files <- list.files(path = "../data/diversity", recursive = T, full.names = T, pattern = "_pi.txt$")

file_prefixes <- gsub("/", "_", gsub("../data/diversity/|.txt", "", pi_files))

plot_list <- list()

for(i in seq_along(pi_files)){
  
  pi_file <- pi_files[i]
  
  cat(paste0("Processing ", pi_file, "...\n"))
  
  nuc_div <- fread(pi_file, header = T)

  # calculate window midpoint
  nuc_div$window_pos_mid <- (nuc_div$window_pos_2 + nuc_div$window_pos_1 + 1)/2
  nuc_div$window_pos_mid <- nuc_div$window_pos_mid/1e6 #per mb
  
  # match chromosome name across files
  names(nuc_div)[2] <- "chrom"
  
  # extract number of individuals in vcf
  num_indiv <- as.numeric(gsub("_pi.txt", "", str_split(pi_file, "/")[[1]][5]))
  site_class <- str_split(pi_file, "/")[[1]][4]
  # if diversity file excludes invariant sites, rescale denominator
  # count_comparisons = choose(<number of diploid individuals>*2,2)*no_sites
  if( (num_indiv == 231 | num_indiv == 2029) & (site_class == "all_sites") ){
    cat("Rescaling pi denominator to assume all missing sites are invariant...\n")
    nuc_div$count_comparisons <- choose(num_indiv*2,2)*(nuc_div$window_pos_2 - nuc_div$window_pos_1 + 1)
    nuc_div$avg_pi <- nuc_div$count_diffs/nuc_div$count_comparisons
  }
  
  if( (num_indiv == 231 | num_indiv == 2029) & (site_class != "all_sites") ){
    if(site_class == "four_fold"){
      num_4_per_win <- fread("~/moi_lab_projects/grenenet-phase2/data/diversity/four_fold/four_fold_per_window.bed", col.names = c("chrom", "window_pos_1", "window_pos_2", "num_4"), colClasses = list(integer = 1:4))
      num_4_per_win$V2 <- num_4_per_win$V2 + 1
      nuc_div %>%
        left_join(num_4_per_win, by = c("chrom", "window_pos_1", "window_pos_2"))
    }
    if(site_class == "zero_fold"){
      
    }
  }

```

## Diversity vs recombination rate

```{r}
# fit model to marey map
best_mods <- recom_map %>%
  group_by(chrom) %>%
  nest() %>%
  mutate(model = purrr::map(data, ~ find_best_poly(.x, 
                                                   x_col = "pos",
                                                   y_col = "cm", 
                                                   max_degree = 7)))
# Plot model fits
grid_data <- best_mods %>%
  mutate(
    grid = purrr::map(data, ~ tibble(x_val = seq(min(.x$pos), max(.x$pos), length.out = 500))))

pred_data <- grid_data %>% 
  mutate(fit  = purrr::map2(model, grid, ~ predict(.x, newdata = .y))
  ) %>%
  select(chrom, grid, fit) %>%
  unnest(c(grid, fit))

names(pred_data)[2] = "pos"

ggplot(pred_data, aes(x = pos, y = fit)) +
  geom_point(data = recom_map, mapping = aes(x = pos, y = cm), shape = 16) +
  geom_line(color = "red") +
  facet_wrap(.~chrom, scales = "free", labeller = as_labeller(function(value) paste0("Chromosome ", value))) +
  labs(x = "Physical position (Mb)", y = "Genetic position (cM)")

ggsave(paste(results_path, "marey_map_fitted.pdf", sep = ""), width = 7, height = 6)

# now loop over diversity files
pi_files <- list.files(path = "../data/diversity", recursive = T, full.names = T, pattern = "_pi.txt$")

file_prefixes <- gsub("/", "_", gsub("../data/diversity/|.txt", "", pi_files))

plot_list <- list()
for(i in seq_along(pi_files)){
  
  cat(paste0("Processing ", pi_files[i], "...\n"))
  
  nuc_div <- fread(pi_files[i], header = T)

  
  # calculate window midpoint
  nuc_div$window_pos_mid <- (nuc_div$window_pos_2 + nuc_div$window_pos_1 + 1)/2
  nuc_div$window_pos_mid <- nuc_div$window_pos_mid/1e6 #per mb
  
  # match chromosome name across files
  names(nuc_div)[2] <- "chrom"
  
  # extract number of individuals in vcf
  num_indiv <- as.numeric(str_split(title, " ")[[1]][3])
  
  # if diversity file excludes invariant sites, rescale denominator
  # count_comparisons = choose(<number of diploid individuals>*2,2)*no_sites
  if(num_indiv == 231 | num_indiv == 2029){
    cat("Rescaling pi denominator to assume all missing sites are invariant...\n")
    nuc_div$count_comparisons <- choose(num_indiv*2,2)*(nuc_div$window_pos_2 - nuc_div$window_pos_1 + 1)
    nuc_div$avg_pi <- nuc_div$count_diffs/nuc_div$count_comparisons
  }

  # Evaluate derivative of model for windows
  rate_vs_div <- nuc_div %>%
    left_join(best_mods, by = "chrom") %>%
    group_by(chrom) %>%
    mutate(rate = eval_deriv(model[[1]], window_pos_mid)) %>%
    select(chrom, window_pos_mid, rate, avg_pi) %>%
    unnest(cols = c(rate))
  
  # filter out negative recombination rates
  rate_vs_div <- rate_vs_div %>% filter(rate > 0)
  
  # recombination rate across chromosome
  ggplot(rate_vs_div, aes(x = window_pos_mid, y = rate)) +
    geom_line() +
    facet_wrap(.~chrom) +
    labs(x = "Position (bp)", y = "Recombination rate (cM/Mb)")
  
  ggsave(paste(results_path, file_prefixes[i],"_recombination_rate.pdf", sep = ""))
  
  # correlation between recombination rate and diversity
  corr_results <- corr_ci_autocorr(
    rate_vs_div$rate[rate_vs_div$rate <= 10],
    rate_vs_div$avg_pi[rate_vs_div$rate <= 10]
    )
  
  # build plot title
    title <- gsub("^ ", "", gsub("/|_", " ", gsub("../data/diversity|_pi.txt", "", pi_files[i])))
 
   title <- paste0(title,
                  "\nr = ", 
                  signif(corr_results$r, 2), 
                  ", p = ", 
                  format.pval(corr_results$p_value, digits = 2, eps = 0.001), 
                  "\n95 % CI = [", 
                  signif(corr_results$ci_lower, 2), 
                  ", ", 
                  signif(corr_results$ci_upper, 2), 
                  "]")
  
  # diversity vs recombination rate
  plot_rate_div <- ggplot(rate_vs_div, aes(x = rate, y = avg_pi)) +
    geom_point() +
    #geom_smooth(method = "lm", color = "lightblue") + 
    #stat_cor() +
    #xlim(0, 10) +
    geom_smooth(
      data = subset(rate_vs_div, rate <= 10), 
      method = "lm", 
      se = TRUE, 
      color = "lightblue"
    ) +
    facet_zoom(rate <= 10) +
    theme(
      # Draws a border line and shaded region for the connecting zoom lines
      zoom = element_rect(colour = "pink", linewidth = 0.8, fill = alpha("pink", 0.01))) +
    labs(x = "Recombination rate (cM/Mb)", 
         y = "Average nucleotide diversity",
         title = title)
  
  plot_list[[i]] <- plot_rate_div
  
  ggsave(paste(results_path, file_prefixes[i],"_recombination_vs_diversity.pdf", sep = ""), plot_rate_div, width = 5.5, height = 8)
}

# adjust plots to fit them all together
plot_list <- purrr::map(plot_list, ~ .x + theme(text = element_text(size = 12)))

ggarrange(plot_list[[1]], plot_list[[2]], plot_list[[3]], nrow = 1, ncol = 3)
ggsave(paste0(results_path, "/all_sites_recombionation_rate_vs_diversity.pdf"), width = 9, height = 5)

ggarrange(plot_list[[4]], plot_list[[5]], plot_list[[6]], nrow = 1, ncol = 3)
ggsave(paste0(results_path, "/four_fold_recombionation_rate_vs_diversity.pdf"), width = 9, height = 5)

ggarrange(plot_list[[7]], plot_list[[8]], plot_list[[9]], nrow = 1, ncol = 3)
ggsave(paste0(results_path, "/zero_fold_recombionation_rate_vs_diversity.pdf"), width = 9, height = 5)
```

## gene density vs diversity

```{r}
gene_density <- fread("/global/scratch/users/milesroberts/genomes/arabidopsis_thaliana/cds_per_window.bed")

names(gene_density) <- c("chrom", "window_pos_1", "window_pos_2", "V4", "V5", "V6", "frac_cds")

gene_density$chrom <- as.integer(gene_density$chrom)
gene_density$window_pos_1 <- gene_density$window_pos_1 + 1

gene_density_vs_div_plot_list <- list()

for(i in seq_along(pi_files)){
  
  # load data
  cat(paste0("Processing ", pi_files[i], "...\n"))
  nuc_div <- fread(pi_files[i], header = T)
  names(nuc_div)[2:4] <- c("chrom", "window_pos_1", "window_pos_2")
  
  # if windows are left shifted, correct them
  nuc_div <- nuc_div %>%
    group_by(chrom) %>%
    mutate(window_pos_2 = window_pos_2 - min(window_pos_1) + 1,
           window_pos_1 = window_pos_1 - min(window_pos_1) + 1)
  
  # combine density and diversity
  nuc_div <- nuc_div %>%
    left_join(gene_density, by = c("chrom", "window_pos_1", "window_pos_2"))
  
  # plot title
  title <- gsub("^ ", "", gsub("/|_", " ", gsub("../data/diversity|_pi.txt", "", pi_files[i])))
  
  # extract number of individuals in vcf
  num_indiv <- as.numeric(str_split(title, " ")[[1]][3])
  
  # calculate window midpoint
  nuc_div$window_pos_mid <- (nuc_div$window_pos_2 + nuc_div$window_pos_1 + 1)/2
  nuc_div$window_pos_mid <- nuc_div$window_pos_mid/1e6 #per mb
  
  # if diversity file excludes invariant sites, rescale denominator
  # count_comparisons = choose(<number of diploid individuals>*2,2)*no_sites
  if(num_indiv == 231 | num_indiv == 2029){
    cat("Rescaling pi denominator to assume all missing sites are invariant...\n")
    nuc_div$count_comparisons <- choose(num_indiv*2,2)*(nuc_div$window_pos_2 - nuc_div$window_pos_1 + 1)
    nuc_div$avg_pi <- nuc_div$count_diffs/nuc_div$count_comparisons
  }
  
  # Evaluate derivative of model for windows
  rate_vs_div <- nuc_div %>%
    left_join(best_mods, by = "chrom") %>%
    group_by(chrom) %>%
    mutate(rate = eval_deriv(model[[1]], window_pos_mid)) %>%
    select(chrom, window_pos_mid, rate, avg_pi, frac_cds) %>%
    unnest(cols = c(rate))
  
  # filter out negative recombination rates
  rate_vs_div <- rate_vs_div %>% filter(rate > 0)
 
  # correlation between gene_density and diversity
  corr_results <- corr_ci_autocorr(
    rate_vs_div$frac_cds,
    rate_vs_div$avg_pi
    )
 
   title <- paste0(title,
                  "\nr = ", 
                  signif(corr_results$r, 2), 
                  ", p = ", 
                  format.pval(corr_results$p_value, digits = 2, eps = 0.001), 
                  "\n95 % CI = [", 
                  signif(corr_results$ci_lower, 2), 
                  ", ", 
                  signif(corr_results$ci_upper, 2), 
                  "]")
   
  # plot
  plot_density_vs_div <- ggplot(rate_vs_div, aes(x = frac_cds, y = avg_pi)) +
    geom_point() +
    geom_smooth(
      method = "lm", 
      se = TRUE, 
      color = "lightblue"
    ) +
    labs(y = "Average nucleotide diversity (\u03c0)", 
         x = "Fraction of window\nannotated CDS",
         title = title) 
  
  gene_density_vs_div_plot_list[[i]] <- plot_density_vs_div
  
    ggsave(paste(results_path, file_prefixes[i],"_gene_density_vs_diversity.pdf", sep = ""), plot_density_vs_div, width = 5.5, height = 8, device = cairo_pdf)
    
}

# adjust plots to fit them all together
gene_density_vs_div_plot_list <- purrr::map(gene_density_vs_div_plot_list, ~ .x + theme(text = element_text(size = 12)))

ggarrange(gene_density_vs_div_plot_list[[1]], 
          gene_density_vs_div_plot_list[[2]],
          gene_density_vs_div_plot_list[[3]], 
          nrow = 1, 
          ncol = 3)
ggsave(paste0(results_path, "/all_sites_gene_density_vs_diversity.pdf"), width = 9, height = 3.5, device = cairo_pdf)

ggarrange(gene_density_vs_div_plot_list[[4]],
          gene_density_vs_div_plot_list[[5]],
          gene_density_vs_div_plot_list[[6]],
          nrow = 1,
          ncol = 3)
ggsave(paste0(results_path, "/four_fold_gene_density_vs_diversity.pdf"), width = 9, height = 3.5, device = cairo_pdf)

ggarrange(gene_density_vs_div_plot_list[[7]],
          gene_density_vs_div_plot_list[[8]],
          gene_density_vs_div_plot_list[[9]],
          nrow = 1,
          ncol = 3)
ggsave(paste0(results_path, "/zero_fold_gene_density_vs_diversity.pdf"), width = 9, height = 3.5, device = cairo_pdf)
  
```

## adaptest recombination map

```{r}
adaptest<-fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/Adaptest_genetic_map_500kb.csv")

#
names(adaptest)[1] <- "chrom"
adaptest$window_pos_1 <- adaptest$bin + 1
adaptest$window_pos_2 <- adaptest$bin + 5e5

# now loop over diversity files
pi_files <- list.files(path = "../data/diversity", recursive = T, full.names = T, pattern = "_pi.txt$")

file_prefixes <- gsub("/", "_", gsub("../data/diversity/|.txt", "", pi_files))

# loop over pixy outputs
plots_adaptest <- list()
for(i in seq_along(pi_files)){
  
  # load data
  nuc_div <- fread(pi_files[i])
  names(nuc_div)[2] <- "chrom"
  
  # if windows are left shifted, correct them
  nuc_div <- nuc_div %>%
    group_by(chrom) %>%
    mutate(window_pos_2 = window_pos_2 - min(window_pos_1) + 1,
           window_pos_1 = window_pos_1 - min(window_pos_1) + 1)
  
  # extract number of individuals in vcf
  num_indiv <- as.numeric(gsub("sites/|fold/", "", str_split(pi_files[i], "_")[[1]][2]))
  
  # if diversity file excludes invariant sites, rescale denominator
  # count_comparisons = choose(<number of diploid individuals>*2,2)*no_sites
  if(num_indiv == 231 | num_indiv == 2029){
    cat("Rescaling pi denominator to assume all missing sites are invariant...\n")
    nuc_div$count_comparisons <- choose(num_indiv*2,2)*(nuc_div$window_pos_2 - nuc_div$window_pos_1 + 1)
    nuc_div$avg_pi <- nuc_div$count_diffs/nuc_div$count_comparisons
  }
  
  rate_vs_div <- nuc_div %>%
    left_join(adaptest, by = c("chrom", "window_pos_1", "window_pos_2"))
  
  # correlation between recombination rate and diversity
  corr_results <- corr_ci_autocorr(
    rate_vs_div$rate_cM_Mb[rate_vs_div$rate_cM_Mb <= 10],
    rate_vs_div$avg_pi[rate_vs_div$rate_cM_Mb <= 10]
  )
  
  # build plot title
  title <- gsub("^ ", "", gsub("/|_", " ", gsub("../data/diversity|_pi.txt", "", pi_files[i])))
  
  title <- paste0(title,
                  "\nr = ", 
                  signif(corr_results$r, 2), 
                  ", p = ", 
                  format.pval(corr_results$p_value, digits = 2, eps = 0.001), 
                  "\n95 % CI = [", 
                  signif(corr_results$ci_lower, 2), 
                  ", ", 
                  signif(corr_results$ci_upper, 2), 
                  "]")
  
  # diversity vs recombination rate
  plot_adaptest <- ggplot(rate_vs_div, aes(x = rate_cM_Mb, y = avg_pi)) +
    geom_point() +
    geom_smooth(
      data = subset(rate_vs_div, rate_cM_Mb <= 10), 
      method = "lm", 
      se = TRUE, 
      color = "lightblue"
    ) +
    facet_zoom(rate_cM_Mb <= 10) +
    theme(
      # Draws a border line and shaded region for the connecting zoom lines
      zoom = element_rect(colour = "pink", linewidth = 0.8, fill = alpha("pink", 0.01))) +
    labs(x = "Recombination rate (cM/Mb)", 
         y = "Average nucleotide diversity",
         title = title)
  
  plots_adaptest[[i]] <- plot_adaptest
  ggsave(paste(results_path, "adaptest_", file_prefixes[i],"_recombination_vs_diversity.pdf", sep = ""), plot_adaptest, width = 5, height = 8)
  
}

# adjust plots to fit them all together
plots_adaptest <- purrr::map(plots_adaptest, ~ .x + theme(text = element_text(size = 12)))

# trios of data
ggarrange(plots_adaptest[[1]], 
          plots_adaptest[[2]],
          plots_adaptest[[3]], 
          nrow = 1, 
          ncol = 3)
ggsave(paste0(results_path, "/adaptest_all_sites_gene_density_vs_diversity.pdf"),
       width = 9,
       height = 4.5, device = cairo_pdf)

ggarrange(plots_adaptest[[4]],
          plots_adaptest[[5]],
          plots_adaptest[[6]],
          nrow = 1,
          ncol = 3)
ggsave(paste0(results_path, "/adaptest_four_fold_gene_density_vs_diversity.pdf"),
       width = 9,
       height = 4.5, device = cairo_pdf)

ggarrange(plots_adaptest[[7]],
          plots_adaptest[[8]],
          plots_adaptest[[9]],
          nrow = 1,
          ncol = 3)
ggsave(paste0(results_path, "/adaptest_zero_fold_gene_density_vs_diversity.pdf"), 
       width = 9,
       height = 4.5, device = cairo_pdf)
```

## gene density vs diversity

```{r}

# compile diversity data, for crying out loud
nuc_div_all <- NULL
for(i in seq_along(pi_files)){
  
  # load data
  nuc_div <- fread(pi_files[i])
  names(nuc_div)[2] <- "chrom"
  
  # if windows are left shifted, correct them
  nuc_div <- nuc_div %>%
    group_by(chrom) %>%
    mutate(window_pos_2 = window_pos_2 - min(window_pos_1) + 1,
           window_pos_1 = window_pos_1 - min(window_pos_1) + 1,
           window_pos_mid_Mb = (window_pos_2 + window_pos_1)/(2*1e6))
  
  # extract number of individuals in vcf
  num_indiv <- as.numeric(gsub("sites/|fold/", "", str_split(pi_files[i], "_")[[1]][2]))
  site_class <- gsub("../data/diversity/", "", str_split(pi_files[i], "_")[[1]][1])
  
  # if diversity file excludes invariant sites, rescale denominator
  # count_comparisons = choose(<number of diploid individuals>*2,2)*no_sites
  if(num_indiv == 231 | num_indiv == 2029){
    cat("Rescaling pi denominator to assume all missing sites are invariant...\n")
    nuc_div$count_comparisons <- choose(num_indiv*2,2)*(nuc_div$window_pos_2 - nuc_div$window_pos_1 + 1)
    nuc_div$avg_pi <- nuc_div$count_diffs/nuc_div$count_comparisons
  }
  
  nuc_div$panel <- num_indiv
  nuc_div$site_class <- site_class
  nuc_div <- as.data.frame(nuc_div)
  nuc_div_all <- rbind(nuc_div_all, nuc_div)
  
}

#
nuc_div_density_all <- nuc_div_all %>%
  left_join(gene_density, by = c("chrom", "window_pos_1", "window_pos_2"))

# Evaluate derivative of model for windows
rate_vs_div_all <- nuc_div_density_all %>%
    left_join(best_mods, by = "chrom") %>%
    group_by(chrom) %>%
    mutate(rate = eval_deriv(model[[1]], window_pos_mid_Mb)) %>%
    select(chrom, window_pos_mid_Mb, panel, site_class, rate, avg_pi, frac_cds) %>%
    unnest(cols = c(rate))

rate_density_cor <- rate_vs_div_all %>%
  group_by(panel, site_class) %>%
  filter(rate <= 10) %>%
  reframe(r = corr_ci_autocorr(rate, frac_cds)$r,
          p = corr_ci_autocorr(rate, frac_cds)$p_value)

# plot density vs recombination rate
ggplot(rate_vs_div_all %>% filter(site_class == "all", panel == 231), aes(y = frac_cds, x = rate)) +
  geom_point(shape = 16) +
  geom_smooth(method = "lm", color = "skyblue", data = subset(rate_vs_div_all, site_class == "all" & panel == 231 & rate <= 10)) +
  facet_zoom(rate <= 10) +
    theme(
      zoom = element_rect(colour = "pink", linewidth = 0.8, fill = alpha("pink", 0.01))) +
  labs(x = "Fraction of window\nannotated CDS", 
       y = "Recombination rate (cM/Mb)",
title = paste0("r = ", signif(rate_density_cor$r, 2), ", p = ", format.pval(rate_density_cor$p, digits = 2, eps = 0.001)))

ggsave(paste0(results_path, "/gene_density_vs_recombination_rate.pdf"), 
       width = 4.5,
       height = 5.5, device = cairo_pdf)

# model of diversity as coding sequence density and recombination rate
mod <- lm(log10(avg_pi) ~ rate + frac_cds, data = rate_vs_div_all %>% filter(site_class == "all", panel == 231, rate <= 10))
summary(mod)
  
```

# Ne and VA(1) estimates

## re-do: standardize by half heterozygosity, weight

### one model per block

```{r}
# make list of sites
sites <- unique(gsub("_.*", "", colnames(hapfire)))
sites <- sites[-30]

# initial sum of site heterozygosity
sum_of_het_1 <- sum_of_het(gen0[,"X3"])

# get correlations that are encompassed within each block
blocks <- create_blocks(gen0, X1, X2, 10000, "_")

output_linked_params <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/", today, "/va_ne_points_for_estimation_10000b_nCorrectionOmitted.txt", sep = "")

#check_file_exists(output_linked_params)

weight = 1/length(unique(blocks$window))
#weight = 1

for(site in unique(sites)){
  print(paste("Site:", site))
  hap_site <- hapfire %>% 
    dplyr::select(starts_with(paste(site, "_", sep = ""))) %>% 
    collect()
  
  hap_site$window <- blocks$window
  
  for(exprep in 1:12){
    print(paste("Rep:", exprep))
    hap_rep <- hap_site %>% 
      dplyr::select(ends_with(paste("_", exprep, sep = "")), window)
    
    if(ncol(hap_rep) <= 2){
      print("Need at least two increments. Skipping to next replicate...")
      next
    }
    
    sum_of_het_vec <- sum_of_het_by_t(hap_rep %>% dplyr::select(-window))
    
    for(b in unique(hap_site$window)){
      
      hap_block <- hap_rep %>% 
        filter(window == b) %>% 
        dplyr::select(-window)
      
      #sum_of_het_vec <- sum_of_het_by_t(hap_block)
      
      gen0_block <- blocks %>% filter(window == b)
      
      #sum_of_het_1 <- sum_of_het(gen0_block[,"X3"])
      
      chrom = gen0_block$X1[1]
      start = min(gen0_block$X2)
      stop = max(gen0_block$X2)
      
      mean_ld_block <- ld_matrix %>% 
        filter( (CHR_A == chrom) & (CHR_B == chrom) & (BP_A >= start) & (BP_A <= stop) & (BP_B >= start) & (BP_B <= stop)) %>% 
        summarize(mean(R2)) %>% 
        collect()
  
      # sample sizes
      rep_n <- grab_sample_sizes(sample_sizes, site_gen_plot, colnames(hap_rep), 1000)
      
      # sort names to be in order of generation
      hap_block <- hap_block %>% 
        relocate(all_of(sort(colnames(hap_block))))
      
      # add initial generation
      hap_block <- cbind(gen0_block$X3, hap_block)
      #hap_rep <- cbind(gen0$X3, hap_rep)
      
      # organize data for linked selection parameters into a table
      linked_sel_param_block <- estim_linked_selection_params_new(hap_block, rep_n$n, sum_of_het_vec, sum_of_het_1, mean_ld_block$`mean(R2)`, weight, correct_for_n = F)
      
      linked_sel_param_block$site = site
      linked_sel_param_block$rep = exprep
      linked_sel_param_block$block = b
      linked_sel_param_block$S = nrow(hap_block)
      
      append_table(linked_sel_param_block, output_linked_params)
      rm(chrom, start, stop, hap_block, linked_sel_param_block, gen0_block, rep_n, mean_ld_block)
      gc()
    }
  }
}



```

### one model for whole genome

```{r}
output_linked_params_whole <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/", today, "/va_ne_points_for_estimation_whole_genome.txt", sep = "")

#check_file_exists(output_linked_params_whole)

# mean ld decay
mean_ld_block <- ld_matrix %>% 
        summarize(mean(R2)) %>% 
        collect()

# initial sum of site heterozygosity
sum_of_het_1 <- sum_of_het(gen0[,"X3"])

#
weight = 1

for(site in unique(sites)){
  print(paste("Site:", site))
  hap_site <- hapfire %>% 
    dplyr::select(starts_with(paste(site, "_", sep = ""))) %>% 
    collect()
  
  for(exprep in 1:12){
    print(paste("Rep:", exprep))
    hap_rep <- hap_site %>% 
      dplyr::select(ends_with(paste("_", exprep, sep = "")))
    
    # if(ncol(hap_rep) <= 2){
    #   print("Need at least two increments. Skipping to next replicate...")
    #   next
    # }
    
    sum_of_het_vec <- sum_of_het_by_t(hap_rep)
    
    # sample sizes
    rep_n <- grab_sample_sizes(sample_sizes, site_gen_plot, colnames(hap_rep), 1000)
    
    # sort names to be in order of generation
    hap_rep <- hap_rep %>% relocate(all_of(sort(colnames(hap_rep))))
    
    # add initial generation
    hap_rep <- cbind(gen0$X3, hap_rep)
    #hap_rep <- cbind(gen0$X3, hap_rep)
    names(hap_rep)[1] = paste(site, 0, exprep, sep = "_")
    
    # organize data for linked selection parameters into a table
    linked_sel_param_rep <- estim_linked_selection_params_new(hap_rep, rep_n$n, sum_of_het_vec, sum_of_het_1, mean_ld_block$`mean(R2)`, weight)
    
    linked_sel_param_rep$site = site
    linked_sel_param_rep$rep = exprep
    
    append_table(linked_sel_param_rep, output_linked_params_whole)
  }
}

```

### fit linear models

```{r}

### one model per genome
 va_ne_whole <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-06-29/va_ne_points_for_estimation_10000b.txt", fill = T) %>%
   distinct(site, rep, s, t, block, .keep_all = T)

# va_ne_whole <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-07-15/va_ne_points_for_estimation_10000b_nCorrectionOmitted.txt", fill = T) %>%
#   distinct(site, rep, s, t, block, .keep_all = T)

### plot relationships
# are there lots of outliers?
# are values structured by year?
for(i in unique(va_ne_whole$site)){
  data_subset <- va_ne_whole %>% 
    filter(s == t & site == i) %>%
    left_join(locations_data, by = "site")
  
  print(summary(lm(value ~ mean_ld + as.factor(s), data_subset)))
  
  if(nrow(data_subset) == 0){
    cat("Skipping site", i, "because there's no data...\n")
    next
  }
  
ggplot(data_subset, aes(x = mean_ld, y = value, color = as.factor(s))) +
  geom_point(shape = 16) +
  geom_smooth(method = "lm") +
  facet_wrap(.~rep) +
  scale_color_scico_d(palette = "lipari") +
  #scale_y_log10() +
  scale_y_continuous(trans = pseudo_log_trans(base = 10, sigma = min(data_subset$value[data_subset$value > 0])), breaks = pretty_breaks(n = 3)) +
  labs(x = "Mean LD", y = "Var(\u0394 pt)/(pt(1-pt))", color = "Generation", title = paste0("(", i, ") ", unique(data_subset$city_country)))

ggsave(paste(results_path, i, "_variance_all_vs_ld.png", sep = ""), height = 9, width = 9)
}

### try some custom modeling

# Think slopes per site make more sense than one global slope
site_slopes <- va_ne_whole %>%
  group_by(site) %>%
  filter(s == t) %>% # variances only
  group_modify(~ {
    m <- lmer(value ~ mean_ld + (1 | rep) + (1 | s), data = .x)
    tibble(
      slope = fixef(m)["mean_ld"],
      n_obs = nrow(.x)
    )
  }) %>%
  ungroup()

# generate block bootstrap sample
# sample the same blocks for every replicate WITHIN a site
block_bootstrap_sample <- function(dat) {

  # unique blocks available per site
  blocks_by_site <- dat %>% distinct(site, block)

  # resample blocks WITH replacement, once per site
  # (identical draw is then applied to all reps within that site)
  sampled_blocks <- blocks_by_site %>%
    group_by(site) %>%
    group_modify(~ {
      n_blocks <- length(unique(.x$block))
      tibble(
        block   = sample(.x$block, size = n_blocks, replace = TRUE),
        boot_id = seq_len(n_blocks)  # unique tag per draw, handles duplicate blocks
      )
    }) %>%
    ungroup()

  # pull ALL rows (all reps) matching each sampled block instance
  boot_dat <- sampled_blocks %>%
    left_join(dat, by = c("site", "block"),
              relationship = "many-to-many")

  boot_dat
}

# one bootstrap replicate:
boot_sample <- block_bootstrap_sample(va_ne_whole %>% filter(s == t))

# full block bootstrapping procedure
B <- 1000
output_boot_slopes <- paste0(results_path, "vardp_meanld_slope_bootstraps.txt")
#check_file_exists(output_boot_slopes)

for (i in 1:B) {
  
  cat("Bootstrap sample: ", i, "\n")
  boot_dat <- block_bootstrap_sample(va_ne_whole %>% filter(s == t))

  boot_results <- boot_dat %>% 
    group_by(site) %>%
  filter(s == t) %>% # variances only
  group_modify(~ {
    m <- lmer(value ~ mean_ld + (1 | rep) + (1 | s), data = .x)
    tibble(
      slope = fixef(m)["mean_ld"],
      n_obs = nrow(.x)
    )
  }) %>%
  ungroup()

  boot_results$boot <- boot_results$boot <- i
  append_table(boot_results, output_boot_slopes)
}

### plot lmm results
lmm_results <- fread(output_boot_slopes)

conf_level <- 0.05 

lmm_cis <- lmm_results %>%
  group_by(site) %>%
  summarize(lwr_slope = quantile(slope, conf_level/2),
            upr_slope = quantile(slope, 1-conf_level/2)) %>%
  mutate(significant = ((lwr_slope > 0) & (upr_slope > 0)) | ((lwr_slope < 0) & (upr_slope < 0))) %>%
  left_join(site_slopes, by = "site")


for(sigma in c(1e-5, 1e-4, 1e-3)){

  ggplot(lmm_cis, aes(x = reorder(site, -slope), y = slope, color = significant)) +
  geom_pointrange(aes(ymin = lwr_slope, ymax = upr_slope)) +
  scale_y_continuous(trans = pseudo_log_trans(base = 10, sigma = sigma)) +
  scale_color_scico_d(palette = "bam") +
    labs(x = "Location", y = "Var(\u0394 pt)/(pt(1-pt)) vs LD slope") +
    theme(legend.position = "none")
  
  ggsave(paste0(results_path, "vardp_ld_slopes_", sigma, ".pdf"), device = cairo_pdf)
  
}






















#### DEPRECATED ####


### estimate va and ne
mod_genome <- va_ne_whole %>%
  group_by(site, rep) %>%
  filter(s == 1 & t == 1 & value > 0) %>%
  #mutate(a = 1/a) %>%
  #filter(s != t) %>%
  #filter(s == t) %>%
  #filter((sum_of_het_ratio < 1) & (value > 0))
  nest() %>%
  mutate(
    # Fit a linear model to each nested group
    #fit = purrr::map(data, ~ MASS::rlm(value ~ a + b - 1, data = .x)),
    fit = purrr::map(data, ~ lm(value ~ a + b - 1, data = .x)),
    # Extract tidy coefficients (intercept and slope)
    tidied = purrr::map(fit, tidy)
  ) %>%
  # Expand the tidied results into a standard data frame
  unnest(tidied) %>% 
  dplyr::select(c(-data, -fit))

va_per_site <- mod_genome %>% 
  filter(estimate > 0 & term == "a") %>% # some reps with negative estimates for Ne are being included here, same for Ne per site
  group_by(site) %>%
  summarize(mean_va = mean(estimate), n = n())



ne_per_site <- mod_genome %>% 
  filter(estimate > 0 & term == "b") %>%
  group_by(site) %>%
  summarize(mean_Ne = 1/mean(estimate), n = n())


### bootstrapping
W = 299
B = 1000
boot_samples <- NULL
for(i in 1:B){
  
  boot_sample <- va_ne_whole %>%
  group_by(site, rep) %>%
  filter(s == 1 & t == 1 & value > 0) %>%
  slice_sample(n = W, replace = TRUE) %>%  # Sample blocks with replacement
  nest() %>%
  mutate(
    fit = purrr::map(data, ~ lm(value ~ a + b - 1, data = .x)),
    tidied = purrr::map(fit, tidy)
  ) %>%
  unnest(tidied) %>% 
  dplyr::select(c(-data, -fit))
  
  boot_sample$boot <- i
  
  boot_samples <- rbind(boot_samples, boot_sample)
}

va_per_site_boot <- boot_samples %>% 
  filter(estimate > 0 & term == "a") %>% # some reps with negative estimates for Ne are being included here, same for Ne per site
  group_by(site, boot) %>%
  summarize(mean_va = mean(estimate)) %>%
  summarize(lwr = quantile(mean_va, 0.005),
            med = median(mean_va),
            upr = quantile(mean_va, 0.995),
            n = n())

va_plot_data <- va_per_site %>% 
  left_join(va_per_site_boot, by = "site") %>% 
  mutate(VALID = mean_va > lwr) %>%
  left_join(worldclim_sitesdata, by = "site")

ne_per_site_boot <- boot_samples %>% 
  filter(estimate > 0 & term == "b") %>% # some reps with negative estimates for Ne are being included here, same for Ne per site
  group_by(site, boot) %>%
  summarize(mean_Ne = 1/mean(estimate)) %>%
  summarize(lwr = quantile(mean_Ne, 0.005),
            med = median(mean_Ne),
            upr = quantile(mean_Ne, 0.995),
            n = n())

ne_plot_data <- ne_per_site %>% 
  left_join(ne_per_site_boot, by = "site") %>% 
  mutate(VALID = mean_Ne > lwr) %>%
  left_join(worldclim_sitesdata, by = "site")

# look at correlations with climate
plot_va_bio1 <- ggplot() +
  geom_pointrange(data = va_plot_data, mapping = aes(y = mean_va, ymin = lwr, ymax = upr, x = bio1)) +
  geom_smooth(
      data = subset(va_plot_data, bio1 <= 19), 
      mapping = aes(x = bio1, y = mean_va),
      method = "lm", 
      se = TRUE, 
      color = "lightblue"
    ) +
  scale_y_log10() +
  labs(x = "Mean Annual Temperature (\U00B0 C)", y = "VA")

save_plot_and_object(plot_obj = plot_va_bio1, base_path = paste(results_path, "va_per_site", sep = ""), img_ext = "pdf")

plot_ne_bio1 <- ggplot() +
    geom_pointrange(data = ne_plot_data, mapping = aes(y = mean_Ne, ymin = lwr, ymax = upr, x = bio1)) +
  geom_smooth(
      data = subset(ne_plot_data, bio1 <= 19), 
      mapping = aes(x = bio1, y = mean_Ne),
      method = "lm", 
      se = TRUE, 
      color = "lightblue"
    ) +
  scale_y_log10()  +
  labs(x = "Mean Annual Temperature (\U00B0 C)", y = "Ne")

save_plot_and_object(plot_obj = plot_ne_bio1, base_path = paste(results_path, "ne_per_site", sep = ""), img_ext = "pdf")


#### DEPRECATED ####















# boot_cis <- boot_samples %>%
#   group_by(site, rep, term) %>%
#   summarize(lwr = quantile(boot_estimate, 0.025),
#             med = median(boot_estimate),
#             upr = quantile(boot_estimate, 0.975))




### one model per block
# va_ne_blocks <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-06-25/va_ne_points_for_estimation_single_locus_version.txt") %>%
#   distinct(site, rep, block, s, t, .keep_all = T)

# va_ne_blocks <- va_ne_blocks %>%
#   mutate(a = (1/sum_of_het_ratio)*0.5*mean_ld)

ggplot(subset(va_ne_blocks, t == 1 & s == 1 & value > 0), aes(x = a, y = value)) +
  geom_point(aes(color = mean_ld)) +
  geom_smooth(method = "lm") +       # Linear regression line
  stat_regline_equation() +
  facet_wrap(.~site + rep, scales = "free") +
  labs(x = "a", y = "Var(dp)/(p(1-p))", color = "Mean LD", title = "First increment variances")

ggsave(paste("../results/by-date/", today, "/va1_ne_bc_regression_single_locus_version.pdf", sep = ""), width = 15, height = 10, device = cairo_pdf())

# un-do window weighting
va_ne_blocks <- va_ne_blocks %>%
  mutate(c = 0.5*mean_ld)

va_ne_blocks %>%
  group_by(site, rep) %>%
  filter(s <= 1 & t <= 1) %>%
  #filter((sum_of_het_ratio < 1) & (value > 0))
  nest() %>%
  mutate(
    # Fit a linear model to each nested group
    fit = purrr::map(data, ~ lm(value ~ c + b - 1, data = .x)),
    # Extract tidy coefficients (intercept and slope)
    tidied = purrr::map(fit, tidy)
  ) %>%
  # Expand the tidied results into a standard data frame
  unnest(tidied) %>% 
  dplyr::select(c(-data, -fit))

# va_ne_blocks <- va_ne_blocks %>% 
#   filter(t <= 2 & s <= 2)
  #filter((sum_of_het_ratio < 1) & (value > 0))

# check for multicollinearity
va_ne_blocks %>%
  group_by(site, rep) %>%
  summarize(cor(a,b))

mod_first_gen <- va_ne_blocks %>%
  group_by(site, rep) %>%
  filter(s <= 1 & t <= 1) %>%
  #filter((sum_of_het_ratio < 1) & (value > 0))
  nest() %>%
  mutate(
    # Fit a linear model to each nested group
    fit = purrr::map(data, ~ lm(value ~ a + b - 1, data = .x)),
    # Extract tidy coefficients (intercept and slope)
    tidied = purrr::map(fit, tidy)
  ) %>%
  # Expand the tidied results into a standard data frame
  unnest(tidied) %>% 
  dplyr::select(c(-data, -fit))

mod_first_gen %>% 
  ungroup() %>%
  filter(term == "a") %>% 
  summarize(sum(estimate > 0))

mod_genome <- va_ne_blocks %>%
  group_by(site, rep) %>%
  #filter(s <= 2 & t <= 2) %>%
  #filter((sum_of_het_ratio < 1) & (value > 0))
  nest() %>%
  mutate(
    # Fit a linear model to each nested group
    fit = purrr::map(data, ~ lm(value ~ a + b - 1, data = .x)),
    # Extract tidy coefficients (intercept and slope)
    tidied = purrr::map(fit, tidy)
  ) %>%
  # Expand the tidied results into a standard data frame
  unnest(tidied) %>% 
  dplyr::select(c(-data, -fit))

mod_genome %>% 
  ungroup() %>%
  filter(term == "a") %>% 
  summarize(sum(estimate > 0))

mod_block <- va_ne_blocks %>%
  group_by(site, rep, block) %>%
  #filter(s <= 2 & t <= 2) %>%
  nest() %>%
  mutate(
    # Fit a linear model to each nested group
    fit = purrr::map(data, ~ lm(value ~ a + b - 1, data = .x)),
    # Extract tidy coefficients (intercept and slope)
    tidied = purrr::map(fit, tidy)
  ) %>%
  # Expand the tidied results into a standard data frame
  unnest(tidied) %>% 
  dplyr::select(c(-data, -fit))

mod_off_diag <- va_ne_blocks %>%
  group_by(site, rep) %>%
  filter(s != t) %>%
  nest() %>%
  mutate(
    # Fit a linear model to each nested group
    fit = purrr::map(data, ~ lm(value ~ a + b - 1, data = .x)),
    # Extract tidy coefficients (intercept and slope)
    tidied = purrr::map(fit, tidy)
  ) %>%
  # Expand the tidied results into a standard data frame
  unnest(tidied) %>% 
  dplyr::select(c(-data, -fit))

mod_off_diag %>% 
  ungroup() %>%
  filter(term == "a") %>% 
  summarize(sum(estimate > 0))

mod_diag <- va_ne_blocks %>%
  group_by(site, rep) %>%
  filter(s == t) %>%
  nest() %>%
  mutate(
    # Fit a linear model to each nested group
    fit = purrr::map(data, ~ lm(value ~ a + b - 1, data = .x)),
    # Extract tidy coefficients (intercept and slope)
    tidied = purrr::map(fit, tidy)
  ) %>%
  # Expand the tidied results into a standard data frame
  unnest(tidied) %>% 
  dplyr::select(c(-data, -fit))

mod_diag %>% 
  ungroup() %>%
  filter(term == "a") %>% 
  summarize(sum(estimate > 0))

mean_est_genome <- mod_genome %>%
  group_by(site, term) %>%
  summarize(estimate = mean(estimate))
  #filter(estimate < 0)

mean_est_block <- mod_block %>%
  group_by(site, rep, term) %>%
  summarize(estimate = mean(estimate)) %>%
  group_by(site, term) %>%
  summarize(estiamte = mean(estimate))

va1_ne_est <- mod_genome %>% 
  subset(term == "b") %>% 
  mutate(Ne=1/(2*estimate)) %>%
  left_join(mod_genome %>% subset(term == "a"), by = c("site", "rep")) %>%
  mutate(VA1 = estimate.y) %>%
  dplyr::select(site, rep, Ne, VA1)

ggplot(va1_ne_est, aes(x = Ne, y = VA1)) +
  geom_point() +
  scale_y_log10() +
  scale_x_log10()

## bootstrap
W = 200
B = 200
boot_samples <- NULL
for(i in 1:B){
  
  boot_sample <- mod_block %>%
    group_by(site,rep,term) %>%
    slice_sample(n = W, replace = TRUE) %>%  # Sample blocks with replacement
    summarize(boot_estimate = mean(estimate))
  
  boot_samples <- rbind(boot_samples, boot_sample)
}

boot_cis <- boot_samples %>%
  group_by(site, rep, term) %>%
  summarize(lwr = quantile(boot_estimate, 0.025),
            med = median(boot_estimate),
            upr = quantile(boot_estimate, 0.975))

boot_cis %>%
  subset(term == "a") 
  
boot_cis %>%
  subset(term == "b") %>% 
  mutate(Ne_lwr=1/(2*upr),
         Ne_med=1/(2*med),
         Ne_upr=1/(2*lwr))
```

## calculate average ld or average ld decay by block

```{r}

# output file
#output_mean_ld <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/", today, "/mean_ld_by_block.tsv", sep = "")
output_mean_ld <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-05-05/mean_ld_by_block.tsv", sep = "")

# read ld_matrix in blocks and calculate average initial R2
ld_matrix <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/plink.ld.tsv")

# get correlations that are encompassed within each block
blocks <- create_blocks(gen0, X1, X2, 15000, "_")

blocks <- blocks %>% 
  group_by(window) %>% 
  summarize(chr = unique(X1), start = min(X2), end = max(X2))

for(i in 1:nrow(blocks)){
  print(i)
  print(Sys.time())
  
  b = blocks[i,]
  start = b$start
  end = b$end
  chr = b$chr
  w = b$window
  
  result <- data.frame(window = w,
                       mean_ld = ld_matrix %>% 
                         filter((CHR_A == chr) & (CHR_B == chr) & (BP_A >= start) & (BP_A <= end) & (BP_B >= start) & (BP_B <= end)) %>% summarize(mean(R2))
  )
  
  write.table(result, output_mean_ld, col.names=!file.exists(output_mean_ld), append = T, row.names = F, sep = ",", quote = F)
    
}

# 
# ld_matrix %>% summarize(mean(R2))

# process_ld <- function(output){
#   function(x, pos) {
#     print(pos)
#     print(Sys.time())
#     y <- x %>% summarize(chr = unique(CHR_A),
#                     start = min(c(BP_A, BP_B)),
#                     end = max(c(BP_A, BP_B)),
#                     avg_r2 = mean(R2),
#                     n = length(unique(c(SNP_A, SNP_B))))
#     
#     write.table(y, output, col.names=!file.exists(output), append = T, row.names = F, sep = "\t", quote = F)
#   }
# }
# 
# mean_ld <- read_tsv_chunked(
#   "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/plink.ld.tsv",
#   callback = SideEffectChunkCallback$new(process_ld("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/plink.ld.mean.tsv")),
#   chunk_size = 99*15000)
```

## use least squares to estimate Ne and Va(1)

Need:
* allele frequencies
* sample sizes
* mean LD decay in windows through time, or just LD in initial generation if assuming no decay through time

```{r}

# load mean ld
mean_ld_blocks <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-05-05/mean_ld_by_block.tsv")

# assign blocks to allele frequencies
# number of snps per window
window_size <- 15000

# create windows, should not bleed between chromosomes
blocks <- gen0 %>%
  group_by(X1) %>%
  mutate(block = (row_number() - 1) %/% window_size + 1,
         window = paste(X1, block, sep = "_"))

length(unique(blocks$window))

hapfire$window <- blocks$window

# make list of sites
sites <- unique(gsub("_.*", "", colnames(hapfire)))
sites <- sites[c(-30, -32)]

linked_params <- NULL

output_linked_params <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/", today, "/va_ne_points_for_estimation_no_fix.txt", sep = "")

# loop over genomic blocks
num_blocks <- length(unique(hapfire$window))

# initial sum of site heterozygosity
sum_of_het_1 <- sum_of_het(gen0[,"X3"])

for(site in unique(sites)){
  print(paste("Site:", site))
  hap_site <- hapfire %>% dplyr::select(starts_with(paste(site, "_", sep = "")), window)
  
  for(exprep in 1:12){
    print(paste("Rep:", exprep))
    hap_rep <- hap_site %>% dplyr::select(ends_with(paste("_", exprep, sep = "")), window)
    
    if(ncol(hap_rep) <= 2){
      print("Need at least two increments. Skipping to next replicate...")
      next
    }
    
    sum_of_het_vec <- sum_of_het_by_t(hap_rep %>% dplyr::select(-window))
    
    for(b in unique(hapfire$window)){
      
      hap_block <- hap_rep %>% filter(window == b) %>% dplyr::select(-window)
      
      gen0_block <- blocks %>% filter(window == b)
      
      mean_ld_block <- mean_ld_blocks %>% filter(window == b)
      
      # get sample sizes
      # rep_n <- sample_sizes[(sample_sizes$site_gen_plot %in% colnames(hap_block)),]
      # rep_n <- rbind(c(0, 1e3), rep_n)
      # rep_n <- rep_n %>% arrange(site_gen_plot)
      
      rep_n <- grab_sample_sizes(sample_sizes, site_gen_plot, colnames(hap_block), 1000)
      
      # sort names to be in order of generation
      hap_block <- hap_block %>% relocate(all_of(sort(colnames(hap_block))))
      
      hap_block <- cbind(gen0_block$X3, hap_block)
      
      names(hap_block)[1] = 0
      
      # create table of linked selection covariates for linear model
      linked_param_block <- estim_linked_selection_params(hap_block, sum_of_het_vec, sum_of_het_1, rep_n$n, mean_ld_block$mean.R2., 1/num_blocks)
      
      linked_param_block$site = site
      linked_param_block$rep = exprep
      linked_param_block$block = b
      linked_param_block$S = nrow(hap_block)
      
      #write.table(linked_param_block, output_linked_params, col.names=!file.exists(output_linked_params), append = T, row.names = F, sep = ",", quote = F)
      
      append_table(linked_param_block, output_linked_params)
      
      #linked_params <- rbind(linked_params, linked_param_block)
    }
  }
}





```

## fit linear model by location

```{r}

# first attempt, observed multicollinearity between Ne and VA(1)
#va_ne_blocks <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-05-06/va_ne_points_for_estimation.txt")

# second attempt, fixed summation in SSH(s) but not SSH(1)
#va_ne_blocks <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-05-18/va_ne_points_for_estimation.txt")

# third attempt, fixed summation in SSH(1) too
#va_ne_blocks <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-05-20/va_ne_points_for_estimation.txt")
#va_ne_blocks <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-05-21/va_ne_points_for_estimation.txt")

# fourth attempt, fixed sites marked as NA
#va_ne_blocks <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-05-21/va_ne_points_for_estimation_no_fix.txt")
va_ne_blocks <- fread("../results/by-date/2026-05-21/va_ne_points_for_estimation_no_fix.txt")

# quick plot of ssh over time
ggplot(va_ne_blocks, aes(x = t, y = sum_of_het_s, color = rep, group = interaction(site, rep, block))) +
  geom_line(alpha = 0.2) +
  facet_wrap(.~site) +
  labs(y = "SSH", x = "t")

ggsave(paste(results_path, "SSH_time_series_no_fix.pdf", sep = ""), width = 8, height = 8)

# plot predictors vs response
ggplot(va_ne_blocks, aes(x = a, y = value, color = as.factor(b))) +
  geom_point(shape = 16, alpha = 0.2) +
  facet_wrap(.~site, scales = "free") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "1/2 SSH(s)/SSH(1) A(R,t,s)", y = "Variance/Covariance value", color = "Var/Cov")

ggsave(paste(results_path, "cov_vs_ssh.pdf", sep = ""), width = 10, height = 10)

# check for multi-collinearity
va_ne_blocks %>%
      group_by(site) %>%
  filter(value != 0) %>%
      summarize(
        cor_a_y = cor(a, abs(value), method = "spearman"),
        cor_b_y = cor(b, abs(value), method = "spearman"),
        cor_a_b = cor(a, b, method = "spearman")
      )

ggplot(va_ne_blocks, aes(x = as.factor(b), y = a)) +
  geom_violin(draw_quantiles = c(0.25, 0.5, 0.75)) +
  facet_wrap(.~site) +
  labs(y = "1/2 SSH(s)/SSH(1) A(R,t,s)", x = "Variance/Covariance value")

ggsave(paste(results_path, "ne_va1_multicollinearity_check.pdf", sep = ""), width = 10, height = 10)

# bootstrap estimates
B = 200 # number of bootstrap samples per site
W = 200 # number of windows in the genome
alpha = 0.05 # ci width

cis <- NULL

for(focal_site in unique(va_ne_blocks$site)){
  
  print(paste("Site:", focal_site))
  boot_samples = NULL
  boot_site <- va_ne_blocks %>%
    filter(site == focal_site)
  
  for(focal_rep in 1:12){
    print(paste("Rep:", focal_rep))
    boot_rep <- boot_site %>%
      filter(rep == focal_rep)
    
    if(nrow(boot_rep) == 0){
      print("No replicate found. Skipping to next replicate...")
      next
    }
    
    for(b in 1:B){
      
      boot_sample <- boot_rep %>%
        nest_by(block) %>% # Group and nest into a list-column
        ungroup() %>%
        slice_sample(n = W, replace = TRUE) %>%  # Sample blocks with replacement
        dplyr::select(block, data) %>%
        unnest(cols = c(data)) %>%
        #filter(value > 1e-5) %>%
        #lm(value ~ b + a - 1, data = .) %>%
        lm(abs(value) ~ b + a - 1, data = .) %>%
        #lm(abs(value) ~ b + a, data = .) %>%
        #vif()
        tidy()
      
      #mutate(
      #cor_a_b = map(data, ~ cor(.x$a, .x$b)),
      # Fit a linear model to each nested group
      #fit = map(data, ~ lm(abs(value) ~ a + b - 1, data = .x)),
      #fit = map(data, ~ lm(value ~ b + a - 1, data = .x)),
      #fit = map(data, ~ lm(value ~ a - 1, data = .x)),
      #fit = map(data, ~ lm(value ~ b - 1, data = .x)),
      # Extract tidy coefficients (intercept and slope)
      #tidied = map(fit, tidy)
      #) %>%
      # Expand the tidied results into a standard data frame
      #unnest(tidied) %>% 
      #dplyr::select(c(-data, -fit))
      
      boot_samples <- rbind(boot_samples, boot_sample)
    }
    
    ci <- boot_samples %>% 
      group_by(term) %>% 
      summarize(lwr = quantile(estimate, alpha/2), 
                med = median(estimate), 
                upr = quantile(estimate, 1-alpha/2))
    
    ci$site <- focal_site
    ci$rep <- focal_rep
    
    cis <- rbind(cis, ci)
  }
}

# save bootstrapping results
write.table(cis, paste(results_path, "VA1_Ne_estimates_200_boots_abs.txt", sep = ""), row.names = F, quote = F, sep = "\t")
```

### POINT ESTIMATES

Fit linear model for each site and replicate. Can look at only absolute value of covariances

```{r}
va_ne_models <- va_ne_blocks %>%
  group_by(site, rep) %>%
  nest() %>%
  mutate(
    # Fit a linear model to each nested group
    fit = map(data, ~ lm(value ~ a + b - 1, data = .x)),
    # Extract tidy coefficients (intercept and slope)
    tidied = map(fit, tidy)
  ) %>%
  # Expand the tidied results into a standard data frame
  unnest(tidied) %>% 
  dplyr::select(c(-data, -fit))

va_ne_site_means <- va_ne_models %>%
  group_by(site, term) %>%
  summarize(
    mean_est = mean(estimate),
    sd_est = sd(estimate),
    n = n(),
    se_est = sd_est / n
  )

ggplot(va_ne_models, aes(x = estimate)) +
  geom_histogram() +
  facet_grid(.~term, scales = "free")

avg_ne <- va_ne_models %>% 
  subset(term == "b") %>% 
  mutate(Ne=1/(2*estimate)) %>% 
  group_by(site) %>%
  summarize(sd = sd(Ne), n = n(), se = sd/n, Ne=mean(Ne)) %>%
  arrange(Ne)


```

# multicollinearity simulations

```{r}
b0 = 0
b1 = 13
b2 = -13

sims <- do.call("rbind", replicate(100, multicol_sim(1000, 10, 10, b0, b1, b2, matrix(c(50,5,5,50), nrow = 2)), simplify = F))

ggplot(sims, aes(x = value, color = term)) +
  geom_vline(xintercept = c(b0, b1, b2), lty = 2) +
  geom_density() +
  facet_wrap(.~type+term, scales = "free")
```

# Waples 1989 Ne estimator

```{r}
# loop over each replicate
# get two furthest time points, separated by at least 2 generations

# make list of sites
sites <- unique(gsub("_.*", "", colnames(hapfire)))
sites <- sites[c(-30)]

output_ne <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/", today, "/waples_1989_ne_esitmates.txt", sep = "")

check_file_exists(output_ne)

for(site in as.numeric(unique(sites))){
  local({
  print(paste("Site:", site))
  hap_site <- hapfire %>% 
    dplyr::select(starts_with(paste(site, "_", sep = "")))
  
  for(exprep in 1:12){
    print(paste("Rep:", exprep))
    hap_rep <- hap_site %>% 
      dplyr::select(ends_with(paste("_", exprep, sep = ""))) %>%
      collect()
    
    # sort names to be in order of generation
    hap_rep <- hap_rep %>% relocate(all_of(sort(colnames(hap_rep))))
    
    if(ncol(hap_rep) == 0){
      print("No observations. Skipping replicate...")
      next
    }
    
    # get last column == last generation
    pt <- hap_rep %>% pull(ncol(hap_rep))
    
    # sample sizes
    rep_n <- grab_sample_sizes(sample_sizes, site_gen_plot, colnames(hap_rep), 1000)
    
    # check for minimum separation to avoid negative Ne estimates
    rep_time <- as.numeric(gsub("_", "", str_extract(colnames(hap_rep), "_._")))
    rep_time <- rep_time[length(rep_time)]
    
    # if(rep_time < 2){
    #   print("Too close to t too close to 0. Skipping replicate...")
    #   next
    # }
    
    # calculate Ne
    ne_estimate <- waples_ne(1.8, 20, gen0$X3, pt, 1000, rep_n$n[nrow(rep_n)], rep_time)
    
    ne_result <- data.frame(
      site = site,
      rep = exprep,
      fc = fc(gen0$X3, pt),
      ne_low = ne_estimate[1],
      ne_high = ne_estimate[2],
      t = rep_time
    )
    
    append_table(ne_result, output_ne)
  }
  })
}

```

## plot waples Ne estimates

```{r}
ne_estimates <- read.csv("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-07-13/waples_1989_ne_esitmates.txt")

ne_summary <- ne_estimates %>%
  #filter(ne_low > 0) %>%
  group_by(site) %>%
  summarize(mufc = mean(fc),
            sefc = sd(fc)/sqrt(n()),
            tmin = min(t),
            tmax = max(t)) %>%
  left_join(locations_data, by = "site")

ne_summary$site_city_country <- paste0("(", ne_summary$site, ") ", ne_summary$city_country)

ggplot(ne_summary, aes(x = reorder(site_city_country, -mufc), 
                       y = mufc, 
                       group = as.factor(site))) +
  geom_pointrange(aes(ymin = mufc - 2*sefc, 
                      ymax = mufc+2*sefc), 
                  position = position_dodge2(width = 0.75)) +
  theme(axis.text.x = element_text(angle = 75, hjust = 1),
        text = element_text(size = 12)) +
  labs(x = "Site", y = "Fb")

ggsave(paste0(results_path, "fc_estimates.pdf"), height = 4, width = 8)
```

# G' (G prime)

```{r}
# ne estimates
ne_estimates <- fread("/global/home/users/milesroberts/moi_lab_projects/grenenet-phase2/results/by-date/2026-05-31/VA1_Ne_estimates_200_boots_abs.txt")

ne_estimates <- ne_estimates %>% 
  filter(term == "b") %>% mutate(Ne = 1/(2*med)) %>%
  dplyr::select(site, rep, Ne) %>%
  as.data.frame()

# make list of sites
sites <- unique(gsub("_.*", "", colnames(hapfire)))
sites <- sites[c(-30, -32)]

output_gp <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/", today, "/gprime_point_esitmates.txt", sep = "")

for(site in as.numeric(unique(sites))){
  print(paste("Site:", site))
  hap_site <- hapfire %>% dplyr::select(starts_with(paste(site, "_", sep = "")))
  
  for(exprep in 1:12){
    print(paste("Rep:", exprep))
    hap_rep <- hap_site %>% dplyr::select(ends_with(paste("_", exprep, sep = "")))
    
    if(ncol(hap_rep) <= 2){
      print("Need at least two increments. Skipping to next replicate...")
      next
    }
    
    # sort names to be in order of generation
    hap_rep <- hap_rep %>% relocate(all_of(sort(colnames(hap_rep))))
    
    # sample sizes
    rep_n <- grab_sample_sizes(sample_sizes, site_gen_plot, colnames(hap_rep), 1000)
    
    # combine 
    hap_rep <- cbind(gen0$X3, hap_rep)
    names(hap_rep)[1] = 0
    
    # grab times
    times <- as.numeric(gsub("_", "", str_extract(colnames(hap_rep), "_.*_")))
    times <- times[-1]
    
    # grab Ne estimate
    Ne <- ne_estimates[(ne_estimates$site == site) & (ne_estimates$rep == exprep), "Ne"]
    
    if(Ne <= 2){
      print("Ne should be >= 2 Skipping to next replicate...")
      next
    }
    
    # estimate G prime
    gp_result <- data.frame(
      site = site,
      rep = exprep,
      gen = c(0, times),
      gp_values = c(0, g_prime(times, hap_rep, Ne, rep_n$n, take_abs = T))
      
    )
    
    append_table(gp_result, output_gp)
  }
}



```

# Final figures

## Feder test

```{r}
# load results of test
# remove invalid points
feder <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-03-31/feder_test_phase2.txt")

# calculate averages and variances for a snp at a site
# feder <- feder %>% group_by(site, CHROM, POS) %>% 
#   summarize(ybar = mean(y, na.rm = T), s2 = var(y, na.rm = T), n = n())
# 
# feder$tvalue <- feder$ybar/sqrt(feder$s2/feder$n)
# feder$pvalue <- dt(feder$tvalue, df = feder$n)

feder <- feder %>%
  filter(is.finite(ybar))

# choose which points to highlight
# Chr1:1185512..1188696 - CRY2
# Chr1:24331373..24333999 - FT
# Chr2:11058944..11063324 - ELF3
# Chr2:11531967..11534358 - CAM5
# Chr2:18807505..18811125 - SOC1
# Chr4:9195247..9199840 - PHYD
# Chr4:11000771..11003010 - TSF
# Chr5:5170162..5173085 - CO
# Chr5:3173382..3179448 - FLC
# Chr5:18589438..18591506 - DOG1

feder$highlight <- "no"
buffer = 10000
chroms = c(1,1,2,2,2,4,4,5,5,5)
starts = c(1185512,24331373,11058944,11531967,18807505,9195247,11000771,5170162,3173382,18589438)
ends = c(1188696,24333999,11063324,11534358,18811125,9199840,11003010,5173085,3179448,18591506)
starts = starts - buffer
ends = ends + buffer

for(i in seq_along(chroms)){
  print(i)
  feder$highlight[(feder$CHROM == chroms[i]) & (feder$POS > starts[i]) & (feder$POS < ends[i])] <- "yes"
}

# test out a few plots
#plot_manhattan_by_site(data = feder, site = 1, today, height = 3, width = 8, quick = F, prefix = "manhattan_flowering_time")

# plot them ALL!
map(unique(feder$SITE), ~plot_manhattan_by_site(data = feder, site = .x, today, height = 3, width = 8, quick = F, prefix = "manhattan_flowering_time"))

# drill down to certain peaks
# s10_c3 <- feder %>% subset(SITE == 10 & CHROM == 3)
# s10_c5 <- feder %>% subset(SITE == 10 & CHROM == 5)
# s32_c1 <- feder %>% subset(SITE == 32 & CHROM == 1)
```

## GLM

```{r}
glm_results <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-05-04/glm_quasibinomial_coef.txt")

names(glm_results)[c(1,2,6,7)] = c("CHROM", "POS", "pvalue", "SITE")

# remove duplicates from re-running glm
glm_results <- glm_results %>%
  distinct(CHROM, POS, SITE, .keep_all = TRUE)

glm_results$pvalue_adj <- p.adjust(glm_results$pvalue, method = "fdr")

glm_results$highlight <- "no"

glm_results$highlight[(glm_results$pvalue_adj < 1e-14)] <- "yes"

for(site in unique(glm_results$SITE)){
  plot_manhattan_by_site(glm_results, site = site, today = today, height = 7, width = 7, prefix = "glm_manhattan")
}

```

## Simulations

```{r}
#sim_curves <- fread("simulated_gt_curves_2026-03-31.txt")
#sim_curves <- fread("simulated_gt_curves_2026-04-08.txt")
#sim_curves <- fread("simulated_gt_curves_2026-04-14.txt")
#sim_curves <- fread("../results/2026-04-16/simulated_gt_curves_2026-04-16.txt")
#sim_curves <- fread("../results/2026-04-26/simulated_gt_curves_2026-04-26.txt")
#sim_curves <- fread("../results/2026-04-27/simulated_gt_curves_2026-04-27.txt")
#sim_curves <- fread("../results/2026-04-28/simulated_gt_curves_2026-04-28.txt")
#sim_curves <- fread("../results/2026-06-15/simulated_gt_curves_2026-06-15.txt")
#sim_curves <- fread("../results/2026-06-15/simulated_gt_curves_with_neutral_2026-06-15.txt")

#sim_curves <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-06-17/simulated_gt_curves_2026-06-17.txt")
#sim_curves <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-06-19/simulated_gt_curves_2026-06-19.txt")
sim_curves <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-06-26/simulated_gt_curves_2026-06-26.txt")

sim_curves$sum_abs_cov/sim_curves$sum_var

#sim_curves$sum_abs_cov/sim_curves$med_sum_abs_cov

#sim_curves <- merge(params[,c("ID", "MIN_AGE")], sim_curves, by = "ID")

# average across reps within a simulation parameter combination
mean_sims = sim_curves %>% 
  group_by(gen, SIGMA, L, VA, adjust) %>% 
  summarize(gt = mean(sum_cov)/mean(sum_cov + sum_var),
            gt_pos = mean(sum_pos_cov)/mean(sum_abs_cov + sum_var),
            gt_abs = mean(sum_abs_cov)/mean(sum_abs_cov + sum_var),
            #gt_abs_2 = mean(sum_abs_cov)/mean(sum_cov + sum_var),
            #gt_abs_cor = mean(cor_sum_abs_cov)/mean(sum_abs_cov + sum_var),
            sum_cov = mean(sum_cov), 
            sum_abs_cov = mean(sum_abs_cov),
            sum_pos_cov = mean(sum_pos_cov),
            sum_neg_cov = mean(sum_neg_cov),
            sum_var = mean(sum_var), 
            ratio = mean(ratio, na.rm = T),
            n = n())

mean_sims$adjust[mean_sims$adjust] <- "Ne constant"
mean_sims$adjust[(mean_sims$adjust == "FALSE")] <- "Ne lowered"

# mean_sims$MIN_AGE[(mean_sims$MIN_AGE == 1)] <- "Non-WF"
# mean_sims$MIN_AGE[(mean_sims$MIN_AGE == 0)] <- "Non-WF reduced"

# gt over time
# ggplot(mean_sims, aes(x = gen, y = gt_abs_cor, group = SIGMA, color = as.factor(SIGMA))) +
#   geom_point(size = 1.5) +
#   scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
#   facet_grid(adjust~VA+L) +
#   geom_hline(yintercept = 0, lwd = 1.1, lty = 2) +
#   geom_line() +
#   theme(legend.position = "top") +
#   labs(x = "Generation", y = "G(t)", color = "Self\nrate")
# 
# ggsave(paste("../results/by-date/", today, "/simulated_linked_selection_gt_abs_cor.pdf", sep = ""), height = 4.5, width = 14, device = cairo_pdf)

# gt over time
ggplot(subset(mean_sims, adjust == "Ne constant"), aes(x = gen, y = gt, group = SIGMA, color = as.factor(SIGMA))) +
  geom_point(size = 1.5) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  facet_grid(VA~L) +
  geom_hline(yintercept = 0, lwd = 1.1, lty = 2) +
  geom_line() +
  scale_x_continuous(breaks = seq(from = 2, to = max(mean_sims$gen), by = 2)) +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "G(t)", color = "Self\nrate")

ggsave(paste("../results/by-date/", today, "/simulated_linked_selection_gt.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)

# gt pos over time
ggplot(mean_sims, aes(x = gen, y = gt_pos, group = SIGMA, color = as.factor(SIGMA))) +
  geom_point(size = 1.5) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  facet_grid(adjust~VA+L) +
  geom_hline(yintercept = 0, lwd = 1.1, lty = 2) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "G+(t)", color = "Self\nrate")

ggsave(paste("../results/by-date/", today, "/simulated_linked_selection_gt_pos.pdf", sep = ""), height = 4.5, width = 14, device = cairo_pdf)

# sum of covariances over time
ggplot(mean_sims, aes(x = gen, y = sum_cov, group = SIGMA, color = as.factor(SIGMA))) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, lty = 2, lwd = 1.1) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  facet_grid(adjust~VA+L, labeller = label_both) +
  geom_line() +
  scale_x_continuous(breaks = seq(from = 2, to = max(mean_sims$gen), by = 2)) +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "Sum of Cov(\u0394pi, \u0394pj)", color = "Self\nrate")

ggsave(paste("../results/by-date/", today, "/simulated_linked_selection_sum_cov.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)

# sum of absolute value of covariances over time
ggplot(mean_sims, aes(x = gen, y = sum_abs_cov, group = SIGMA, color = as.factor(SIGMA))) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, lty = 2, lwd = 1.1) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  facet_grid(adjust~VA + L) +
  scale_x_continuous(breaks = seq(from = 2, to = max(mean_sims$gen), by = 2)) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "Sum of |Cov(\u0394pi, \u0394pj)|", color = "Self\nrate")

ggsave(paste("../results/by-date/", today, "/simulated_linked_selection_sum_abs_cov.pdf", sep = ""), height = 7, width = 10, device = cairo_pdf)

# sum variances over time
ggplot(mean_sims, aes(x = gen, y = sum_var, group = SIGMA, color = as.factor(SIGMA))) +
  geom_point(size = 1.5) +
  #geom_abline(intercept = 0, slope = 1/16000) +
  geom_hline(yintercept = 0, lty = 2, lwd = 1.1) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  facet_grid(adjust~VA+L, scales = "free") +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "Sum of Var(\u0394pi)", color = "Self\nrate")

ggsave(paste("../results/by-date/", today, "/simulated_linked_selection_sum_var.pdf", sep = ""), height = 4.5, width = 14, device = cairo_pdf)

# importance of fluctuating vs directional selection
ggplot(mean_sims, aes(x = gen, y = ratio, group = SIGMA, color = as.factor(SIGMA))) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0.5, lty = 2, lwd = 1.1) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  facet_grid(adjust~VA+L) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "Proportion of Cov(\u0394pi, \u0394pj) > 0", color = "Self\nrate")

ggsave(paste("../results/by-date/", today, "/simulated_linked_selection_ratio.pdf", sep = ""), height = 4.5, width = 14, device = cairo_pdf)

# absolute value of covariance
#ggplot(subset(mean_sims,gen <=10), aes(x = gen, y = sum_abs_cov/(sum_abs_cov + sum_var), group = interaction(sigma, alpha), color = as.factor(sigma))) +
ggplot(mean_sims, aes(x = gen, y = gt_abs, group = SIGMA, color = as.factor(SIGMA))) +
  geom_point(size = 1.5) +
  #geom_vline(xintercept = 5) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  facet_grid(adjust~VA+L) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "|G(t)|", color = "Self\nrate")

ggsave(paste("../results/by-date/", today, "/simulated_linked_selection_gt_abs.pdf", sep = ""), height = 4.5, width = 14, device = cairo_pdf)

# sum of negative vs positive cov
ggplot(mean_sims, aes(x = gen, y = sum_neg_cov, group = SIGMA, color = as.factor(SIGMA))) +
  geom_point(size = 1.5) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  facet_grid(adjust~VA+L) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "Sum Cov < 0", color = "Self\nrate")

ggsave(paste("../results/by-date/", today, "/simulated_linked_selection_sum_neg_cov.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)

ggplot(mean_sims, aes(x = gen, y = sum_pos_cov, group = SIGMA, color = as.factor(SIGMA))) +
  geom_point(size = 1.5) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  facet_grid(adjust~VA+L) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "Sum Cov > 0", color = "Self\nrate")

ggsave(paste("../results/by-date/", today, "/simulated_linked_selection_sum_pos_cov.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)
```

## covariance

### global summary of covariance

```{r}
if(!exists(unbiased_cov_pdiffs)){
  unbiased_cov_pdiffs <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/unbiased_increment_covariances.tsv")
  unbiased_cov_pdiffs_mat <- as.matrix(unbiased_cov_pdiffs[,-1])
  rownames(unbiased_cov_pdiffs_mat) <- unbiased_cov_pdiffs$V1
  rm(unbiased_cov_pdiffs)
}

# plot the entire matrix as a heatmap
plot_var_cov_matrix(cov2cor(unbiased_cov_pdiffs_mat), paste(results_path, "unbiased_cor_mat_heatmap.pdf"), include_values = F, include_tick_labels = F, width = 7, height = 7, tile_or_raster = "raster")

plot_var_cov_matrix(sign(unbiased_cov_pdiffs_mat) == 1, paste(results_path, "unbiased_var_cov_mat_heatmap.pdf"), include_values = F, include_tick_labels = F, width = 7, height = 7, tile_or_raster = "raster", discrete_or_continuous = "discrete")

# Try doing a PCA from bias-corrected increment covariances
cor_matrix   <- cov2cor(unbiased_cov_pdiffs_mat)
eigen_decomp <- eigen(cor_matrix)

eigenvalues  <- eigen_decomp$values   # Variance of each component
eigenvectors <- eigen_decomp$vectors  # The projection matrix (V)

# 3. Standardize the raw data
# This step is critical to match the correlation-based eigenvectors
scaled_data <- scale(pdiffs, center = TRUE, scale = TRUE)

# 4. Project data into PCA space using matrix multiplication (%*%)
pca_scores <- scaled_data %*% eigenvectors

# Optional: Add column names for clarity
colnames(pca_scores) <- paste0("PC", 1:ncol(pca_scores))

# plot pca, randomize plotting order
ggplot(pca_scores, aes(x = PC1, y = PC2)) +
  #geom_point()
geom_density_2d_filled(alpha = 0.8) +
theme(legend.position = "none") # Removes massive color legend

```

### covariances by site

```{r}

if(!exists(unbiased_cov_pdiffs)){
  unbiased_cov_pdiffs <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/unbiased_increment_covariances.tsv")
}

for(site in sites){
  
  site_pattern <- paste("^", site, "_", sep = "")
  
  melt_cov <- reshape::melt(unbiased_cov_pdiffs[grep(site_pattern, rownames(unbiased_cov_pdiffs)), grep(site_pattern, colnames(unbiased_cov_pdiffs))])
  
  melt_cov <- melt_cov[(melt_cov$X1 != melt_cov$X2),]
  
  melt_cov <- melt_cov[!duplicated(melt_cov$value),]
  
  melt_cov$tmin <- pmin(as.numeric(gsub("_", "", str_extract(as.character(melt_cov$X1), "_.*_"))),as.numeric(gsub("_", "", str_extract(as.character(melt_cov$X2), "_.*_"))))
  
  melt_cov$tmax <- pmax(as.numeric(gsub("_", "", str_extract(as.character(melt_cov$X1), "_.*_"))),as.numeric(gsub("_", "", str_extract(as.character(melt_cov$X2), "_.*_"))))
  
  melt_cov <- melt_cov[gsub(".*_", "", melt_cov$X1) == gsub(".*_", "", melt_cov$X2),]
  
  melt_cov$rep <- gsub(".*_", "", melt_cov$X1)
  
  ggplot(melt_cov, aes(x = as.factor(tmax), y = value, group = interaction(tmin, rep), color = as.factor(tmin))) +
    geom_point() +
    scale_color_scico_d(palette = "turku", begin = 0.1, end = 0.9) +
    geom_line() +
    theme(legend.position = "top") +
    labs(x = "t", y = "Cov(\u0394ps, \u0394pt)", color = "s", title = paste("Site #", site, sep = ""))
  
  ggsave(paste(results_path, "genome_wide_covariances_site_", site, ".pdf", sep = ""), device = cairo_pdf)
}

```

## G(t): linked selection

```{r}
# permutation-based null distributions
# gabs_null_1 <- fread("../results/2026-05-02/gt_abs_permutation_intervals.txt")
# gabs_null_2 <- fread("../results/2026-05-02/gt_abs_permutation_intervals_2.txt")
# gabs_null <- rbind(gabs_null_1, gabs_null_2)

gabs_null <- fread("../results/2026-05-02/gt_abs_permutation_intervals_200.txt")

# bootstrapping-based confidence intervals
gabs_cis <- read.table("../results/by-date/2026-04-15/gt_abs_bootstrap_intervals.txt")

gabs_cis <- merge(gabs_cis, worldclim_sitesdata, by = "site")
  
ggplot(gabs_cis, aes(x = gen, y = med, group = site, color = bio1)) +
  geom_point() +
  geom_line() +
  geom_errorbar(aes(ymin=lwr, ymax=upr), width=.2) +
  geom_ribbon(data = gabs_null, mapping = aes(x=gen, ymin=lwr, ymax=upr), alpha = 0.2, color = "grey", size = 0.01) + 
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:5)) +
  labs(x = "Generation", y = expression(G[abs]), color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste(results_path, "linked_selection_abs_with_null_200_perms.pdf"), width = 6, height = 4)
```

## Replicate G(t)

```{r}


```

## Ne and VA(1)

```{r}

if(!exists("cis")){
  cis <- fread("../results/by-date/2026-05-31/VA1_Ne_estimates_200_boots_abs.txt")
}

# convert to Ne
avg_boot_ne <- cis %>%
  subset(term == "b") %>%
  mutate(Ne_upr=1/(2*lwr),
         Ne_med=1/(2*med),
         Ne_lwr=1/(2*upr)) %>%
  group_by(site) %>%
  filter((Ne_lwr > 1) &
                (Ne_upr > 1) & 
                (Ne_med > 1)) %>%
  mutate(site_Ne_med = median(Ne_med)) %>%
  arrange(desc(site_Ne_med))

#avg_boot_ne$Ne_upr[(avg_boot_ne$Ne_upr < 0)] <- 1e5 # if upper bound is uninterpretable, just give it a large number
#dodge_width <- position_jitter(width = 0.1, height = 0)
avg_boot_ne$site <- factor(avg_boot_ne$site, levels = unique(avg_boot_ne$site))

ggplot(avg_boot_ne,
       aes(
         x = as.factor(rep),
         y = Ne_med,
         ymin = Ne_lwr,
         ymax = Ne_upr
       )) +
  geom_pointrange(size = 0.1) +
  geom_hline(aes(yintercept = site_Ne_med), avg_boot_ne, lty = 2, color = "red") +
  facet_wrap(. ~ site, nrow = 1, scales = "free_x") +
  scale_y_log10() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  labs(y = "Ne", x = "Replicate", title = "Site")

ggsave(paste(results_path, "Ne_estimates_200_boots_abs.pdf", sep = ""), width = 11, height = 3)

# plot Va(1) estimates, set negative Va(1) estimates to zero

#cis[(cis < 0)] <- 0 # if upper bound is uninterpretable, just give it a large number

avg_boot_va1 <- cis %>%
  subset(term == "a") %>%
  group_by(site) %>%
  filter((lwr > 0) &
                (upr > 0) &
                (med > 0)) %>%
  mutate(site_VA1_med = median(med)) %>%
  arrange(desc(site_VA1_med))

avg_boot_va1$site <- factor(avg_boot_va1$site, levels = levels(avg_boot_ne$site))

ggplot(avg_boot_va1,
       aes(
         x = as.factor(rep),
         y = med,
         ymin = lwr,
         ymax = upr
       )) +
  geom_pointrange(size = 0.1) +
  geom_hline(aes(yintercept = site_VA1_med),
             avg_boot_va1,
             lty = 2,
             color = "red") +
  facet_wrap(. ~ site, nrow = 1, scales = "free_x") +
  scale_y_log10() +
  theme(
    axis.title.x = element_blank(),
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.line.x = element_blank(),
    panel.grid.major.x = element_blank()
  ) +
  labs(y = "VA(1)", title = "Site")

ggsave(paste(results_path, "VA1_estimates_200_boots_abs.pdf", sep = ""), width = 11, height = 3)

# plot correlation between Ne and VA1
ne_vs_va1 <- merge(avg_boot_ne[,c("site", "rep", "Ne_med")], avg_boot_va1[,c("site", "rep", "med")], by = c("site", "rep"))

ggplot(ne_vs_va1, aes(x = Ne_med, y = med)) +
  geom_point() +
  geom_smooth(method = "lm") +
  scale_x_log10() +
  scale_y_log10() +
  labs(x = "Median Ne estimate", y = "Median VA(1) estimate")

```

## G' 

```{r}
gp <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-06-02/gprime_point_esitmates.txt")

gp <- gp %>% 
  group_by(site, gen) %>% 
  summarize(gp = median(gp_values))

ggplot(gp, aes(x = gen, y = gp, group = site)) +
  geom_point() +
  geom_line() +
  facet_zoom(ylim = c(0,1), show.area = T) +
  labs(x = "Generation", y = "Median G' across replicates")

  ggsave(paste(results_path, "median_g_prime_across_reps.pdf", sep = ""), device = cairo_pdf, width = 8, height = 5)
```

## linked selection vs Ne, VA(1)

Hypothesis: A rapid increase in Gabs(t) should correspond to a high VA(1). Also, a low Ne should correspond to a low repeatability.

```{r}

va_cis <- read.table("../results/by-date/2026-05-19/Va1_estimates_200_boots.txt", header = T)

gabs_cis$term <- "va"

gabs_va <- merge(gabs_cis, va_cis, by = "site")

ggplot(gabs_va %>% filter(gen == 2), aes(x = med.x, y = med.y)) +
  geom_point()

```

## Convergence correlation

### between environments, first generation, TO-DO: CORRELATE WITH CLIMATE

```{r}
conv_cor_cis <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-05-10/convergence_correlation_bootstrap_intervals.txt")

conv_cor_cis_copy <- conv_cor_cis
names(conv_cor_cis_copy)[1:2] <-c("site_b", "site_a")

conv_cor_cis_diagonal <- data.frame(
  site_a = unique(c(conv_cor_cis$site_a, conv_cor_cis$site_b)),
  site_b = unique(c(conv_cor_cis$site_a, conv_cor_cis$site_b)),
  lwr = 1,
  med = 1,
  upr = 1
)

conv_cor_cis_mat <- rbind(conv_cor_cis, conv_cor_cis_copy, conv_cor_cis_diagonal) %>%
  dplyr::select(site_a, site_b, med) %>%
  pivot_wider(names_from = site_a, values_from = med) %>%
  arrange(site_b)  %>%
  as.data.frame()

rownames(conv_cor_cis_mat) <- conv_cor_cis_mat$site_b

conv_cor_cis_mat <- conv_cor_cis_mat[,-1]

# create environment distance matrix
env_dist <- worldclim_sitesdata %>%
  mutate(id = row_number()) %>%
  cross_join(., .) %>%
  #filter(site.x > site.y) %>% # Key step: keeps only one version of each pair
  mutate(dist = abs(bio1.x - bio1.y))

env_dist_mat <- env_dist %>%
  dplyr::select(site.x, site.y, dist) %>%
  pivot_wider(names_from = site.x, values_from = dist) %>%
  as.data.frame()

rownames(env_dist_mat) <- env_dist_mat$site.y

env_dist_mat <- env_dist_mat[,-1]

# do mantel test
# hyp: convergence correlation decays with temperature
env_dist_bool <- rownames(env_dist_mat) %in% rownames(conv_cor_cis_mat)

env_dist_mat <- env_dist_mat[env_dist_bool,env_dist_bool]

ecodist::mantel(as.dist(env_dist_mat) ~ as.dist(conv_cor_cis_mat))

# what about all the environmental variables?
worldclim_subset <- worldclim_sitesdata[worldclim_sitesdata$site %in% unique(c(conv_cor_cis$site_a, conv_cor_cis$site_b)),]

ecodist::mantel(as.dist(conv_cor_cis_mat) ~ dist(worldclim_subset[,-1]))

# conv_cor_cis$site_a <- factor(conv_cor_cis$site_a, levels = worldclim_sitesdata$site[hclust(dist(worldclim_sitesdata[,-1]))$order])
# 
# conv_cor_cis$site_b <- factor(conv_cor_cis$site_b, levels = worldclim_sitesdata$site[hclust(dist(worldclim_sitesdata[,-1]))$order])

# convergence correaltion heatmap
plot_pairs_heatmap(
  conv_cor_cis,
  as.factor(site_a),
  as.factor(site_b),
  med,
  T,
  lwr,
  upr,
  "cork",
  width = 8,
  height = 8,
  legend_name = "Convergence\ncorrelation",
  output_name = paste(
    "../results/by-date/",
    today,
    "/convergence_correlations.pdf",
    sep = ""
  )
)

# temp diff heat map
plot_pairs_heatmap(
  env_dist,
  as.factor(site.x),
  as.factor(site.y),
  dist,
  F,
  scico_palette = "bamako",
  midpoint = median(env_dist$dist),
  direction = -1,
  height = 8,
  width = 8,
  legend_name = "|bio1(x) - bio1(y)|",
  output_name = paste(
    "../results/by-date/",
    today,
    "/pairwise_temp_diff.pdf",
    sep = ""
  )
)

# plotting convergence vs temp difference
env_dist$site_pair <- paste(env_dist$site.y, "_", env_dist$site.x, sep = "")
conv_cor_cis$site_pair <- paste(conv_cor_cis$site_a, "_", conv_cor_cis$site_b, sep = "")

conv_cor_vs_env_dist <- merge(conv_cor_cis, env_dist, by = "site_pair")

ggplot(conv_cor_vs_env_dist, aes(x = dist, y = med)) +
  geom_smooth(method = "lm", color = "red") +
  #stat_cor(method = "pearson", label.x.npc = "middle") +
  geom_point(shape = 16) +
  labs(x = "|bio1(site a) - bio1(site b)|", y = "Median convergence correlation\n(1000 bootstrap samples)")

ggsave(paste(results_path, "convergence_correlation_vs_environment_distance.pdf"), width = 6, height = 5)
```

### within environments, first generation

```{r}
conv_cor_cis <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-05-28/convergence_correlation_wn_env_bootstrap_intervals.txt")

ggplot(conv_cor_cis, aes(x = reorder(site, -med), y = med)) +
  # 1. Draw the range lines first (colored grey)
  geom_linerange(aes(ymin = lwr, ymax = upr), color = "darkgrey", linewidth = 1) +
  # 2. Draw the center points on top (colored red)
  geom_point(color = "red", size = 3, pch = 1) +
  labs(x = "Site", y = "Convergence correlation\nin 1st increment")


ggsave(paste(results_path, "convergence_correlation_wn_env_first_increment.pdf"), width = 6, height = 5)
```

## Conceptual images

```{r}

#### selection, drift, fluctuating selection ####
gen = 100
N = 100
h = 0.5
s = 0.05
q0 = 0.5

selection = WF.sel(N, q0, h, rep(s, times = gen), gen)

drift = WF.sel(N, q0, h, rep(0,times = gen), gen)

fluct <- WF.sel(N, q0, h, c(rep(s, times = gen/2), rep(-3*s, times = gen/2)), gen)

plotdata <- data.frame(
  gen = 0:gen,
  selection = c(q0*N, selection)/N,
  drift = c(q0*N, drift)/N,
  fluct = c(q0*N, fluct)/N
)

plotdata <- melt(plotdata, id = "gen")

ggplot(plotdata, aes(x = gen, y = value)) +
  geom_abline(slope = 0, intercept = q0, lty = 2) +
  geom_point() +
  geom_line(lwd = 1.1) +
  facet_wrap(~variable, labeller = as_labeller(c("selection" = "Directional\nselection", "drift" = "Drift", "fluct" = "Fluctuating\nselection"))) +
  labs(x = "Generation", y = "Allele frequency")

ggsave(paste("../results/by-date/", today, "/allele_frequency_changes.pdf", sep = ""), height = 4, width = 8)

#### Replicate covariances ####
set.seed(123)

plotdata <- data.frame(
  gen = 0:gen,
  selection_1 = c(q0, WF.sel(N, q0, h, rep(s, times = gen), gen)/N),
  selection_2 = c(q0, WF.sel(N, q0, h, rep(s, times = gen), gen)/N),
  drift_1 = c(q0, WF.sel(N, q0, h, rep(0, times = gen), gen)/N),
  drift_2 = c(q0, WF.sel(N, q0, h, rep(0, times = gen), gen)/N)
)

plotdata <- reshape2::melt(plotdata, id = "gen")

plotdata$mech <- gsub("_.*", "", plotdata$variable)
plotdata$rep <- gsub(".*_", "", plotdata$variable)

ggplot(plotdata, aes(x = gen, y = value, group = rep)) +
  geom_abline(slope = 0, intercept = q0, lty = 2) +
  geom_point() +
  geom_line(lwd = 1.1) +
  facet_wrap(~mech, labeller = as_labeller(c("selection" = "Directional\nselection", "drift" = "Drift"))) +
  labs(x = "Generation", y = "Allele frequency")

ggsave(paste("../results/by-date/", today, "/replicate_covariances.pdf", sep = ""), height = 4, width = 8)

#### Many replicates of drift ####

plotdata2 <- data.frame(
  gen = 0:gen,
  n1 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n2 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n3 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n4 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n5 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n6 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n7 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n8 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n9 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n10=  c(q0, WF.sel(N, q0, h, 0, gen)/N)
)

plotdata2 <- melt(plotdata2, id = "gen")

ggplot(plotdata2, aes(x = gen, y = value, group = variable)) +
  #geom_abline(slope = 0, intercept = q0, lty = 2) +
  geom_line(lwd = 2, color = "green") +
  theme_void() +
  labs(x = "Generation", y = "Allele frequency")

ggsave(paste("../results/", today, "/neutral_examples.pdf", sep = ""), height = 4, width = 7)


#### allele frequency vs environmental increments ####
dT_vs_dp <- data.frame(
  t = 1:3,
  p = c(0.5, 0.7, 0.75, 0.5, 0.4, 0.55),
  temp = c(20, 21, 21.5, 20, 19, 20),
  s = c(rep("directional", times = 3), rep("fluctuating", times = 3))
)

p1 <- ggplot(dT_vs_dp, aes(x = t, y = p)) +
  geom_point() +
  geom_line() +
  facet_wrap(~s) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank()
  ) +
  labs(y = "Allele frequency") +
  scale_x_continuous(breaks = seq(1, 3, by = 1))

p2 <- ggplot(dT_vs_dp, aes(x = t, y = temp)) +
  geom_point() +
  geom_line() +
  facet_wrap(~s) +
  theme(
    strip.background = element_blank(),
    strip.text = element_blank()
  ) +
  labs(y = "Temperature", x = "Generation") +
  scale_x_continuous(breaks = seq(1, 3, by = 1))

ggarrange(p1, p2, nrow = 2, ncol = 1)

ggsave(paste("../results/", today, "/example_environment_driver.pdf", sep = ""), height = 7, width = 7)

# More cartoons of directional vs fluctuating vs drift
gen <- 0:5

dat <- data.frame(
  generation = rep(gen, 6),
  frequency = c(
    # High fitness (selected, increasing) - blue
    0.50, 0.58, 0.67, 0.75, 0.82, 0.88,
    # Low fitness (selected, decreasing) - red
    0.50, 0.42, 0.35, 0.28, 0.22, 0.17,
    # Fluctuating selection (up-down-up pattern) - purple
    0.50, 0.65, 0.42, 0.68, 0.38, 0.62,
    # Drift (3 replicates, simulated random walk)
    0.50, 0.53, 0.56, 0.54, 0.57, 0.55,
    0.50, 0.47, 0.44, 0.46, 0.43, 0.41,
    0.50, 0.52, 0.48, 0.50, 0.53, 0.51
  ),
  type = rep(c("Positive selection", "Negative selection",
               "Fluctuating selection", "Drift", "Drift", "Drift"), each = 6),
  replicate = rep(c(1, 1, 1, 1, 2, 3), each = 6)
)

dat$type <- factor(dat$type, levels = c("Positive selection", "Negative selection",
                                         "Fluctuating selection", "Drift"))
dat$group <- paste(dat$type, dat$replicate)

cols <- c("Positive selection" = "#4A90D9",
          "Negative selection" = "#D94A4A",
          "Fluctuating selection" = "#9B59B6",
          "Drift" = "grey55")

p <- ggplot(dat, aes(x = generation, y = frequency, colour = type, group = group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_colour_manual(values = cols, name = NULL) +
  scale_x_continuous(breaks = 0:5, limits = c(0, 5)) +
  scale_y_continuous(limits = c(0.1, 1.0), breaks = seq(0.1, 1.0, 0.1)) +
  labs(x = "Generation", y = "Frequency") +
  theme_classic(base_size = 14) +
  theme(
    legend.position = c(0.22, 0.92),
    legend.background = element_blank(),
    legend.text = element_text(size = 11),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )

p
```

## seed mix: compare freqk estimated allele frequencies to vcf snp calls

```{r}
# greneNet allele counts
real_calls <- read.table("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/resources/greneNet_allele_counts.tsv", sep = "\t", header = F, col.names = c("CHROM", "POS", "REF", "ALT", "AN", "AC"))

# calculate reference allele frequency
real_calls$RF_real = 1 - real_calls$AC/real_calls$AN

# loop over seed mixes
plots_freqk_seedmix <- list()
for(i in paste("S", 1:8, sep = "")){
  cat(i)
  # freqk estimates
  #freqk_calls <- read.table("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/resources/calls.txt", sep = "|", header = F, col.names = c("RF_freqk", "AF_freqk"))
  freqk_calls <- read.table(paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/freqk_call_results/", i, ".txt", sep = ""), sep = "|", header = F, col.names = c("RF_freqk", "AF_freqk"))
  
  freqk_index <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/resources/ref_index_greneNet.txt", sep = ",", header = F, data.table = F, select = c(2,3))
  
  freqk_calls <- cbind(freqk_index, freqk_calls)
  
  freqk_calls$V3 = freqk_calls$V3 + 2
  
  # merge
  names(freqk_calls)[1:2] = c("CHROM", "POS")
  
  real_vs_freqk <- merge(real_calls, freqk_calls, by = c("CHROM", "POS"))
  
  # plot
  cor_estimate = signif(cor.test(real_vs_freqk$RF_real, real_vs_freqk$RF_freqk, method = "pearson")$estimate, 2)
  
  plots_freqk_seedmix[[i]] <- ggplot(real_vs_freqk, aes(x = RF_real, y = RF_freqk)) +
    geom_bin2d() +
    geom_abline(slope = 1, intercept = 0, color = "black", lwd = 2) +
    scale_fill_scico(palette = "bilbao", trans = "log10", direction = -1) +
    labs(x = "Called REF frequency", y = "freqk estimate", fill = "log10(count)", title = paste(i, ": r = ", cor_estimate, sep = ""))
  
}

ggarrange(plotlist=plots_freqk_seedmix, ncol = 2, nrow = 4, common.legend = T)

ggsave(paste("../results/", today, "/bin2d_greneNet_vs_freqk.pdf", sep = ""), height = 11, width = 6, device = cairo_pdf)
```


# Pool-seq power analysis simulations

```{r}

# Parameters ===========================================
a = 0.05 # confidence 
r = 12*4*3 # number of replicates (replicates per location, per treatment)
t = 60 # number of time points (2 generations per year, for 3 years)
p0 = 0.125 # minimum minor allele frequency
N = 100000 # number of diploids in population
s = 0.001 # selection coefficient
n = 250 # average pool size in number of individuals
d = 10000 # high depth, because of reconstruction
h = 0.5

r*t*n # want to target 100,000 survivors

# Simulation loop ======================================
all_mod_res <- NULL

num_sims <- 1000

for(i in 1:num_sims){
  
  # sampling size variation over time
  sampling_sizes <- 2*rpois(t, n)
  
  # allele frequency trajectory
  true_freq_traj <- reshape2::melt(replicate(r, WF.sel(2*N, p0, h, rep(s, times = t), t)))
  
  # assuming constant depth across time points
  obs_freq_traj <- mapply(simulate_af_est, true_freq_traj$value, sampling_sizes, MoreArgs = list(d = d))
  
  # create dataframe of observations
  frame_freq_traj <- true_freq_traj
  frame_freq_traj$p_obs <- obs_freq_traj
  frame_freq_traj$n <- sampling_sizes

  # fit linear model                     
  mod_lm <- lm(log(p_obs/(1-p_obs)) ~ Var1, data = frame_freq_traj)
  
  # create dataframe of observations
  mod_res <- tidy(mod_lm) %>% filter(term == "Var1")
  all_mod_res <- rbind(all_mod_res, mod_res)
}

# calculate power: number of significant slopes detected
all_mod_res$detect = (all_mod_res$p.value < a)
all_mod_res %>% 
  summarize(sum(detect)) / num_sims

# Power of about 0.77 (80 %) with these parameters:
# a = 0.05
# r = 32
# t = 3
# p0 = 0.2
# N = 1000
# s = 0.03
# n = 1000
# d = 10000

# Power of about 0.9 with these parameters:
# a = 0.05
# r = 144
# t = 6 # 2 generations per year
# p0 = 0.1
# N = 1000
# s = 0.03
# n = 200
# d = 10000

ggplot(true_freq_traj, aes(x = Var1, y = value, group = Var2)) +
  geom_point() + 
  geom_line() +
  geom_hline(yintercept = p0)

true_freq_traj %>% group_by(Var1) %>% summarize(p = mean(value))
```
## claude attempt

```{r}

 
## ---- core functions --------------------------------------------------------
 
#' Allele frequency after one generation of selection (diploid, q = selected allele).
#' AA = 1, Aa = 1 + h*s, aa = 1 + s.
fitfreq <- function(q, h, s) {
  p   <- 1 - q
  num <- q^2 * (1 + s) + p * q * (1 + h * s)
  den <- 1 + s * q * (2 * h * p + q)
  qp  <- num / den
  pmin(pmax(qp, 0), 1)
}
 
#' Wright-Fisher trajectory under selection.
#' N here is the number of GENE COPIES (chromosomes). Pass 2 * N_diploid.
#' s, h may be scalars or length-(G-1) vectors. Returns a length-G frequency vector.
WF.sel <- function(N, q, h, s, G) {
  if (length(s) == 1) s <- rep(s, G - 1)
  if (length(h) == 1) h <- rep(h, G - 1)
  cnt    <- numeric(G)
  cnt[1] <- round(N * q)                 # start exactly at q (gen 0)
  for (i in 2:G) {
    p_sel  <- fitfreq(cnt[i - 1] / N, h[i - 1], s[i - 1])
    cnt[i] <- rbinom(1, N, p_sel)
  }
  cnt / N
}
 
#' Pool-seq allele-frequency estimate (Buffalo & Coop 2020 PNAS structure).
#' p = true frequency; n = pool size in INDIVIDUALS (2*n chromosomes sampled);
#' d = mean read depth. One consistent realized depth d_obs is used throughout.
simulate_af_est <- function(p, n, d) {
  d_obs <- rpois(1, d)
  if (d_obs == 0) return(NA_real_)
  chroms <- 2 * n                                  # diploid pool
  x_obs  <- rbinom(1, size = chroms, prob = p)     # allele count in the pool
  r_obs  <- rbinom(1, size = d_obs, prob = x_obs / chroms)  # reads carrying allele
  (r_obs + 0.5) / (d_obs + 1)                      # continuity correction -> (0,1)
}
 
## ---- parameters ------------------------------------------------------------
 
a       <- 0.05          # significance threshold
r       <- 12 * 4 * 3    # number of replicate cages
t       <- 6             # number of sampled time points (generations 0 .. t-1)
p0      <- 0.125         # starting allele frequency
N_dip   <- 1000          # diploid population size
s       <- 0.02          # selection coefficient
h       <- 0.5           # dominance
n       <- 250           # pool size in INDIVIDUALS per sample
d       <- 10000         # mean read depth
one_sided <- FALSE        # directional test (we expect s > 0)
 
num_sims <- 1000
 
## ---- simulation loop -------------------------------------------------------
 
set.seed(1)
all_mod_res <- vector("list", num_sims)
 
for (i in seq_len(num_sims)) {
 
  ## allele frequency trajectories: t x r matrix -> long (Var1 = gen, Var2 = rep)
  traj <- replicate(r, WF.sel(2 * N_dip, p0, h, rep(s, t), t))
  df   <- reshape2::melt(traj)              # columns: Var1, Var2, value
  names(df)[names(df) == "value"] <- "p_true"
 
  ## per-(generation, replicate) pool sizes; mean n, Poisson variation
  df$n_pool <- rpois(nrow(df), n)
 
  ## observed pool-seq frequency estimate, one realized depth each
  df$p_obs <- mapply(simulate_af_est, df$p_true, df$n_pool, MoreArgs = list(d = d))
 
  ## generation clock starts at 0 so intercept corresponds to p0
  df$gen <- df$Var1 - 1
 
  ## quasibinomial GLM on observed frequency, weighted by realized pool chromosomes.
  ## (weights approximate the sampling precision of each point.)
  df_fit <- df[!is.na(df$p_obs), ]
  mod <- glm(p_obs ~ gen, data = df_fit,
             weights = 2 * n_pool,
             family = quasibinomial(link = "logit"))
 
  res <- broom::tidy(mod) %>% filter(term == "gen")
 
  ## one-sided p-value for H1: slope > 0
  if (one_sided) {
    res$p.value <- pt(res$statistic, df = mod$df.residual, lower.tail = FALSE)
  }
 
  all_mod_res[[i]] <- res
}
 
all_mod_res <- bind_rows(all_mod_res)
all_mod_res$detect <- all_mod_res$p.value < a
 
power <- mean(all_mod_res$detect)
cat(sprintf("Estimated power: %.3f\n", power))
cat(sprintf("Mean slope estimate: %.4f (true per-gen logit signal ~ %.4f)\n",
            mean(all_mod_res$estimate), s * (h + (1 - 2 * h) * p0)))


```

# Fst 

## historical vs modern pairings

```{r}
pairings <- read.csv("../data/comb_hist+1001g.prs_geo+gen.csv")

historical <- pairings[, 1:2]

historical$type <- "historical"

colnames(historical) <- gsub("historical_", "", colnames(historical))

modern <- pairings[, 5:6]

modern$type <- "modern"

colnames(modern) <- gsub("modern_", "", colnames(modern))

slim_pairings <- rbind(historical, modern)

slim_pairings$time <- slim_pairings$year - min(slim_pairings$year) + 1

slim_pairings <- slim_pairings %>% group_by(type, time) %>% summarize(n = n()) %>% arrange(time, .by_group = T)

slim_pairings$ID = 1:nrow(slim_pairings)

write.table(slim_pairings, "../config/sampling_times.tsv", quote = F, row.names = F, sep = "\t")

sum(slim_pairings$n)

```


## Fst - historical vs modern

```{r}
cutoff = 0.05

msprime <- vroom(
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/msprime_fst.txt",
  delim = "\t",
  col_names = F
)

msprime$ID <- as.numeric(gsub(".*/|.txt", "", fst$X1))

names(msprime)[2] = "msprime_fst"

gcta <- vroom(
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/gcta_fst.txt",
  delim = "\t",
  col_names = F
)

gcta$ID <- as.numeric(gsub(".*/|.fst:1", "", gcta$X1))

gcta <- gcta %>% group_by(ID) %>% summarize(gcta_fst = mean(X7), S = n())

gcta_msprime <- merge(msprime, gcta, by = "ID")

# compare gcta and msprime
nei_weir_fst <- ggplot(gcta_msprime, aes(x = msprime_fst, y = gcta_fst)) +
  geom_point(shape = 16) +
  #geom_smooth(method = "lm") +
  geom_abline(slope = 1, intercept = 0, lwd = 1.2, color = "green") +
  labs(x = "Nei 1973 Fst (msprime)", y = "Weir 1996 Fst (gcta)")

# merge with window sizes and calculate p-values
params <- vroom(
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/config/fst_params.tsv",
  delim = "\t"
)

params_fst <- merge(params, gcta_msprime, by = "ID")

params_fst %>% 
  #filter(gcta_fst > 0) %>% 
  group_by(L) %>% 
  summarize(upb = quantile(gcta_fst, probs = 1 - cutoff/2, na.rm = T),
            lwb = quantile(gcta_fst, probs = cutoff / 2, na.rm = T),
            ost1_filter_test = 1 - ecdf(gcta_fst)(0.003892181),
            erecta_filter_test = 1 - ecdf(gcta_fst)(0.032451811),
            ost_unfilter_test = 1 - ecdf(gcta_fst)(0.011700622),
            erecta_unfilter_test = 1 - ecdf(gcta_fst)(0.014078437))

# tact distribution of fst from gcta
fst_density <- ggplot(params_fst, aes(x = gcta_fst)) +
  geom_density() +
  facet_wrap(L~., ncol = 1, labeller = "label_both") +
  geom_vline(data = filter(params_fst, L == 3081), aes(xintercept = 0.003892181), lty = 1) +
  geom_vline(data = filter(params_fst, L == 3081), aes(xintercept = 0.011700622), lty = 2) +
  geom_vline(data = filter(params_fst, L == 6583), aes(xintercept = 0.032451811), lty = 1) +
  geom_vline(data = filter(params_fst, L == 6583), aes(xintercept = 0.014078437), lty = 2) +
  labs(x = "Neutral Weir 1996 Fst\n(Sampling desing, Ne = 20,000)")


ggarrange(fst_density, nei_weir_fst, ncol = 2, labels = c("A", "B"), font.label = list(size = 16, family = "Helvetica"))

ggsave(paste("../results/", today, "/baseline_fst_historical_modern.pdf", sep = ""), height = 4, width = 8, device = cairo_pdf)
```


## load scRNA-seq data

```{r}
global <- readRDS("/global/scratch/users/milesroberts/tmp/GSE226097_global_integration_221009.rds")

meta <- global[[]]

#flower <- readRDS("../data/GSE226097_flower_250805.rds")

#flower[["data"]]
```

## check overlap between GEA and AEGIS

```{r}
lfmm_files <- list.files("lfmm.*", path = "../data", full.names = T)

lfmm <- lapply(lfmm_files, read.csv, header = T)

names(lfmm) <- lfmm_files

lfmm <- lfmm %>% 
  bind_rows(.id = "groups")

lfmm$chrom <- as.numeric(gsub("_.*", "", lfmm$snp_id))
lfmm$pos <- as.numeric(gsub(".*_", "", lfmm$snp_id))

lfmm$clim <- gsub(".*_", "", gsub("_k.*", "", lfmm$groups))

# Chr1:1185512..1188696 - CRY2
# Chr1:24331373..24333999 - FT
# Chr2:11058944..11063324 - ELF3
# Chr2:11531967..11534358 - CAM5
# Chr2:18807505..18811125 - SOC1
# Chr4:9195247..9199840 - PHYD
# Chr4:11000771..11003010 - TSF
# Chr5:3173382..3179448 - FLC
# Chr5:5170162..5173085 - CO
# Chr5:18589438..18591506 - DOG1
buffer = 10000
chroms = c(1,1,2,2,2,4,4,5,5,5)
starts = c(1185512,24331373,11058944,11531967,18807505,9195247,11000771,5170162,3173382,18589438)
ends = c(1188696,24333999,11063324,11534358,18811125,9199840,11003010,5173085,3179448,18591506)
starts = starts - buffer
ends = ends + buffer

lfmm_ft_only <- NULL
for(i in seq_along(chroms)){
  print(i)
  lfmm_ft <- lfmm %>% subset((chrom == chroms[i]) & (pos >= starts[i]) & (pos <= ends[i]) & (significant == "True"))
  lfmm_ft_only <- rbind(lfmm_ft_only, lfmm_ft)
}


head(lfmm_ft_only)
```
>>>>>>> Stashed changes
