---
title: "Linked selection in GrENE-net phase 2"
output: 
  html_notebook: 
    toc: true
---

# set-up

```{r}
rm(list = ls())
gc()
.rs.restartR()
```

# packages

```{r}
library(ggplot2)

library(scico)

library(reshape2)

library(tidyverse)

library(vroom)

library(broom)

library(data.table)

box::use(./R/gt[...])
box::use(./R/plot[...])
box::use(./R/utils[...])

today <- Sys.Date()

results_path <- paste("../results/", today, "/", sep = "")

# global ggplot2 theme
theme_set(
  theme_minimal() +
    theme(
      text = element_text(
        size = 16,
        family = "Helvetica",
        colour = "black"
      ),
      axis.text = element_text(colour = "black"),
      plot.title = element_text(hjust = 0.5),
      strip.text = element_text(colour = "black")
    )
)
```

# functions

```{r}
# All functions moved to box modules (R/gt.R, R/plot.R, R/utils.R)
```

# Conceptual images

```{r}
gen = 100
N = 100
h = 0.5
s = 0.05
q0 = 0.5

selection = WF.sel(N, q0, h, rep(s, times = gen), gen)

drift = WF.sel(N, q0, h, rep(0,times = gen), gen)

fluct <- WF.sel(N, q0, h, c(rep(s, times = gen/2), rep(-3*s, times = gen/2)), gen)

plotdata <- data.frame(
  gen = 0:gen,
  selection = c(q0*N, selection)/N,
  drift = c(q0*N, drift)/N,
  fluct = c(q0*N, fluct)/N
)

plotdata <- melt(plotdata, id = "gen")

ggplot(plotdata, aes(x = gen, y = value)) +
  geom_abline(slope = 0, intercept = q0, lty = 2) +
  geom_point() +
  geom_line(lwd = 1.1) +
  facet_wrap(~variable, labeller = as_labeller(c("selection" = "Directional\nselection", "drift" = "Drift", "fluct" = "Fluctuating\nselection"))) +
  labs(x = "Generation", y = "Allele frequency")

ggsave(paste("../results/", today, "/allele_frequency_changes.pdf", sep = ""), height = 4, width = 8)

plotdata2 <- data.frame(
  gen = 0:gen,
  n1 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n2 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n3 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n4 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n5 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n6 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n7 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n8 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n9 =  c(q0, WF.sel(N, q0, h, 0, gen)/N),
  n10=  c(q0, WF.sel(N, q0, h, 0, gen)/N)
)

plotdata2 <- melt(plotdata2, id = "gen")

ggplot(plotdata2, aes(x = gen, y = value, group = variable)) +
  #geom_abline(slope = 0, intercept = q0, lty = 2) +
  geom_line(lwd = 2, color = "green") +
  theme_void() +
  labs(x = "Generation", y = "Allele frequency")

ggsave(paste("../results/", today, "/neutral_examples.pdf", sep = ""), height = 4, width = 7)


# matching allele frequency increments with environmental increments
dT_vs_dp <- data.frame(
  t = 1:3,
  p = c(0.5, 0.7, 0.75, 0.5, 0.4, 0.55),
  temp = c(20, 21, 21.5, 20, 19, 20),
  s = c(rep("directional", times = 3), rep("fluctuating", times = 3))
)

p1 <- ggplot(dT_vs_dp, aes(x = t, y = p)) +
  geom_point() +
  geom_line() +
  facet_wrap(~s) +
  theme(
    axis.text.x = element_blank(),
    axis.ticks.x = element_blank(),
    axis.title.x = element_blank()
  ) +
  labs(y = "Allele frequency") +
  scale_x_continuous(breaks = seq(1, 3, by = 1))

p2 <- ggplot(dT_vs_dp, aes(x = t, y = temp)) +
  geom_point() +
  geom_line() +
  facet_wrap(~s) +
  theme(
    strip.background = element_blank(),
    strip.text = element_blank()
  ) +
  labs(y = "Temperature", x = "Generation") +
  scale_x_continuous(breaks = seq(1, 3, by = 1))

ggarrange(p1, p2, nrow = 2, ncol = 1)

ggsave(paste("../results/", today, "/example_environment_driver.pdf", sep = ""), height = 7, width = 7)

# More cartoons of directional vs fluctuating vs drift
gen <- 0:5

dat <- data.frame(
  generation = rep(gen, 6),
  frequency = c(
    # High fitness (selected, increasing) - blue
    0.50, 0.58, 0.67, 0.75, 0.82, 0.88,
    # Low fitness (selected, decreasing) - red
    0.50, 0.42, 0.35, 0.28, 0.22, 0.17,
    # Fluctuating selection (up-down-up pattern) - purple
    0.50, 0.65, 0.42, 0.68, 0.38, 0.62,
    # Drift (3 replicates, simulated random walk)
    0.50, 0.53, 0.56, 0.54, 0.57, 0.55,
    0.50, 0.47, 0.44, 0.46, 0.43, 0.41,
    0.50, 0.52, 0.48, 0.50, 0.53, 0.51
  ),
  type = rep(c("Positive selection", "Negative selection",
               "Fluctuating selection", "Drift", "Drift", "Drift"), each = 6),
  replicate = rep(c(1, 1, 1, 1, 2, 3), each = 6)
)

dat$type <- factor(dat$type, levels = c("Positive selection", "Negative selection",
                                         "Fluctuating selection", "Drift"))
dat$group <- paste(dat$type, dat$replicate)

cols <- c("Positive selection" = "#4A90D9",
          "Negative selection" = "#D94A4A",
          "Fluctuating selection" = "#9B59B6",
          "Drift" = "grey55")

p <- ggplot(dat, aes(x = generation, y = frequency, colour = type, group = group)) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_colour_manual(values = cols, name = NULL) +
  scale_x_continuous(breaks = 0:5, limits = c(0, 5)) +
  scale_y_continuous(limits = c(0.1, 1.0), breaks = seq(0.1, 1.0, 0.1)) +
  labs(x = "Generation", y = "Frequency") +
  theme_classic(base_size = 14) +
  theme(
    legend.position = c(0.22, 0.92),
    legend.background = element_blank(),
    legend.text = element_text(size = 11),
    axis.title = element_text(size = 14),
    axis.text = element_text(size = 12)
  )

p
```

## Manhattan plots of sweep statistics in founders

```{r}

sweeps <- read.table("../data/sweeps.txt", header = T, sep = "\t")

sweeps <- melt(sweeps, id = c("Contig", "Var_Start", "Var_Stop", "Pos_Start", "Pos_Stop"))

sweeps <- sweeps[!(sweeps$variable %in% c("Theta_W", "Theta_Pi")),]

# manhattan plots of 
sweeps$variable <- gsub("_", " ", sweeps$variable)

sweeps$Contig <- as.factor(sweeps$Contig)

sweeps$Contig <- revalue(sweeps$Contig, c("0" = "1", "1" = "2", "2" = "3", "3" = "4", "4" = "5"))

ggplot(sweeps, aes(x = (Pos_Start + Pos_Stop)/2, y = value)) +
  geom_point(shape = 16) +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  facet_wrap(variable~Contig, ncol = 5, nrow = 5, scales = "free") +
  labs(x = "Position", y = "Statistic")

ggsave(paste("../results/", today, "/sweep_stats.pdf", sep = ""), height = 11, width = 13)
```

## seed mix: compare freqk estimated allele frequencies to vcf snp calls

```{r}
# greneNet allele counts
real_calls <- read.table("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/resources/greneNet_allele_counts.tsv", sep = "\t", header = F, col.names = c("CHROM", "POS", "REF", "ALT", "AN", "AC"))

# calculate reference allele frequency
real_calls$RF_real = 1 - real_calls$AC/real_calls$AN

# loop over seed mixes
plots_freqk_seedmix <- list()
for(i in paste("S", 1:8, sep = "")){
  cat(i)
  # freqk estimates
  #freqk_calls <- read.table("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/resources/calls.txt", sep = "|", header = F, col.names = c("RF_freqk", "AF_freqk"))
  freqk_calls <- read.table(paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/freqk_call_results/", i, ".txt", sep = ""), sep = "|", header = F, col.names = c("RF_freqk", "AF_freqk"))
  
  freqk_index <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/resources/ref_index_greneNet.txt", sep = ",", header = F, data.table = F, select = c(2,3))
  
  freqk_calls <- cbind(freqk_index, freqk_calls)
  
  freqk_calls$V3 = freqk_calls$V3 + 2
  
  # merge
  names(freqk_calls)[1:2] = c("CHROM", "POS")
  
  real_vs_freqk <- merge(real_calls, freqk_calls, by = c("CHROM", "POS"))
  
  # plot
  cor_estimate = signif(cor.test(real_vs_freqk$RF_real, real_vs_freqk$RF_freqk, method = "pearson")$estimate, 2)
  
  plots_freqk_seedmix[[i]] <- ggplot(real_vs_freqk, aes(x = RF_real, y = RF_freqk)) +
    geom_bin2d() +
    geom_abline(slope = 1, intercept = 0, color = "black", lwd = 2) +
    scale_fill_scico(palette = "bilbao", trans = "log10", direction = -1) +
    labs(x = "Called REF frequency", y = "freqk estimate", fill = "log10(count)", title = paste(i, ": r = ", cor_estimate, sep = ""))
  
}

ggarrange(plotlist=plots_freqk_seedmix, ncol = 2, nrow = 4, common.legend = T)

ggsave(paste("../results/", today, "/bin2d_greneNet_vs_freqk.pdf", sep = ""), height = 11, width = 6, device = cairo_pdf)
```

# Prepare allele frequencies

## randomly switch tracked allele frequencies

```{r}
set.seed(1111)

# LD-pruned data
#gen0 <- read.table("../data/average_seedmix_p0_LDpruned.txt", sep = "\t")
#hapfire <- read.table("../data/merged_hapFIRE_allele_frequency_LDpruned.txt", sep = "\t", header = T)

# full dataset
gen0 <- vroom("/global/scratch/users/milesroberts/tmp/average_seedmix_p0.txt", delim = "\t", col_names = F)

# randomly swith which allele is being tracked
# right now, the non-reference allele is always the one we're tracking
idx <- sample(c(T,F), size = nrow(gen0), replace = T)
gen0[idx,"X3"] <- 1 - gen0[idx,"X3"]

write.csv(gen0, "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/seedmix_hapfire_allele_frequency_rng_switched.csv", quote = F, row.names = F)

#hapfire_phase1 <- read_csv("/global/scratch/users/milesroberts/tmp/merged_hapFIRE_allele_frequency.csv")

# start <- Sys.time()
# hapfire_phase1 <- vroom("/global/scratch/users/milesroberts/tmp/merged_hapFIRE_allele_frequency.csv", delim = ",")
# end <- Sys.time()
# end - start
# 
# hapfire_phase2 <- vroom("/global/scratch/users/milesroberts/tmp/PHASE2_merged_hapFIRE_allele_frequency.csv", delim = ",")
# 
# hapfire <- cbind(hapfire_phase1, hapfire_phase2)
# 
# write.csv(hapfire, "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_2_hapfire_allele_frequency.csv", quote = F, row.names = F)

# randomly switch which allele is being tracked, same as gen0
# j = 0
# for(i in idx){
#   if(j %% 1000 == 0){
#     print(j)
#   }
#   hapfire[idx,] <- 1 - hapfire[idx,]
#   j = j + 1
# }

process_one_minus <- function(idx, chunk_size, output) {
  function(x, pos) {
    print(pos)
    print(Sys.time())
    # calculate end
    chunk_size <- nrow(x)
    end <- pos + chunk_size - 1
    # subtract 1 from some alleles
    print(paste("Start:", pos, "End:", end))
    idx_sub <- idx[pos:end]
    x[idx_sub,] <- 1 - x[idx_sub,]
    write.table(x, output, col.names=!file.exists(output), append = T, row.names = F, sep = ",", quote = F)
  }
}

# phase 1
read_csv_chunked(
  "/global/scratch/users/milesroberts/tmp/PHASE_1_2_COMBINED_hapFIRE_allele_frequency.csv",
  callback = SideEffectChunkCallback$new(
    process_one_minus(
      idx,
      5000,
      "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_2_COMBINED_hapFIRE_allele_frequency_switched.csv"
    )
  ),
  chunk_size = 5000
)

# phase 2
# read_csv_chunked(
#   "/global/scratch/users/milesroberts/tmp/PHASE2_merged_hapFIRE_allele_frequency.csv",
#   callback = SideEffectChunkCallback$new(
#     process_one_minus(
#       idx,
#       5000,
#       "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_2_hapfire_allele_frequency_rng_switched.csv"
#     )
#   ),
#   chunk_size = 5000
# )
```

# load allele frequencies

```{r}
#gen0 <- vroom("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/seedmix_hapfire_allele_frequency_rng_switched.csv", delim = ",", col_names = T)

# year 1-3 sample sizes
sample_sizes_13 <- fread("/global/home/users/milesroberts/moi_lab_projects/grenenet-phase2/data/samples_data_fix57.csv")

sample_sizes_13$site_gen_plot <- paste(sample_sizes_13$site, sample_sizes_13$fix_57_generation, sample_sizes_13$plot, sep = "_")

sample_sizes_13 <- sample_sizes_13 %>% group_by(site_gen_plot) %>% summarize(n = sum(flowerscollected))

# year 4-5 sample sizes
sample_sizes_45 <- fread("/global/home/users/milesroberts/moi_lab_projects/grenenet-phase2/data/grenenet_year4_5_rep_metadata_predicted_flower_final.csv")

sample_sizes_45$site_gen_plot <- paste(sample_sizes_45$site, sample_sizes_45$fixed_57_generation, sample_sizes_45$TrayID, sep = "_")

sample_sizes_45 <- sample_sizes_45 %>% group_by(site_gen_plot) %>% summarize(n = sum(predicted_flower_counts))

# combine all sizes
sample_sizes <- rbind(sample_sizes_13, sample_sizes_45)

# number of chromosomes = 2*flower count
sample_sizes$n <- 2*sample_sizes$n
all(sample_sizes$n > 1)

# first generation
gen0 <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/seedmix_hapfire_allele_frequency_rng_switched.csv", data.table = F)

#hapfire_phase1 <- vroom("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_hapfire_allele_frequency_rng_switched.csv", delim = ",")

#hapfire_phase1 <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_hapfire_allele_frequency_rng_switched.csv", data.table = F)

#hapfire_phase2 <- vroom("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_2_hapfire_allele_frequency_rng_switched.csv", delim = ",")

#hapfire_phase2 <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_2_hapfire_allele_frequency_rng_switched.csv", data.table = F)

#hapfire <- cbind(hapfire_phase1, hapfire_phase2)

#rm(hapfire_phase1, hapfire_phase2)
#gc()
hapfire <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/PHASE_1_2_COMBINED_hapFIRE_allele_frequency_switched.csv", data.table = F)

# omit allele frequencies near abosrbing boundaries
non_abosrbing_freq <- (gen0$X3 >= 1/231) & (gen0$X3 <= 1 - 1/231)
gen0 <- gen0[non_abosrbing_freq,]
hapfire <- hapfire[non_abosrbing_freq,]
```

## feder test (LD pruned dataset)

```{r}
# subset different generations
gen1 <- hapfire[,grepl("_1_", names(hapfire))] 

gen2 <- hapfire[,grepl("_2_", names(hapfire))] 

gen3 <- hapfire[,grepl("_3_", names(hapfire))]

gen4 <- hapfire[,grepl("_4_", names(hapfire))] 

# subset to only plots present in all 3 years
shared_gen1 <- (gsub("_1_","_",names(gen1)) %in% gsub("_2_","_",names(gen2)))& (gsub("_1_","_",names(gen1)) %in% gsub("_3_","_",names(gen3))) & (gsub("_1_","_",names(gen1)) %in% gsub("_4_","_",names(gen4)))

shared_gen2 <- (gsub("_2_","_",names(gen2)) %in% gsub("_3_","_",names(gen3))) & (gsub("_2_","_",names(gen2)) %in% gsub("_1_","_",names(gen1))) & (gsub("_2_","_",names(gen2)) %in% gsub("_4_","_",names(gen4)))

shared_gen3 <- (gsub("_3_","_",names(gen3)) %in% gsub("_2_","_",names(gen2))) & (gsub("_3_","_",names(gen3)) %in% gsub("_1_","_",names(gen1))) & (gsub("_3_","_",names(gen3)) %in% gsub("_4_","_",names(gen4)))

shared_gen4 <- (gsub("_4_","_",names(gen4)) %in% gsub("_2_","_",names(gen2))) & (gsub("_4_","_",names(gen4)) %in% gsub("_1_","_",names(gen1))) & (gsub("_4_","_",names(gen4)) %in% gsub("_3_","_",names(gen3)))

gen1_sub <- gen1[,shared_gen1]

gen2_sub <- gen2[,shared_gen2]

gen3_sub <- gen3[,shared_gen3]

gen4_sub <- gen4[,shared_gen4]

# sort columns to match
gen4_sub <- gen4_sub[,paste("1_4", c(1,4,5,6,7,8,9,10,11,12), sep = "_")]

# calculate statistic
y1 <- (gen1_sub - gen0$V3)/sqrt(2*gen0$V3*(1-gen0$V3)*(1-0))
y1$gen <- 1
y1$CHROM <- gen0$V1
y1$POS <- gen0$V2
y1 <- melt(y1, id = c("gen", "CHROM", "POS"))

y2 <- (gen2_sub - gen1_sub)/sqrt(2*gen1_sub*(1-gen1_sub)*(2-1))
y2$gen <- 2
y2$CHROM <- gen0$V1
y2$POS <- gen0$V2
y2 <- melt(y2, id = c("gen", "CHROM", "POS"))

y3 <- (gen3_sub - gen2_sub)/sqrt(2*gen2_sub*(1-gen2_sub)*(3-2))
y3$gen <- 3
y3$CHROM <- gen0$V1
y3$POS <- gen0$V2
y3 <- melt(y3, id = c("gen", "CHROM", "POS"))

yvalues <- rbind(y1, y2, y3)
yvalues$SITE <- gsub("X", "", gsub("_.*", "", yvalues$variable))
yvalues$REP <- gsub(".*_", "", yvalues$variable)

t_test_stat <- yvalues %>% group_by(CHROM, POS, SITE) %>% summarize(ybar = mean(value, na.rm = T), s2 = var(value, na.rm = T), n = n())

t_test_stat$tvalue <- t_test_stat$ybar/sqrt(t_test_stat$s2/t_test_stat$n)

t_test_stat$pvalue <- dt(t_test_stat$tvalue, df = t_test_stat$n)

# Manhattan plot
map(unique(t_test_stat$SITE), ~plot_manhattan_by_site(data = t_test_stat, site = .x, today, height = 4, width = 10))

#load("../data/allele_allyears_frequencies_long_raw_climate.rda")
# calculate mean and average

# ybar <- (y1 + y2 + y3)/3
# 
# s2 <- (1/(3-1))*( (y1 - ybar)^2 + (y2 - ybar)^2 + (y3 - ybar)^2 )
# 
# t_test_stat <- ybar/sqrt(s2/3)
# 
# # 
# t_test_stat$CHROM <- gen0$V1
# 
# t_test_stat$POS <- gen0$V2
# 
# t_test_stat <- melt(t_test_stat, id = c("CHROM", "POS"))
# 
# t_test_stat$SITE <- gsub("X", "", gsub("_.*", "", t_test_stat$variable))
# 
# t_test_stat$REP <- gsub(".*_", "", t_test_stat$variable)
# 
# # average across reps of each pos at each site
# mean_t_stat <- t_test_stat %>% 
#   group_by(CHROM, POS, SITE) %>% 
#   summarize(MEANT = mean(value, na.rm = T), SDT = sd(value, na.rm = T))
# 
# # calculate p-values
# #
# mean_t_stat$PVALUE <- dt(mean_t_stat$MEANT, df = 2)
# 
# # adjust p-values
# mean_t_stat$ADJPVALUE <- p.adjust(mean_t_stat$PVALUE, method = "bonferroni")
# 
# # plot distribution of stats across positions and sites
# ggplot(mean_t_stat, aes(x = MEANT)) +
#   geom_density()
#   xlim(c(-10, 10))
#   
# # Manhattan plot
# map(unique(mean_t_stat$SITE), ~manhattan_by_site(data = mean_t_stat, site = .x, today, height = 4, width = 10))

#
#summary(unlist(as.vector(t_test_stat)))

```

## feder test (full frequency matrix)

```{r}

# Source - https://stackoverflow.com/a/33051935
# Posted by scoa
# Retrieved 2026-02-27, License - CC BY-SA 3.0
# con <- file("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/merged_hapFIRE_allele_frequency.csv","r")
# F_header <- readLines(con,n=1)
# close(con)
# 
# F_subset <- read_csv("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/merged_hapFIRE_allele_frequency.csv", col_select = "1_1_1")

gen0 <- read.table("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/average_seedmix_p0.txt", sep = "\t")

process_F <- function(gen0, chunk_size, output) {
  function(x, pos) {
    print(pos)
    print(Sys.time())
    gen0 <- gen0[pos:(pos + chunk_size - 1),]
    # generations
    header_names <- names(x)
    gen1 <- x[,grepl("_1_", header_names)] 
    gen2 <- x[,grepl("_2_", header_names)] 
    gen3 <- x[,grepl("_3_", header_names)]
    
    # subset to populations shared across generations
    names_gen1 <- names(gen1)
    names_gen2 <- names(gen2)
    names_gen3 <- names(gen3)
    
    shared_gen1 <- (gsub("_1_","_",names_gen1) %in% gsub("_2_","_",names_gen2)) & (gsub("_1_","_",names_gen1) %in% gsub("_3_","_",names(gen3)))
    shared_gen2 <- (gsub("_2_","_",names_gen2) %in% gsub("_3_","_",names_gen3)) & (gsub("_2_","_",names_gen2) %in% gsub("_1_","_",names_gen1))
    shared_gen3 <- (gsub("_3_","_",names_gen3) %in% gsub("_2_","_",names_gen2)) & (gsub("_3_","_",names_gen3) %in% gsub("_1_","_",names_gen1))
    
    gen1_sub <- gen1[,shared_gen1]
    gen2_sub <- gen2[,shared_gen2]
    gen3_sub <- gen3[,shared_gen3] 
    # calculate statistic
    y1 <- (gen1_sub - gen0$V3)/sqrt(2*gen0$V3*(1-gen0$V3)*(1-0))
    y1$gen <- 1
    y1$CHROM <- gen0$V1
    y1$POS <- gen0$V2
    y1 <- melt(y1, id = c("gen", "CHROM", "POS"))
    
    y2 <- (gen2_sub - gen1_sub)/sqrt(2*gen1_sub*(1-gen1_sub)*(2-1))
    y2$gen <- 2
    y2$CHROM <- gen0$V1
    y2$POS <- gen0$V2
    y2 <- melt(y2, id = c("gen", "CHROM", "POS"))
    
    y3 <- (gen3_sub - gen2_sub)/sqrt(2*gen2_sub*(1-gen2_sub)*(3-2))
    y3$gen <- 3
    y3$CHROM <- gen0$V1
    y3$POS <- gen0$V2
    y3 <- melt(y3, id = c("gen", "CHROM", "POS"))
    
    yvalues <- rbind(y1, y2, y3)
    yvalues$SITE <- gsub("X", "", gsub("_.*", "", yvalues$variable))
    yvalues$REP <- gsub(".*_", "", yvalues$variable)
    
    t_test_stat <- yvalues %>% group_by(CHROM, POS, SITE) %>% summarize(ybar = mean(value, na.rm = T), s2 = var(value, na.rm = T), n = n())
    
    t_test_stat$tvalue <- t_test_stat$ybar/sqrt(t_test_stat$s2/t_test_stat$n)
    
    t_test_stat$pvalue <- dt(t_test_stat$tvalue, df = t_test_stat$n)
    
    write.table(t_test_stat, output, col.names=!file.exists(output), append = T, row.names = F, sep = ",", quote = F)
  }
}

read_csv_chunked(
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/merged_hapFIRE_allele_frequency.csv",
  #callback = SideEffectChunkCallback$new(function(x, pos){print(pos)}),
  callback = SideEffectChunkCallback$new(process_F(gen0, 1000, "feder_t_test.txt")),
  chunk_size = 1000)
                

```

## feder test - phase 2

```{r}
# make list of sites
sites <- unique(gsub("_.*", "", colnames(hapfire)))

# for now, ignore sites that lack reps with contiguous generations
#sites <- sites[c(1,3,5:17,19:21,23,25:29,31)]
sites <- sites[-30]

sites <- sites[-1:-9]

# loop over sites
output <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-03-31/", "feder_test_phase2.txt", sep = "")

for(site in sites){
  print(paste("Site:", site))
  yvalues <- NULL
  for(exprep in 1:12) {
    print(paste("Rep:", exprep))
    # subset to one site
    print("Subsetting to one replicate...")
    hap_rep <- hapfire %>% 
      select(starts_with(paste(site, "_", sep = "")) & ends_with(paste("_", exprep, sep = "")))
    
    # partition data by generation
    # should work for an arbitrary number of generations
    print("Paritioning data by generation...")
    colnames(hap_rep) <- gsub("_", "", sort(unlist(str_extract_all(colnames(hap_rep), "_.*_"))))
    
    # sort names to be in order of generation
    hap_rep$`0` <- gen0$X3
    hap_rep <- hap_rep %>% relocate(all_of(sort(colnames(hap_rep))))
    
    if(ncol(hap_rep) == 1){
      print("No columns in allele frequency matrix, skipping to next replicate")
      next
    }
    
    # calculate allele frequency differences between generations
    print("Calculating allele frequency increments...")

    years <- as.numeric(colnames(hap_rep))
    dts <- (years - lag(years))[-1]
    print("Time increments:")
    print(dts)
    # calculate y values for t test
    for (i in 1:(length(years) - 1)) {
      print(paste("Generation ", i, sep = ""))
      p0 <- hap_rep[,i]
      p1 <- hap_rep[,i + 1]
      dp <- p1 - p0
      
      yvalue <- data.frame(
        y = dp/sqrt(2*p0*(1-p0)*dts[i]),
        CHROM = gen0$X1,
        POS = gen0$X2,
        gen = i,
        SITE = site,
        rep = exprep
      )
      
      yvalues <- rbind(yvalues, yvalue)
    }
  }
  
  # calculate averages and variances for a snp at a site
  print("Averaging across replicates and generations at a site")
  yvalues <- yvalues %>% group_by(SITE, CHROM, POS) %>% 
    summarize(ybar = mean(y, na.rm = T), 
              s2 = var(y, na.rm = T),
              n = n())
  
  yvalues$tvalue <- yvalues$ybar/sqrt(yvalues$s2/yvalues$n)
  yvalues$pvalue <- dt(yvalues$tvalue, df = yvalues$n)
  
  write.table(yvalues, output, col.names=!file.exists(output), append = T, row.names = F, sep = ",", quote = F)
}
```

## plot - full feder test

```{r}
# load results of test
# remove invalid points
feder <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-03-31/feder_test_phase2.txt")

# calculate averages and variances for a snp at a site
# feder <- feder %>% group_by(site, CHROM, POS) %>% 
#   summarize(ybar = mean(y, na.rm = T), s2 = var(y, na.rm = T), n = n())
# 
# feder$tvalue <- feder$ybar/sqrt(feder$s2/feder$n)
# feder$pvalue <- dt(feder$tvalue, df = feder$n)

feder <- feder %>%
  filter(is.finite(ybar))

# choose which points to highlight
# Chr1:1185512..1188696 - CRY2
# Chr1:24331373..24333999 - FT
# Chr2:11058944..11063324 - ELF3
# Chr2:11531967..11534358 - CAM5
# Chr2:18807505..18811125 - SOC1
# Chr4:9195247..9199840 - PHYD
# Chr4:11000771..11003010 - TSF
# Chr5:5170162..5173085 - CO
# Chr5:3173382..3179448 - FLC
# Chr5:18589438..18591506 - DOG1

feder$highlight <- "no"
buffer = 10000
chroms = c(1,1,2,2,2,4,4,5,5,5)
starts = c(1185512,24331373,11058944,11531967,18807505,9195247,11000771,5170162,3173382,18589438)
ends = c(1188696,24333999,11063324,11534358,18811125,9199840,11003010,5173085,3179448,18591506)
starts = starts - buffer
ends = ends + buffer

for(i in seq_along(chroms)){
  print(i)
  feder$highlight[(feder$CHROM == chroms[i]) & (feder$POS > starts[i]) & (feder$POS < ends[i])] <- "yes"
}

# test out a few plots
#plot_manhattan_by_site(data = feder, site = 1, today, height = 3, width = 8, quick = F, prefix = "manhattan_flowering_time")

# plot them ALL!
map(unique(feder$SITE), ~plot_manhattan_by_site(data = feder, site = .x, today, height = 3, width = 8, quick = F, prefix = "manhattan_flowering_time"))

# drill down to certain peaks
# s10_c3 <- feder %>% subset(SITE == 10 & CHROM == 3)
# s10_c5 <- feder %>% subset(SITE == 10 & CHROM == 5)
# s32_c1 <- feder %>% subset(SITE == 32 & CHROM == 1)
```

## modify feder test

```{r}

p0 = 0.5
N0 = 500
L = 1:30
lambda = 100

truth = c(p0)
estims = c(p0)
covs = c(lambda)

# simulate 
for(i in L){
  # generate new generation
  x1 <- rbinom(1, N0, prob = p0)
  p1 <- x1/N0
  # sequencing
  coverage <- rpois(1, lambda = lambda)
  x1_hat <- rbinom(1, size = coverage, prob = p1)
  p1_hat <- x1_hat/coverage
  # save results
  truth = c(truth, p1)
  estims = c(estims, p1_hat)
  covs = c(covs, coverage)
}

results = data.frame(
  p = truth,
  p_hat = estims,
  cov = covs,
  i = c(0, L)
)

variance_table = data.frame(
  dt = diff(results$i),
  cov1 = results$cov[1:(nrow(results) - 1)],
  cov2 = results$cov[2:nrow(results)],
  dp = 2*results$p_hat[-1]*(1 - results$p_hat[-1])
)


variance_table$var = sqrt(variance_table$dt * variance_table$dp * (1/variance_table$cov1 + 1/variance_table$cov2))

z_stats <- diff(results$p_hat)/variance_table$var

hist(z_stats)

shapiro.test(z_stats)
```

## site climatic data

```{r}
#load("../data/sites.clim.rda")
load("../data/worldclim_sitesdata.rda")

#sites.clim <- sites.clim[,c("SITE_CODE", paste("bio", 1:19, sep = ""))]
```

## temporal autocovariance

```{r}

# change in allele frequency between t = 0 and t = 1
dp0 <- gen1_sub - gen0$V3
dp1 <- gen2_sub - gen1_sub
dp2 <- gen3_sub - gen2_sub
dp3 <- gen4_sub - gen3_sub

results = NULL

for(i in 1:ncol(dp0)){
  tmat <- cov(data.frame(
    dp0 = dp0[,i],
    dp1 = dp1[,i],
    dp2 = dp2[,i],
    dp3 = dp3[,i]
  ))
  
  sampleid = names(dp0)[i]
  site = gsub("X|_.*", "", sampleid)
  replicate = gsub(".*_", "", sampleid)
  
  # plot variance covariance matrix
  plot_var_cov_matrix(tmat, paste("../results/", today, "/", site, "_", replicate, "_var_cov_mat.pdf", sep = ""))
  
  result = gt_curves(tmat, calcabs = TRUE)
  
  result$site = site
  result$replicate = replicate
  
  results = rbind(results, result)
}

# average across reps in a site
mean_results = results %>% 
  group_by(site, gen) %>% 
  summarize(gt = mean(gt), 
            sum_cov = mean(sum_cov), 
            sum_var = mean(sum_var), 
            sum_abs_cov = mean(sum_abs_cov), 
            ratio = mean(ratio, na.rm = T))

# merge in climate data
mean_results <- merge(mean_results, worldclim_sitesdata, by = "site")

# plot mean of averages
ggplot(results, aes(
  x = gen,
  y = gt,
  group = interaction(site, replicate)
)) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(
    data = mean_results,
    aes(
      x = gen,
      y = gt,
      group = site,
      color = bio1
    ),
    lwd = 2,
    alpha = 1
  ) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name = "Generation", limits = c(1:4)) +
  labs(x = "Generation", y = "Proportion of allele frequency\nchange due to linked selection", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste("../results/", today, "/linked_selection.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)

# plot covariances
ggplot(results, aes(x = gen, y = sum_cov, group = interaction(site, replicate))) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(data = mean_results, aes(x = gen, y = sum_cov, group = site, color = bio1), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:3)) +
  labs(x = "Generation", y = "Sum of Cov(\u0394pi, \u0394pj)", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste("../results/", today, "/linked_selection_sum_cov.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)

# plot variances
ggplot(results, aes(x = gen, y = sum_var, group = interaction(site, replicate))) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(data = mean_results, aes(x = gen, y = sum_var, group = site, color = bio1), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:3)) +
  labs(x = "Generation", y = "Var(\u0394pi)", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste("../results/", today, "/linked_selection_sum_var.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)

# plot abs of covariances
ggplot(results, aes(x = gen, y = sum_abs_cov, group = interaction(site, replicate))) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(data = mean_results, aes(x = gen, y = sum_abs_cov, group = site, color = bio1), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:3)) +
  labs(x = "Generation", y = "Sum of |Cov(\u0394pi, \u0394pj)|", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste("../results/", today, "/linked_selection_sum_abs_cov.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)

# means of components
mean_results$gt2 <- mean_results$sum_abs_cov/(mean_results$sum_abs_cov + mean_results$sum_var)

ggplot(results, aes(x = gen, y = gt, group = interaction(site, replicate))) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(data = mean_results, aes(x = gen, y = gt2, group = site, color = bio1), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:3)) +
  labs(x = "Generation", y = "Proportion of allele frequency\nchange due to linked selection", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

# importance of fluctuating vs directional selection
ggplot(subset(results, gen == 3), aes(x = reorder(site, ratio, FUN = median), y = ratio)) +
  geom_boxplot() +
  geom_jitter() +
  geom_hline(yintercept = 0.5, lty = 2, lwd = 1.5) +
  labs(x = "Site", y = "Proportion of Cov(\u0394pi, \u0394pj) < 0", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste("../results/", today, "/linked_selection_directional_vs_fluctuating.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)
```

# GLM 

https://doi.org/10.1038/s41586-024-07834-x

## calculate GLM results

```{r}
# make list of sites
sites <- unique(gsub("_.*", "", colnames(hapfire)))

# for now, ignore sites that lack reps with contiguous generations
#sites <- sites[c(1,3,5:17,19:21,23,25:29,31)]
sites <- sites[-30]

# number of snps per window
window_size <- 3000

# create windows, should not bleed between chromosomes
blocks <- gen0 %>%
  group_by(X1) %>%
  mutate(block = (row_number() - 1) %/% window_size + 1,
         block = paste(X1, block, sep = "_"))

length(unique(blocks$block))

hapfire$block <- blocks$block

#output_glm <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/", today, "/glm_quasibinomial_coef.txt", sep = "")
output_glm <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-05-04/glm_quasibinomial_coef.txt", sep = "")

unique_blocks <- unique(hapfire$block)

for(b in unique_blocks[59:length(unique_blocks)]){
  print(paste("Window:", b))
  hap_block <- hapfire %>% filter(block == b)
  for(site in unique(sites)){
    print(paste("Site:", site))
    hap_site <- hap_block %>% dplyr::select(starts_with(paste(site, "_", sep = "")))
    gen0_block <- blocks %>% filter(block == b) %>% dplyr::select(starts_with("X"))
    
    # get sample sizes
    rep_n <- sample_sizes[(sample_sizes$site_gen_plot %in% colnames(hap_site)),]
    rep_n <- rbind(c(0, 1e3), rep_n)
    rep_n <- rep_n %>% arrange(site_gen_plot)
    
    # sort names to be in order of generation
    hap_site <- hap_site %>% relocate(all_of(sort(colnames(hap_site))))
    
    hap_site <- cbind(hap_site, gen0_block)
    
    hap_site <- reshape2::melt(hap_site, id = c("X1", "X2"))
    
    names(hap_site)[3] <- "site_gen_plot"
    
    hap_site <- merge(hap_site, rep_n, by = "site_gen_plot", all.x = T)
    
    hap_site$t <- gsub("_", "", str_extract(hap_site$site_gen_plot, "_.*_"))
    
    hap_site$t[hap_site$site_gen_plot == "X3"] <- 0
    
    hap_site$t <- as.numeric(hap_site$t)
    
    print("Building GLMs...")
    mod_res <- hap_site %>% group_by(X1, X2) %>%  
      filter( (value <= 1) & (value >= 0) ) %>%
      nest() %>%
      mutate(
        # calculate intercept in same scale as logit
        #fixed_intercept = map_dbl(data, ~ as.numeric(.x[(.x$site_gen_plot == "X3"), "value"])),
        #fixed_intercept_logit = log(fixed_intercept/(1-fixed_intercept)),
        # Fit the model to each nested data frame (.x)
        #model = map(data, ~glm(value ~ t - 1 + offset(rep(fixed_intercept_logit, length(value))), data = .x, weights = n, family = quasibinomial(link = "logit"))),
        # fit glm, but don't force fixed intercept
        model = map(data, ~glm(value ~ t, data = .x, weights = n, family = quasibinomial(link = "logit"))),
        # Extract coefficients in a tidy format
        tidied = map(model, tidy)
      ) %>%
      # Flatten the results back into a regular data frame
      unnest(tidied) %>% 
      filter(term == "t") %>%
      dplyr::select(c("X1", "X2", "estimate", "std.error", "statistic", "p.value"))
    
    print("Saving output...")
    mod_res$site = site # add site label
    mod_res$block = b # add site label
    write.table(mod_res, output_glm, col.names=!file.exists(output_glm), append = T, row.names = F, sep = "\t", quote = F)
    print(Sys.time())
  }
}

```
## plot GLM results

```{r}
glm_results <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-04-29/glm_quasibinomial_coef_2026-04-29.txt")

names(glm_results)[c(1,2,6,7)] = c("CHROM", "POS", "pvalue", "SITE")

glm_results$pvalue_adj <- p.adjust(glm_results$pvalue, method = "fdr")

glm_results$highlight <- "no"

glm_results$highlight[(glm_results$pvalue < 1e-14)] <- "yes"

for(site in unique(glm_results$SITE)){
  plot_manhattan_by_site(glm_results, site = site, today = today, height = 7, width = 7, prefix = "glm_manhattan")
}

```

# Simualte G(t) curves

## analyze simulations

```{r}

params <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/config/gt_params.tsv", sep = "\t", header = T, data.table = F)

neutfreq_files <- list.files("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results", pattern = "*freqs.tsv", full.names = T)

#neutfreq_files <- sample(neutfreq_files, size = 1000, replace = F)

files_to_plot <- c(
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results/1_neutfreqs.tsv",
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results/10_neutfreqs.tsv",
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results/100_neutfreqs.tsv",
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results/751_neutfreqs.tsv",
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results/841_neutfreqs.tsv",
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results/1000_neutfreqs.tsv",
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/slim_results/1001_neutfreqs.tsv"
)

sim_curves <- NULL
output <- paste("../results/", today, "/simulated_gt_curves_" , today, ".txt", sep = "")

gens_to_skip <- 0

for(neutfreq_file in neutfreq_files){
  simid <- as.numeric(gsub(".*/", "", gsub("_neutfreqs.tsv", "", neutfreq_file)))
  
  print(paste(Sys.time(),": Simulation ID: ", simid, sep = ""))
  
  no_col <- max(count.fields(neutfreq_file, sep = "\t"))

  neutfreqs <- fread(neutfreq_file, fill = T, col.names = paste("X",1:no_col, sep=""), sep = "\t", data.table = F)

  neutfreqs <- reshape2::melt(neutfreqs, id = "X1")

  neutfreqs$pos <- str_extract(neutfreqs$value, ";.*;")
  neutfreqs$freq <- str_extract(neutfreqs$value, ";0.*$")

  neutfreqs$pos <- as.numeric(gsub(";", "", neutfreqs$pos))
  neutfreqs$freq <- as.numeric(gsub(";", "", neutfreqs$freq))

  neutfreqs <- neutfreqs %>%
    group_by(pos) %>%
    mutate(dp = freq - lag(freq))

  # Source - https://stackoverflow.com/a/5890689
  # Posted by Ista, modified by community. See post 'Timeline' for change history
  # Retrieved 2026-02-21, License - CC BY-SA 3.0
  neutfreqs <- neutfreqs[,c("X1", "pos", "dp")]
  neutfreqs <- neutfreqs[complete.cases(neutfreqs),]
  neutfreqs <- neutfreqs[!duplicated(neutfreqs[,c("X1", "pos")]),]
  dpmat <- reshape2::dcast(neutfreqs, pos ~ X1, value.var = "dp")

  #dpmat <- dpmat %>% filter(rowSums(is.na(.)) <= ncol(dpmat)*0.5)
  dpmat <- dpmat[,c(-1,-2)]
  
  #dpmat <- dpmat[complete.cases(dpmat),]

  dpcov_mat <- cov(dpmat, use = "pairwise.complete.obs")
  
  if(neutfreq_file %in% files_to_plot){
    print("Plotting variance covariance matrix...")
    plot_var_cov_matrix(dpcov_mat, paste("../results/", today, "/", simid, "_var_cov_mat.pdf", sep = ""), include_values = F)
  }
  
  sim_curve <- gt_from_covmat(dpcov_mat)

  param_row <- params[(params$ID == simid),]
  
  sim_curve$ID <- simid
  sim_curve$alpha <- param_row[, "alpha"]
  sim_curve$sigma <- param_row[, "sigma"]
  sim_curve$tmu <- param_row[, "tmu"]
  sim_curve$N <- param_row[, "N"]
  sim_curve$adjust <- param_row[, "adjust"]
  
  write.table(sim_curve, output, col.names=!file.exists(output), append = T, row.names = F, sep = ",", quote = F)
  
}

```

## plot simulations

```{r}
#sim_curves <- fread("simulated_gt_curves_2026-03-31.txt")
#sim_curves <- fread("simulated_gt_curves_2026-04-08.txt")
#sim_curves <- fread("simulated_gt_curves_2026-04-14.txt")
#sim_curves <- fread("../results/2026-04-16/simulated_gt_curves_2026-04-16.txt")
#sim_curves <- fread("../results/2026-04-26/simulated_gt_curves_2026-04-26.txt")
#sim_curves <- fread("../results/2026-04-27/simulated_gt_curves_2026-04-27.txt")
sim_curves <- fread("../results/2026-04-28/simulated_gt_curves_2026-04-28.txt")

sim_curves <- merge(params[,c("ID", "MIN_AGE")], sim_curves, by = "ID")

# average across reps in a site
mean_sims = sim_curves %>% 
  group_by(gen, sigma, alpha, adjust, MIN_AGE) %>% 
  summarize(gt = mean(sum_cov)/mean(sum_cov + sum_var),
            gt_abs = mean(sum_abs_cov)/mean(sum_abs_cov + sum_var), 
            sum_cov = mean(sum_cov), 
            sum_abs_cov = mean(sum_abs_cov),
            sum_pos_cov = mean(sum_pos_cov),
            sum_neg_cov = mean(sum_neg_cov),
            sum_var = mean(sum_var), 
            ratio = mean(ratio, na.rm = T),
            n = n())

mean_sims$adjust[mean_sims$adjust] <- "Ne constant"
mean_sims$adjust[(mean_sims$adjust == "FALSE")] <- "Ne lowered"

mean_sims$MIN_AGE[(mean_sims$MIN_AGE == 1)] <- "Non-WF"
mean_sims$MIN_AGE[(mean_sims$MIN_AGE == 0)] <- "Non-WF reduced"

# gt over time
ggplot(mean_sims, aes(x = gen, y = gt, group = sigma, color = as.factor(sigma))) +
  geom_point(size = 1.5) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  facet_grid(adjust+MIN_AGE~alpha, scales = "free") +
  geom_hline(yintercept = 0, lwd = 1.1, lty = 2) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "G(t)", color = "Self\nrate")

ggsave(paste("../results/", today, "/simulated_linked_selection_gt.pdf", sep = ""), height = 4.5, width = 8, device = cairo_pdf)

# sum of covariances over time
ggplot(mean_sims, aes(x = gen, y = sum_cov, group = sigma, color = as.factor(sigma))) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, lty = 2, lwd = 1.1) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  facet_grid(adjust+MIN_AGE~alpha) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "Sum of Cov(\u0394pi, \u0394pj)", color = "Self\nrate")

ggsave(paste("../results/", today, "/simulated_linked_selection_sum_cov.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)

# sum of absolute value of covariances over time
ggplot(mean_sims, aes(x = gen, y = sum_abs_cov, group = sigma, color = as.factor(sigma))) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, lty = 2, lwd = 1.1) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  facet_grid(adjust+MIN_AGE~alpha) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "Sum of |Cov(\u0394pi, \u0394pj)|", color = "Self\nrate")

ggsave(paste("../results/", today, "/simulated_linked_selection_sum_abs_cov.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)

# sum variances over time
ggplot(mean_sims, aes(x = gen, y = sum_var, group = interaction(sigma, alpha), color = as.factor(sigma))) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, lty = 2, lwd = 1.1) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  facet_grid(adjust+MIN_AGE~alpha) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "Sum of Var(\u0394pi)", color = "Self\nrate", linetype = "|s| > 0")

ggsave(paste("../results/", today, "/simulated_linked_selection_sum_var.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)

# importance of fluctuating vs directional selection
ggplot(mean_sims, aes(x = gen, y = ratio, group = sigma, color = as.factor(sigma))) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0.5, lty = 2, lwd = 1.1) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  facet_grid(adjust+MIN_AGE~alpha) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "Proportion of Cov(\u0394pi, \u0394pj) > 0", color = "Self\nrate")

ggsave(paste("../results/", today, "/simulated_linked_selection_ratio.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)

# try using averages of variances components
mean_sims$gt2 = mean_sims$sum_cov/(mean_sims$sum_cov + mean_sims$sum_var)

ggplot(mean_sims, aes(x = gen, y = gt2, group = interaction(sigma, alpha), color = as.factor(sigma))) +
  geom_point(size = 1.5) +
  geom_hline(yintercept = 0, lty = 2, lwd = 1.1) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  facet_grid(adjust+MIN_AGE~alpha) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "G(t)\nmean components", color = "Self\nrate")

ggsave(paste("../results/", today, "/simulated_linked_selection_gt_mean_components.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)

# absolute value of covariance
#ggplot(subset(mean_sims,gen <=10), aes(x = gen, y = sum_abs_cov/(sum_abs_cov + sum_var), group = interaction(sigma, alpha), color = as.factor(sigma))) +
ggplot(mean_sims, aes(x = gen, y = gt_abs, group = interaction(sigma, alpha), color = as.factor(sigma))) +
  geom_point(size = 1.5) +
  #geom_vline(xintercept = 5) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  facet_grid(adjust+MIN_AGE~alpha) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "G(t)\nabs. value", color = "Self\nrate")

ggsave(paste("../results/", today, "/simulated_linked_selection_gt_abs.pdf", sep = ""), height = 7, width = 5, device = cairo_pdf)

# sum of negative vs positive cov
ggplot(mean_sims, aes(x = gen, y = sum_neg_cov, group = interaction(sigma, alpha), color = as.factor(sigma))) +
  geom_point(size = 1.5) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  facet_grid(adjust+MIN_AGE~alpha) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "Sum Cov < 0", color = "Self\nrate")

ggsave(paste("../results/", today, "/simulated_linked_selection_sum_neg_cov.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)

ggplot(mean_sims, aes(x = gen, y = sum_pos_cov, group = interaction(sigma, alpha), color = as.factor(sigma))) +
  geom_point(size = 1.5) +
  scale_color_scico_d(palette = "acton", begin = 0.1, end = 0.9) +
  facet_grid(adjust+MIN_AGE~alpha) +
  geom_line() +
  theme(legend.position = "top") +
  labs(x = "Generation", y = "Sum Cov > 0", color = "Self\nrate")

ggsave(paste("../results/", today, "/simulated_linked_selection_sum_pos_cov.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)
```

# Temporal covariances - phase 2

```{r}
# make list of sites
sites <- unique(gsub("_.*", "", colnames(hapfire)))

# for now, ignore sites that lack reps with contiguous generations
#sites <- sites[c(1,3,5:17,19:21,23,25:29,31)]
sites <- sites[-30]

# loop over sites
all_gt_curves = NULL
melted_cor_dps = NULL
melted_cov_dps = NULL
for(site in sites) {
  print(paste("Site:", site))
  for(exprep in 1:12) {
    print(paste("Rep:", exprep))
    # subset to one site
    print("Subsetting to one replicate...")
    hap_rep <- hapfire %>% select(starts_with(paste(site, "_", sep = "")) & ends_with(paste("_", exprep, sep = "")))
    
    # partition data by generation
    # should work for an arbitrary number of generations
    print("Paritioning data by generation...")
    colnames(hap_rep) <- gsub("_", "", sort(unlist(str_extract_all(colnames(hap_rep), "_.*_"))))
    
    # sort names to be in order of generation
    hap_rep <- hap_rep %>% relocate(all_of(sort(colnames(hap_rep))))
    
    if(ncol(hap_rep) == 0){
      print("No columns in allele frequency matrix, skipping to next replicate")
      next
    }
    
    gens <- list()
    gens[[1]] <- gen0$X3
    names(gens)[1] <- "0"
    for(i in 1:ncol(hap_rep)){
      gens[[i + 1]] <- hap_rep[,i]
      names(gens)[i+1] <- colnames(hap_rep)[i]
    }
    
    # calculate allele frequency differences between generations
    print("Calculating allele frequency increments...")
    dps <- NULL
    for (i in 1:(length(gens) - 1)) {
      print(i)
      dps <- cbind(dps, gens[[i + 1]] - gens[[i]])
      colnames(dps)[i] <- paste(names(gens)[i], "->", names(gens)[i+1], sep = "")
    }
    
    # calculate and save cov matrix
    cov_dps <- cov(dps)
    cor_dps <- cor(dps)
    
    melted_cov_dp <- reshape2::melt(cov_dps)
    melted_cov_dp$site <- site
    melted_cov_dp$replicate <- exprep
    
    melted_cov_dps <- rbind(melted_cov_dps, melted_cov_dp)
    melted_cor_dps <- rbind(melted_cor_dps, reshape2::melt(cor_dps))
    
    # calculate gt curves
    rep_gt <- gt_from_covmat(cov_dps)
    
    # save result
    rep_gt$site <- site
    rep_gt$replicate <- exprep
    rep_gt$year <- colnames(hap_rep)
    all_gt_curves = rbind(all_gt_curves, rep_gt)
  }
}

# save gt curves, clean-up table
all_gt_curves$gen <- all_gt_curves$year
all_gt_curves <- all_gt_curves %>% select(!c("gt", "gt_abs", "year"))

write.table(all_gt_curves, paste("../results/", today, "/gt_curves_", today, ".txt", sep = ""), sep = "\t", row.names = F, quote = F)
```

## plot temporal covariances - phase 2
```{r}
all_gt_curves <- fread("../results/2026-04-10/gt_curves_2026-04-10.txt", data.table = F)

# prepare cov for line plot
contiguous_increments <- c("0->1", "1->2", "2->3", "3->4", "4->5")
melted_cov_dps <- melted_cov_dps[(melted_cov_dps$Var1 %in% contiguous_increments) & (melted_cov_dps$Var2 %in% contiguous_increments),]

#melted_cov_dps <- melted_cov_dps[!duplicated(melted_cov_dps$value),]

melted_cov_dps$t2 <- gsub(".*->", "", melted_cov_dps$Var2)

melted_cov_dps <- melted_cov_dps[(melted_cov_dps$Var1 != melted_cov_dps$Var2),]

melted_cov_dps <- merge(melted_cov_dps, worldclim_sitesdata, by = "site")

ggplot(subset(melted_cov_dps, Var1 == "4->5") %>% group_by(site, bio1, Var1, t2) %>% summarize(mean_cov = mean(value)), aes(x = t2, y = mean_cov, group = interaction(site, Var1), color = bio1, shape = Var1)) +
  scale_color_scico(palette = "berlin") +
  geom_line(lwd = 1.1) +
  geom_point(size =2) +
  #geom_line(mapping = aes(x = t2, y = mean_cov, group = Var1), data = melted_cov_dps %>% group_by(Var1, t2) %>% summarize(mean_cov = mean(value)), color = "black") +
  labs(x = "\u0394pj", y = "Mean Cov(\u0394pi, \u0394pj)", shape = "\u0394pi")

ggsave(paste("../results/", today, "/linked_selection_mean_cov_4-5.pdf", sep = ""), height = 7, width = 11, device = cairo_pdf)

# remove sites with only one generation of data
all_gt_curves <- all_gt_curves %>% filter(site != "26" & site != "23" & site != "37" & site != "60")

# calculate modified gt based on absolute value of cov
# and use the mean of the components
all_gt_curves$gt <- all_gt_curves$sum_cov/(all_gt_curves$sum_abs_cov + all_gt_curves$sum_var)

all_gt_curves$gt2 <- all_gt_curves$sum_abs_cov/(all_gt_curves$sum_abs_cov + all_gt_curves$sum_var)

# remove replicates with just one observation outside of year 1
all_gt_curves <- all_gt_curves[!((all_gt_curves$gt2 == 0) & (all_gt_curves$gen > 1)),]

# average across reps in a site
mean_gts = all_gt_curves %>% 
  group_by(site, gen) %>% 
  summarize(sum_cov = mean(sum_cov), 
            sum_var = mean(sum_var), 
            sum_abs_cov = mean(sum_abs_cov),
            sum_pos_cov = mean(sum_pos_cov),
            sum_neg_cov = mean(sum_neg_cov),
            ratio = mean(ratio, na.rm = T),
            n = n())

write.table(mean_gts, paste("../results/", today, "/mean_gt_curves_", today, ".txt", sep = ""), sep = "\t", row.names = F, quote = F)

#mean_gts <- mean_gts %>% filter(n > 2 & n <= 12)

# merge in climate data
mean_gts <- merge(mean_gts, worldclim_sitesdata, by = "site")

# estimate gt as the mean of its components
mean_gts$gt <- mean_gts$sum_cov/(mean_gts$sum_cov + mean_gts$sum_var)
mean_gts$gt2 <- mean_gts$sum_abs_cov/(mean_gts$sum_abs_cov + mean_gts$sum_var)

# absolute value version
ggplot(all_gt_curves, aes(x = as.numeric(gen), y = gt2, group = interaction(site, replicate))) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(data = mean_gts, aes(x = as.numeric(gen), y = gt2, group = site, color = bio1), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:5)) +
  labs(x = "Generation", y = "Proportion of allele frequency\nchange due to linked selection", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste("../results/", today, "/linked_selection_mean_components_abs.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)

# regular sum
ggplot(all_gt_curves, aes(x = as.numeric(gen), y = gt, group = interaction(site, replicate))) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(data = mean_gts, aes(x = as.numeric(gen), y = gt, group = site, color = bio1), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:5)) +
  labs(x = "Generation", y = "Proportion of allele frequency\nchange due to linked selection", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste("../results/", today, "/linked_selection_mean_components.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)

# plot sites individually
for (s in unique(all_gt_curves$site)){
  ggplot(subset(all_gt_curves, site == s), aes(x = as.numeric(gen), y = gt, group = interaction(site, replicate))) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(data = subset(mean_gts, site == s), aes(x = as.numeric(gen), y = gt, group = site), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:5)) +
  labs(x = "Generation", y = "Proportion of allele frequency\nchange due to linked selection")

ggsave(paste("../results/", today, "/linked_selection_mean_components_", s, ".pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)

ggplot(subset(all_gt_curves, site == s), aes(x = as.numeric(gen), y = gt2, group = interaction(site, replicate))) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(data = subset(mean_gts, site == s), aes(x = as.numeric(gen), y = gt2, group = site), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:5)) +
  labs(x = "Generation", y = "Proportion of allele frequency\nchange due to linked selection")

ggsave(paste("../results/", today, "/linked_selection_mean_components_abs", s, ".pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)
}


# another plot where we filter to only the super well-replicated data

# number of replicates by year, by site
ggplot(mean_gts, aes(x = year, y = n)) +
  geom_point() +
  geom_jitter() +
  geom_boxplot()

# covariances across all sites
mean_cor_dps <- melted_cor_dps %>% group_by(Var1, Var2) %>% summarize(sum_cor = mean(value)) 

#contiguous_increments <- c("0->1", "1->2", "2->3", "3->4", "4->5")
contiguous_increments <- c("1->2", "2->3", "3->4", "4->5")

mean_cor_dps <- mean_cor_dps[(mean_cor_dps$Var1 %in% contiguous_increments) &(mean_cor_dps$Var2 %in% contiguous_increments),]

mean_cor_dps_tib <- mean_cor_dps %>% pivot_wider(names_from = Var1, values_from = sum_cor)

# Convert to matrix while excluding the name column
mean_cor_dps_mat <- as.matrix(mean_cor_dps_tib[, -1])
# Assign row names from the original column
rownames(mean_cor_dps_mat) <- mean_cor_dps_tib$Var2

plot_var_cov_matrix(mean_cor_dps_mat, paste("../results/", today, "/mean_correlation_across_all_reps.pdf", sep = ""), legend_name = "Mean\ncor(\u0394pi, \u0394pj)")

# plot mean of averages
# ggplot(all_gt_curves, aes(
#   x = as.numeric(year),
#   y = gt,
#   group = interaction(site, replicate)
# )) +
#   geom_point(alpha = 0.1) +
#   geom_line(alpha = 0.1) +
#   geom_line(
#     data = mean_gts,
#     aes(
#       x = as.numeric(year),
#       y = gt,
#       group = site,
#       color = bio1
#     ),
#     lwd = 2,
#     alpha = 1
#   ) +
#   scale_color_scico(palette = "berlin") +
#   scale_x_discrete(name = "Generation", limits = c(1:5)) +
#   labs(x = "Generation", y = "Proportion of allele frequency\nchange due to linked selection", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")
# 
# ggsave(paste("../results/", today, "/linked_selection.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)
# 

# plot covariances
ggplot(all_gt_curves, aes(x = year, y = sum_cov, group = interaction(site, replicate))) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(data = mean_gts, aes(x = year, y = sum_cov, group = site, color = bio1), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:5)) +
  labs(x = "Generation", y = "Sum of Cov(\u0394pi, \u0394pj)", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")
  #theme(legend.position = "top")
  #facet_zoom(ylim = c(0.05, -0.05), show.area = T)

ggsave(paste("../results/", today, "/linked_selection_sum_cov.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)

# # plot variances
# ggplot(all_gt_curves, aes(x = gen, y = sum_var, group = interaction(site, replicate))) +
#   geom_point(alpha = 0.1) +
#   geom_line(alpha = 0.1) +
#   geom_line(data = mean_gts, aes(x = gen, y = sum_var, group = site, color = bio1), lwd = 2, alpha = 1) +
#   scale_color_scico(palette = "berlin") +
#   scale_x_discrete(name ="Generation", limits=c(1:5)) +
#   labs(x = "Generation", y = "Var(\u0394pi)", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")
# 
# ggsave(paste("../results/", today, "/linked_selection_sum_var.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)
# 
# # plot positive and negative covariances
# all_gt_curves_posneg <- all_gt_curves %>% pivot_longer(cols = c(sum_pos_cov, sum_neg_cov))
# 
# mean_gts_posneg <- mean_gts %>% pivot_longer(cols = c(sum_pos_cov, sum_neg_cov))
# 
# all_gt_curves_posneg[(all_gt_curves_posneg$name == "sum_neg_cov"),"value"] <- all_gt_curves_posneg[(all_gt_curves_posneg$name == "sum_neg_cov"),"value"]*(-1)
# 
# mean_gts_posneg[(mean_gts_posneg$name == "sum_neg_cov"),"value"] <- mean_gts_posneg[(mean_gts_posneg$name == "sum_neg_cov"),"value"]*(-1)
#                                                                                                                                                           
# ggplot(all_gt_curves_posneg, aes(x = gen, y = value, group = interaction(site,name, replicate))) +
#   facet_wrap(.~name, nrow = 2, ncol = 1, scales = "free_y")+
#   geom_point(alpha = 0.1) +
#   geom_line(alpha = 0.1) +
#   geom_hline(yintercept = 0, lwd = 1.2, lty = 2) +
#   geom_line(data = mean_gts_posneg, aes(x = gen, y = value, group = interaction(site,name), color = bio1), lwd = 2, alpha = 1) +
#   scale_color_scico(palette = "berlin") +
#   scale_x_discrete(name ="Generation", limits=c(1:5)) +
#   labs(x = "Generation", y = "Sum of Cov(\u0394pi, \u0394pj)", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")
# 
# ggsave(paste("../results/", today, "/linked_selection_sum_cov_components.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)
# 
# # plot abs of covariances
# ggplot(all_gt_curves, aes(x = gen, y = sum_abs_cov, group = interaction(site, replicate))) +
#   geom_point(alpha = 0.1) +
#   geom_line(alpha = 0.1) +
#   geom_line(data = mean_gts, aes(x = gen, y = sum_abs_cov, group = site, color = bio1), lwd = 2, alpha = 1) +
#   scale_color_scico(palette = "berlin") +
#   scale_x_discrete(name ="Generation", limits=c(1:5)) +
#   labs(x = "Generation", y = "Sum of |Cov(\u0394pi, \u0394pj)|", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")
# 
# ggsave(paste("../results/", today, "/linked_selection_sum_abs_cov.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)
# 
# 
# 
# # importance of fluctuating vs directional selection
# ggplot(subset(all_gt_curves, gen == 5), aes(x = reorder(site, ratio, FUN = median), y = ratio)) +
#   geom_boxplot() +
#   geom_jitter() +
#   geom_hline(yintercept = 0.5, lty = 2, lwd = 1.5) +
#   labs(x = "Site", y = "Proportion of Cov(\u0394pi, \u0394pj) < 0", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")
# 
# ggsave(paste("../results/", today, "/linked_selection_directional_vs_fluctuating.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)
```

## compare temperature covariances to allele frequency covariances
```{r}

load("../data/climate_sites_experiment.rda")
load("../data/climate_sites_yearly.rda")

# Generations 1-5 → years 2018-2022
# Transition 0->1 uses P0 (common seed stock, no site climate), skip it for climate
# Allele transitions: 0->1, 1->2, 2->3, 3->4, 4->5
# Climate transitions: 1->2, 2->3, 3->4, 4->5 (year-to-year at sites)
gen_year <- data.frame(generation = 1:5, year = 2018:2022)

# Get yearly TerraClimate temperature (bio1) per site
clim_annual <- climate_sites_yearly %>%
  mutate(site = as.numeric(site)) %>%
  dplyr::select(site, year, bio1) %>%
  filter(year >= 2018, year <= 2022) %>%
  left_join(gen_year, by = "year") %>%
  arrange(site, year)

cat(sprintf("TerraClimate data: %d sites, years %s\n",
            length(unique(clim_annual$site)),
            paste(sort(unique(clim_annual$year)), collapse = ", ")))

clim_increments <- clim_annual %>%
  group_by(site) %>%
  arrange(generation) %>%
  mutate(
    dT = bio1 - lag(bio1),
    interval = paste0(lag(generation), "->", generation)
  ) %>%
  filter(!is.na(dT)) %>%
  ungroup()

cat(sprintf("\nClimate increments: %d observations across %d sites\n",
            nrow(clim_increments), length(unique(clim_increments$site))))

# Build climate increment matrix: sites x intervals
clim_mat <- clim_increments %>%
  dplyr::select(site, interval, dT) %>%
  pivot_wider(id_cols = site, names_from = interval, values_from = dT)

clim_avail <- sort(setdiff(colnames(clim_mat), "site"))
clim_matrix <- as.matrix(clim_mat[, clim_avail])
clim_comp <- complete.cases(clim_matrix)
clim_matrix <- clim_matrix[clim_comp, ]

cat(sprintf("Sites with complete climate data: %d\n", nrow(clim_matrix)))

# Q_climate: covariance of climate increments across sites
Q_climate <- cov(clim_matrix)

# Correlation version
R_climate <- cor(clim_matrix)

plot_var_cov_matrix(R_climate, paste("../results/", today, "/climate_increment_correlations.pdf", sep = ""), legend_name = "Mean\ncor(\u0394Ti, \u0394Tj)", low_col = "#023198", mid_col = "white", high_col = "#7E1700")

# see if covariant allele frequency trajectories follow covariant temperate trajectories, by site
#mean_dp_cov_by_site <- as.data.frame(melted_cov_dps %>% group_by(site, Var1, Var2) %>% summarize(mean_cov = mean(value)) %>% filter(!str_detect(Var1, "0") & !str_detect(Var2, "0")))

mean_dp_cov_by_site <- as.data.frame(melted_cov_dps %>% filter(!str_detect(Var1, "0") & !str_detect(Var2, "0")))

mean_dp_cov_by_site <- mean_dp_cov_by_site[!duplicated(mean_dp_cov_by_site$value),]

# observation
obs_dp_dt_match <- check_dp_match_dt(mean_dp_cov_by_site, clim_increments)

# permutation test
permuted_dp <- permute(mean_dp_cov_by_site, n = 1000, value)

permuted_dt <- permute(clim_increments, n = 1000, dT)

list_of_dps <- permuted_dp$perm %>% 
  map(as.data.frame)

list_of_dts <-permuted_dt$perm %>% 
  map(as.data.frame)

null_dist <- mapply(check_dp_match_dt, x = list_of_dps, y = list_of_dts)

null_dist <- as.data.frame(t(null_dist))

ggplot(null_dist, aes(x = `TRUE`/(`TRUE`+`FALSE`))) +
  geom_density(lwd = 2) +
  geom_vline(xintercept = obs_dp_dt_match["TRUE"]/sum(obs_dp_dt_match), lwd = 2, color = "red",) +
  labs(x = "Proportion of sgn Cov(\u0394pi, \u0394pj) = sgn \u0394Ti\u0394Tj", title = "Permuted null")

ggsave(paste("../results/", today, "/linked_selection_cov_matches_bio1.pdf", sep = ""), height = 7, width = 7, device = cairo_pdf)

# Compute anomalies
# "anomaly" is bio1 - average bio1 (residual bio1), which means there will by definition be some positive and some negative anomalies at each site
clim_anom <- clim_annual %>%
  group_by(site) %>%
  mutate(site_mean = mean(bio1),
         anomaly = bio1 - site_mean) %>%
  ungroup() %>%
  left_join(climate_sites %>% mutate(site = as.numeric(site)) %>%
              dplyr::select(site, site_bio1), by = "site")

# Anomaly heatmap: sites (sorted by temperature) x years
anom_wide <- clim_anom %>%
  dplyr::select(site, year, anomaly, site_bio1) %>%
  arrange(site_bio1)

p_anom_heat <- anom_wide %>%
  mutate(site = factor(site, levels = unique(site[order(site_bio1)]))) %>%
  ggplot(aes(x = factor(year), y = site, fill = anomaly)) +
  geom_tile(color = "grey90") +
  geom_text(aes(label = sprintf("%.1f", anomaly)), size = 2.5) +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red",
                       midpoint = 0, name = "Anomaly\n(°C)") +
  labs(title = "Temperature anomaly per site per year (TerraClimate)",
       subtitle = "Sites sorted by mean temperature (cold at bottom, warm at top)",
       x = "Year (generation)", y = "Site") +
  theme_minimal(base_size = 11) +
  theme(axis.text.y = element_text(size = 6))

# Mean anomaly across sites per year
mean_anom <- clim_anom %>%
  group_by(year) %>%
  summarise(mean_anomaly = mean(anomaly), se = sd(anomaly) / sqrt(n()), .groups = "drop")

p_anom_bar <- mean_anom %>%
  ggplot(aes(x = factor(year), y = mean_anomaly, fill = mean_anomaly > 0)) +
  geom_col(alpha = 0.7, width = 0.6) +
  geom_errorbar(aes(ymin = mean_anomaly - 1.96*se, ymax = mean_anomaly + 1.96*se), width = 0.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_fill_manual(values = c("TRUE" = "red", "FALSE" = "blue"), guide = "none") +
  labs(title = "Mean temperature anomaly across sites",
       subtitle = "Warm/cold years drive allele frequency reversals",
       x = "Year (generation)", y = "Mean anomaly (°C)") +
  theme_minimal(base_size = 12)

p_anom_heat

p_anom_bar

# compare anomalies to bioclim
anom_vs_bioclim <- clim_annual %>%
  left_join(climate_sites %>% mutate(site = as.numeric(site)) %>%
              dplyr::select(site, site_bio1), by = "site") %>%
  mutate(anomaly_bioclim = bio1 - site_bio1)

cat("Anomaly = TerraClimate bio1 - bioclim baseline (site_bio1)\n")

# Build anomaly matrix: sites x years
anom_wide_bioclim <- anom_vs_bioclim %>%
  dplyr::select(site, year, anomaly_bioclim) %>%
  pivot_wider(id_cols = site, names_from = year, values_from = anomaly_bioclim)

anom_matrix <- as.matrix(anom_wide_bioclim[, -1])
rownames(anom_matrix) <- anom_wide_bioclim$site

# Covariance and correlation of anomalies across sites
Cov_anom <- cov(anom_matrix)
Cor_anom <- cor(anom_matrix)

cat("\nAnomaly covariance matrix (TerraClim - bioclim baseline):\n")

plot_var_cov_matrix(Cor_anom, paste("../results/", today, "/climate_anomalies_correlations.pdf", sep = ""), legend_name = "Mean\ncor(Ai, Aj)", low_col = "#023198", mid_col = "#C0E9C2", high_col = "#7E1700")

# try cmip6 model
bbox <- get_1km_bbox(37.8715, -122.2730)

area <- c(
  bbox$xmin,
  bbox$xmax,
  bbox$ymin,
  bbox$ymax
)

download_cmip6_ecmwfr(
  scenario = "ssp126",
  area = area,
  model = "default",
  service = "cds",
  frequency = "monthly",
  variable = c("Tmin", "Tmax"),
  year_start = 2000,
  year_end = 2015,
  month = 1:12,
  sec_wait = 3600,
  n_try = 10,
  update_everything = FALSE,
  path_download = "cmip6_downloaded"
)


```

## temporal covariances - phase 2 - sample size correction
```{r}
# make list of sites
sites <- unique(gsub("_.*", "", colnames(hapfire)))

# for now, ignore sites that lack reps with contiguous generations
#sites <- sites[c(1,3,5:17,19:21,23,25:29,31)]
sites <- sites[-30]

# loop over sites
all_gt_curves = NULL

for(site in sites) {
  print(paste("Site:", site))
  for(exprep in 1:12) {
    print(paste("Rep:", exprep))
    # subset to one site
    hap_rep <- hapfire %>% select(starts_with(paste(site, "_", sep = "")) & ends_with(paste("_", exprep, sep = "")))
    
    if(ncol(hap_rep) <= 1){
      print("Need at least two increments. Skipping to next replicate...")
      next
    }
    
    # get sample sizes
    rep_n <- sample_sizes[(sample_sizes$site_gen_plot %in% colnames(hap_rep)),]
    rep_n <- rbind(c(0, 1e5), rep_n)
    rep_n <- rep_n %>% arrange(site_gen_plot)
    
    # sort names to be in order of generation
    hap_rep <- hap_rep %>% relocate(all_of(sort(colnames(hap_rep))))
    
    # mark fixed or lost alleles as missing
    #hap_rep[( (hap_rep < 1/231) | (hap_rep > 1 - 1/231) )] <- NA
    
    # create matrix of allele frequencies
    # L X G
    pmat <- NULL
    pmat <- cbind(pmat, gen0$X3)
    colnames(pmat)[1] <- "0"
    for(i in 1:ncol(hap_rep)){
      pmat <- cbind(pmat, hap_rep[,i])
      colnames(pmat)[i+1] <- colnames(hap_rep)[i]
    }
    
    rep_gt <- gt_n_adjust(pmat, rep_n$n)
    
    # save result
    rep_gt$site <- site
    rep_gt$replicate <- exprep
    rep_gt$gen <- gsub("_", "", str_extract(colnames(hap_rep), "_.*_"))
    all_gt_curves <- rbind(all_gt_curves, rep_gt)
  }
}

#### plot ####

# remove sites with only one generation of data
#all_gt_curves <- all_gt_curves %>% filter(site != "26" & site != "23" & site != "37" & site != "60")

# calculate modified gt based on absolute value of cov
# and use the mean of the components
all_gt_curves$gt <- all_gt_curves$sum_cov/(all_gt_curves$sum_abs_cov + all_gt_curves$sum_var)

all_gt_curves$gt_abs <- all_gt_curves$sum_abs_cov/(all_gt_curves$sum_abs_cov + all_gt_curves$sum_var)

# remove replicates with just one observation outside of year 1
all_gt_curves <- all_gt_curves[!((all_gt_curves$gt_abs == 0) & (all_gt_curves$gen > 1)),]

# average across reps in a site
mean_gts = all_gt_curves %>% 
  group_by(site, gen) %>% 
  summarize(gt = mean(sum_cov)/mean(sum_cov + sum_var),
            gt_abs = mean(sum_abs_cov)/mean(sum_abs_cov + sum_var),
            sum_cov = mean(sum_cov), 
            sum_var = mean(sum_var), 
            sum_abs_cov = mean(sum_abs_cov),
            sum_pos_cov = mean(sum_pos_cov),
            sum_neg_cov = mean(sum_neg_cov),
            ratio = mean(ratio, na.rm = T),
            n = n())

write.table(mean_gts, paste("../results/", today, "/mean_gt_curves_", today, ".txt", sep = ""), sep = "\t", row.names = F, quote = F)

# merge in climate data
mean_gts <- merge(mean_gts, worldclim_sitesdata, by = "site")

# estimate gt as the mean of its components
# mean_gts$gt <- mean_gts$sum_cov/(mean_gts$sum_cov + mean_gts$sum_var)
# mean_gts$gt_abs <- mean_gts$sum_abs_cov/(mean_gts$sum_abs_cov + mean_gts$sum_var)

# linked selection curves
ggplot(all_gt_curves, aes(x = as.numeric(gen), y = gt, group = interaction(site, replicate))) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(data = mean_gts, aes(x = as.numeric(gen), y = gt, group = site, color = bio1), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:5)) +
  #ylim(0,1) +
  labs(x = "Generation", y = "Proportion of allele frequency\nchange due to linked selection", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

# absolute value version
ggplot(all_gt_curves, aes(x = as.numeric(gen), y = gt_abs, group = interaction(site, replicate))) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(data = mean_gts, aes(x = as.numeric(gen), y = gt_abs, group = site, color = bio1), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:5)) +
  #ylim(0,1) +
  labs(x = "Generation", y = "Proportion of allele frequency\nchange due to linked selection", color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste("../results/", today, "/linked_selection_mean_components_abs.pdf", sep = ""), height = 5, width = 7, device = cairo_pdf)

for (s in unique(all_gt_curves$site)){
  ggplot(subset(all_gt_curves, site == s), aes(x = as.numeric(gen), y = gt, group = interaction(site, replicate))) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(data = subset(mean_gts, site == s), aes(x = as.numeric(gen), y = gt, group = site), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:5)) +
  labs(x = "Generation", y = "Proportion of allele frequency\nchange due to linked selection")

ggsave(paste("../results/", today, "/linked_selection_mean_components_", s, ".png", sep = ""), height = 5, width = 7)

ggplot(subset(all_gt_curves, site == s), aes(x = as.numeric(gen), y = gt_abs, group = interaction(site, replicate))) +
  geom_point(alpha = 0.1) +
  geom_line(alpha = 0.1) +
  geom_line(data = subset(mean_gts, site == s), aes(x = as.numeric(gen), y = gt_abs, group = site), lwd = 2, alpha = 1) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:5)) +
  labs(x = "Generation", y = "Proportion of allele frequency\nchange due to linked selection")

ggsave(paste("../results/", today, "/linked_selection_mean_components_abs", s, ".png", sep = ""), height = 5, width = 7)
}
```

## block bootstrap procedure to get uncertainty estimates

### create blocks

```{r}
window_size <- 15000

blocks <- gen0 %>%
  group_by(X1) %>%
  mutate(block = (row_number() - 1) %/% window_size + 1,
         window = paste(X1, block, sep = "_"))

length(unique(blocks$window))

hapfire$window <- blocks$window

blocks_gt_curves = NULL

for(site in sites) {
  print(paste("Site:", site))
  hap_site <- hapfire %>% group_by(window) %>% select(starts_with(paste(site, "_", sep = "")))
  for(w in unique(hap_site$window)){
    #print(paste("Window:", w))
    hap_window <- hap_site %>% filter(window == w)
    hap_window <- hap_window[,-1]
    for(exprep in 1:12) {
      print(paste("Rep:", exprep))
      # subset to one site
      hap_rep <- hap_window %>% select(ends_with(paste("_", exprep, sep = "")))
      
      if(ncol(hap_rep) <= 1){
        print("Need at least two increments. Skipping to next replicate...")
        next
      }
      
      # get sample sizes
      rep_n <- sample_sizes[(sample_sizes$site_gen_plot %in% colnames(hap_rep)),]
      rep_n <- rbind(c(0, 1e5), rep_n)
      rep_n <- rep_n %>% arrange(site_gen_plot)
      
      # sort names to be in order of generation
      hap_rep <- hap_rep %>% relocate(all_of(sort(colnames(hap_rep))))
      
      # create matrix of allele frequencies
      # L X G
      pmat <- blocks[(blocks$window == w), "X3"]
      colnames(pmat)[1] <- "0"
      for(i in 1:ncol(hap_rep)){
        pmat <- cbind(pmat, hap_rep[,i])
        colnames(pmat)[i+1] <- colnames(hap_rep)[i]
      }
      
      # calculate rolling sum of covariances
      rep_gt <- gt_n_adjust(pmat, rep_n$n)
      
      # save result
      rep_gt$site <- site
      rep_gt$replicate <- exprep
      rep_gt$window <- w
      rep_gt$S <- nrow(hap_rep)
      rep_gt$gen <- gsub("_", "", str_extract(colnames(hap_rep), "_.*_"))
      blocks_gt_curves <- rbind(blocks_gt_curves, rep_gt)
    }
  }
}

# save results
write.table(blocks_gt_curves, paste(results_path, "block_gt_for_bootstrapping.txt", sep = ""), row.names = F, quote = F, sep = "\t")
```


### make confidence intervals

```{r}

blocks_gt_curves <- fread("../results/2026-04-15/block_gt_for_bootstrapping.txt")

sites <- unique(blocks_gt_curves$site) # make list of sites
B = 250 # number of bootstrap samples per site
W = 200 # number of windows in the genome

cis <- NULL

for(s in sites){
  boot_samples = NULL
  for(b in 1:B){
    boot_sample <- blocks_gt_curves %>%
      filter(site == s) %>%
      nest_by(replicate, window) %>% # Group and nest into a list-column
      ungroup() %>%
      slice_sample(n = W, replace = TRUE) %>%  # Sample groups with replacement
      unnest(data) %>% 
      group_by(site, gen) %>% 
      summarize(
        gt_abs = weighted.mean(sum_abs_cov, S)/weighted.mean(sum_abs_cov + sum_var, S))
    
    boot_samples <- rbind(boot_samples, boot_sample)
  }
  
  ci <- boot_samples %>% group_by(gen) %>% summarize(lwr = quantile(gt_abs, 0.025), med = median(gt_abs), upr = quantile(gt_abs, 0.975))
  ci$site <- s
  
  cis <- rbind(cis, ci)
}

write.table(cis, paste(results_path, "gt_abs_confidence_intervals.txt", sep = ""))

#
cis <- read.table("../results/2026-04-15/gt_abs_confidence_intervals.txt")

cis <- merge(cis, worldclim_sitesdata, by = "site")

jitter_pos <- position_jitter(width = 0.1, h = 0, seed = 123)

cis$gen_jitter <- cis$gen
cis$gen_jitter[(cis$gen >= 2)] <- cis$gen_jitter[(cis$gen >= 2)] + runif(sum(cis$gen >= 2), -0.3, 0.3)
  
ggplot(cis, aes(x = gen_jitter, y = med, group = site, color = bio1)) +
  #geom_point(position = jitter_pos) +
  #geom_line(position = jitter_pos) +
  #geom_errorbar(aes(ymin=lwr, ymax=upr), width=.2, position = jitter_pos) +
  geom_point() +
  geom_line() +
  geom_errorbar(aes(ymin=lwr, ymax=upr), width=.2) +
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:5)) +
  labs(x = "Generation", y = expression(G[abs]), color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste(results_path, "bootstrap_gt_abs_jitter.pdf", sep = ""), width = 5, height = 4)
```

## convergent covariances across replicates

### calculate frequency increments

```{r}
# make list of sites
sites <- unique(gsub("_.*", "", colnames(hapfire)))

# for now, ignore sites that do not have at least 2 incremenent
#sites <- sites[c(-11, -14, -18, -19, -30, -31)]
sites <- sites[-30]

# loop over sites
pdiffs <- NULL
gc()

for(site in sites) {
  print(paste("Site:", site))
  for (i in 1:12) {
    print(paste("Rep:", i))
    hap_rep_i <- hapfire %>% dplyr::select(starts_with(paste(site, "_", sep = "")) & ends_with(paste("_", i, sep = "")))
    # if (ncol(hap_rep_i) <= 1) {
    #   print("Need at least two increments. Skipping to next replicate...")
    #   n_valid_reps <- n_valid_reps - 1
    #   next
    # }
    
    # get sample sizes
    # n_i <- sample_sizes[(sample_sizes$site_gen_plot %in% colnames(hap_rep_i)), ]
    # n_i <- rbind(c(0, 1e5), n_i)
    # n_i <- n_i %>% arrange(site_gen_plot)
    
    # sort names to be in order of generation
    hap_rep_i <- hap_rep_i %>% relocate(all_of(sort(colnames(hap_rep_i))))
    # create matrix of allele frequencies
    # L X G
    pmat_i <- NULL
    pmat_i <- cbind(pmat_i, gen0$X3)
    colnames(pmat_i)[1] <- "0"
    pmat_i <- cbind(pmat_i, hap_rep_i)
    #colnames(pmat_i)[2:ncol(pmat_i)] <- colnames(hap_rep_i)
    
    # for (w in 1:ncol(hap_rep_i)) {
    #   pmat_i <- cbind(pmat_i, hap_rep_i[, w])
    #   colnames(pmat_i)[w + 1] <- colnames(hap_rep_i)[w]
    # }
    
    # calculate increments
    pdiff_i <- as.matrix(pmat_i[, -1] - pmat_i[, -ncol(pmat_i)])
    colnames(pdiff_i) <- colnames(hap_rep_i)
    pdiffs <- cbind(pdiffs, pdiff_i)
  }
}

# save increment matrix
write.csv(pdiffs, "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/allele_frequency_increments.csv", row.names = F, quote = F)

#fwrite(pdiffs, "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/allele_frequency_increments.csv", row.names = F, quote = F)
  
```

### calculate covariances

```{r}
if(!exists("pdiffs")){
  pdiffs <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/allele_frequency_increments.csv", header = T)
}

# calculate all covariances
# system.time(
#  cov(pdiffs[,1:50])
# )
# 
# system.time(
#  cov(as.matrix(pdiffs[,1:10]))
# )

# save un-corrected covariance matrices
cov_pdiffs <- cov(pdiffs)

write.table(cov_pdiffs, "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/increment_covariances.tsv",row.names = T, col.names = T, quote = F, sep = "\t")
```

### correct covariances for sample size
```{r}
# load covariance matrix as matrix object
if(!exists("cov_pdiffs")){
  cov_pdiffs <- as.matrix(read.table("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/raw_increment_covariances.txt", header = TRUE, row.names = 1))
  colnames(cov_pdiffs) <- gsub("X", "", colnames(cov_pdiffs))
}

# make list of sites
sites <- unique(gsub("_.*", "", colnames(hapfire)))

# for now, ignore sites that lack reps with contiguous generations
#sites <- sites[c(1,3,5:17,19:21,23,25:29,31)]
sites <- sites[-30]

# rows and replicates
rows_cov_pdiffs <- rownames(cov_pdiffs)
cols_cov_pdiffs <- colnames(cov_pdiffs)

# create new matrix to store bias-corrected cov matrices
unbiased_cov_pdiffs <- cov_pdiffs

# loop over sites
for(site in sites) {
  print(paste("Site:", site))
  for(exprep in 1:12) {
    print(paste("Rep:", exprep))
    # subset to one site
    hap_rep <- hapfire %>% select(starts_with(paste(site, "_", sep = "")) & ends_with(paste("_", exprep, sep = "")))
    
    if(ncol(hap_rep) <= 1){
      print("Need at least two increments. Skipping to next replicate...")
      next
    }
    
    # get sample sizes
    rep_n <- sample_sizes[(sample_sizes$site_gen_plot %in% colnames(hap_rep)),]
    rep_n <- rbind(c(0, 1e5), rep_n)
    rep_n <- rep_n %>% arrange(site_gen_plot)
    
    # sort names to be in order of generation
    hap_rep <- hap_rep %>% relocate(all_of(sort(colnames(hap_rep))))
    
    # create matrix of allele frequencies
    # L X G
    pmat <- NULL
    pmat <- cbind(pmat, gen0$X3)
    colnames(pmat)[1] <- "0"
    for(i in 1:ncol(hap_rep)){
      pmat <- cbind(pmat, hap_rep[,i])
      colnames(pmat)[i+1] <- colnames(hap_rep)[i]
    }
    
    # subset covariance matrix
    covmat <- cov_pdiffs[grep(paste("^", site, "_.*_", exprep, "$", sep = ""), rows_cov_pdiffs), grep(paste("^", site, "_.*_", exprep, "$", sep = ""), cols_cov_pdiffs)]
    
    # correct covariance matrix for sample size
    correct_covmat <- correct_covmat_for_n(covmat, pmat, rep_n$n)
    
    unbiased_cov_pdiffs[grep(paste("^", site, "_.*_", exprep, "$", sep = ""), rows_cov_pdiffs), grep(paste("^", site, "_.*_", exprep, "$", sep = ""), cols_cov_pdiffs)] <- correct_covmat
  }
}

# check that correction behaved as expected
boolean_cov_pdiffs <- (unbiased_cov_pdiffs > cov_pdiffs)

# save corrected covariance matrix
write.table(cov_pdiffs, "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/unbiased_increment_covariances.tsv", row.names = T, col.names = T, quote = F, sep = "\t")
```

### plot covariances by site

````{r}

for(site in sites){
  
  site_pattern <- paste("^", site, "_", sep = "")
  
  melt_cov <- reshape::melt(unbiased_cov_pdiffs[grep(site_pattern, rownames(unbiased_cov_pdiffs)), grep(site_pattern, colnames(unbiased_cov_pdiffs))])
  
  melt_cov <- melt_cov[(melt_cov$X1 != melt_cov$X2),]
  
  melt_cov <- melt_cov[!duplicated(melt_cov$value),]
  
  melt_cov$tmin <- pmin(as.numeric(gsub("_", "", str_extract(as.character(melt_cov$X1), "_.*_"))),as.numeric(gsub("_", "", str_extract(as.character(melt_cov$X2), "_.*_"))))
  
  melt_cov$tmax <- pmax(as.numeric(gsub("_", "", str_extract(as.character(melt_cov$X1), "_.*_"))),as.numeric(gsub("_", "", str_extract(as.character(melt_cov$X2), "_.*_"))))
  
  melt_cov <- melt_cov[gsub(".*_", "", melt_cov$X1) == gsub(".*_", "", melt_cov$X2),]
  
  melt_cov$rep <- gsub(".*_", "", melt_cov$X1)
  
  ggplot(melt_cov, aes(x = as.factor(tmax), y = value, group = interaction(tmin, rep), color = as.factor(tmin))) +
    geom_point() +
    scale_color_scico_d(palette = "turku", begin = 0.1, end = 0.9) +
    geom_line() +
    theme(legend.position = "top") +
    labs(x = "t", y = "Cov(\u0394ps, \u0394pt)", color = "s", title = paste("Site #", site, sep = ""))
  
  ggsave(paste(results_path, "genome_wide_covariances_site_", site, ".pdf", sep = ""), device = cairo_pdf)
}

````

### block covariances across all replicates

```{r}
pdiffs <- as.data.frame(pdiffs)
if(!exists("pdiffs")){
  pdiffs <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/allele_frequency_increments.csv", header = T)
}

# if(!exists("cov_pdiffs")){
#   cov_pdiffs <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/unbiased_increment_covariances.tsv", header = T, row.names = 1))
# 
#   colnames(cov_pdiffs) <- gsub("X", "", colnames(cov_pdiffs))
# }

# number of snps per window
window_size <- 15000

# create windows, should not bleed between chromosomes
blocks <- gen0 %>%
  group_by(X1) %>%
  mutate(block = (row_number() - 1) %/% window_size + 1,
         window = paste(X1, block, sep = "_"))

length(unique(blocks$window))

pdiffs$window <- blocks$window

conv_gt_results <- NULL

for(w in unique(pdiffs$window)){
  print(paste("Window:", w))
  pdiff_w <- pdiffs %>% filter(window == w) %>% select(-window)
  
  # increment covariances
  cov_pdiffs <- cov(pdiff_w, use = "pairwise.complete.obs")
  
  # covariances within a rep
  row_rep <- gsub("_.*_", "", rownames(cov_pdiffs))
  col_rep <- gsub("_.*_", "", colnames(cov_pdiffs))
  
  row_site <- gsub("_.*", "", rownames(cov_pdiffs))
  col_site <- gsub("_.*", "", colnames(cov_pdiffs))
  
  row_gen <- gsub("_", "", str_extract(rownames(cov_pdiffs), "_.*_"))
  col_gen <- gsub("_", "", str_extract(colnames(cov_pdiffs), "_.*_"))
  
  total_var <- 0
  nreps <- 0
  for(k in unique(row_rep)){
    cov_k <- cov_pdiffs[k == row_rep, k == col_rep]
    # TO DO: correct covariances within reps by sample size
    total_var <- total_var + sum(cov_k)
    nreps <- nreps + 1
  }
  
  # covariances across reps
  total_covar <- 0
  for(k in unique(row_site)){
    for(g in unique(row_gen)){
      cov_k <- cov_pdiffs[(k == row_site) & (g == row_gen), (k != col_site) & (g != col_gen)]
      total_covar <- total_covar + sum(cov_k)
    }
  }
  
  # save results
  # npairs is actually number of pairs*2 because I'm double-counting covariances above
  npairs = ((length(unique(row_site))*length(unique(row_gen)))^2 - length(unique(row_site))*length(unique(row_gen)))
    
  conv_gt_result = data.frame(window = w,
                      S = nrow(pdiff_w),
                      nreps = nreps,
                      npairs = npairs,
                      total_covar = total_covar,
                      total_var = total_var,
                      mean_covar = total_covar/npairs,
                      mean_var = total_var/nreps)
  
  conv_gt_results <-  rbind(conv_gt_results, conv_gt_result)
}

# save results
write.table(conv_gt_results, paste(results_path, "block_conv_gt_for_bootstrapping.txt", sep = ""), row.names = F, quote = F, sep = "\t")

```

### make confidence intervals

```{r}
B = 1000 # number of bootstrap samples per site
W = 200 # number of windows in the genome

#cis <- NULL

boot_samples = NULL
for(b in 1:B){
  boot_sample <- conv_gt_results %>%
    slice_sample(n = W, replace = TRUE) %>%  # Sample groups with replacement
    summarize(
      gt_shared = weighted.mean(mean_covar, S)/weighted.mean(mean_var, S)
    )
  boot_samples <- rbind(boot_samples, boot_sample)
}

boot_samples %>% summarize(lwr = quantile(gt_shared, 0.025), med = median(gt_shared), upr = quantile(gt_shared, 0.975))

```

### covariances across pairs of environments

```{r}
if(!exists("cov_pdiffs")){
  cov_pdiffs <- as.matrix(read.table("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/unbiased_increment_covariances.tsv", header = TRUE, row.names = 1))
  colnames(cov_pdiffs) <- gsub("X", "", colnames(cov_pdiffs))
}

sites <- unique(gsub("_.*", "", colnames(cov_pdiffs)))

row_site <- gsub("_.*", "", rownames(cov_pdiffs))
col_site <- gsub("_.*", "", colnames(cov_pdiffs))

row_rep <- gsub("_.*_", "", rownames(cov_pdiffs))
col_rep <- gsub("_.*_", "", colnames(cov_pdiffs))

pair_gt_results = NULL

for(i in 1:(length(sites) - 1)){
  a = sites[i]
  for(j in (i+1):length(sites)){
    
    b = sites[j]
    
    print(paste("Sites:", i, j))
    
    # subset cov to site pair
    cov_pair <- cov_pdiffs[row_site == a, col_site == b]
    
    # remove overlapping increments
    sub_row_rep <- gsub("_.*_", "", rownames(cov_pair))
    sub_col_rep <- gsub("_.*_", "", colnames(cov_pair))
  
    sub_row_site <- gsub("_.*", "", rownames(cov_pair))
    sub_col_site <- gsub("_.*", "", colnames(cov_pair))
  
    sub_row_gen <- gsub("_", "", str_extract(rownames(cov_pair), "_.*_"))
    sub_col_gen <- gsub("_", "", str_extract(colnames(cov_pair), "_.*_"))
    
    # covariances across reps, non-overlapping increments
    total_covar <- 0
    for(k in colnames(cov_pair)){
      col_gen <- gsub("_", "", str_extract(k, "_.*_"))
      cov_k <- cov_pair[sub_row_gen != col_gen, k]
      total_covar <- total_covar + 2*sum(cov_k)
    }
    
    a_b_reps <- unique(c(sub_row_rep, sub_col_rep))
    total_var <- 0
    nreps <- 0
    for(k in a_b_reps){
      cov_k <- cov_pdiffs[k == row_rep, k == col_rep]
      total_var <- total_var + sum(cov_k)
      nreps <- nreps + 1
    }
    
    npairs = length(unique(sub_row_rep))*length(unique(sub_col_rep))
    
    pair_gt_result = data.frame(site_a = a,
                                site_b = b,
                      nreps = nreps,
                      npairs = npairs,
                      total_covar = total_covar,
                      total_var = total_var,
                      mean_covar = total_covar/npairs,
                      mean_var = total_var/nreps
                      )
    
    pair_gt_results = rbind(pair_gt_results, pair_gt_result)
    
  }
}

pair_gt_results$ge_t <- pair_gt_results$mean_covar/pair_gt_results$mean_var

pair_gt_results_clim <- merge(pair_gt_results, worldclim_sitesdata, by.x = "site_a", by.y = "site")

pair_gt_results_clim <- merge(pair_gt_results_clim, worldclim_sitesdata, by.x = "site_b", by.y = "site", suffixes = c("_a", "_b"))

pair_gt_results_clim$delta_bio1 <- (pair_gt_results_clim$bio1_a - pair_gt_results_clim$bio1_b)

ggplot(pair_gt_results_clim, aes(x = delta_bio1, y = ge_t)) +
  geom_point()
```

### sign permute allele frequency changes in blocks

```{r}

if(!exists("pdiffs")){
  pdiffs <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/allele_frequency_increments.csv", header = T)

  #pdiffs <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/allele_frequency_increments.csv", header = T, select = c("1_1_1", "1_2_1", "1_3_1", "1_1_2", "1_2_2", "1_3_2", "1_1_3", "1_2_3", "1_3_3", "1_1_4", "1_2_4", "1_3_4"))
}

sign_permute_replicates <- 200

sign_permute_blocks <- create_blocks(gen0, X1, X2, 15000, "_")

pdiffs$window = sign_permute_blocks$window

rm(sign_permute_blocks)
gc()

# make list of sites
sites <- unique(gsub("_.*", "", colnames(pdiffs)))
sites <- sites[c(-31)]

#
output_gt_perm <- paste(results_path, "gt_abs_permutations.txt", sep = "")

# create permutations of allele frequency changes by site
mean_gt_perms = NULL
# subset to one site
for (site in sites[26:30]) {
  print(paste("Site:", site))
  pdiffs_site <- pdiffs %>%
    dplyr::select(starts_with(paste(site, "_", sep = "")), window)
  # now permute many times
  for (p in 1:sign_permute_replicates) {
    # subset to one rep
    print(paste("Replicate:", p))
    # choose a random replicate
    # within a block, randomly multiply allele frequency change by +1 or -1
    pdiffs_perm <- pdiffs_site %>%
      group_by(window) %>%
      mutate(across(where(is.numeric), ~ .x * sample(c(1, -1), size = 1))) %>%
      ungroup() %>%
      dplyr::select(-window)
    
    gt_perms = NULL
    # loop over each replicate
    for (exprep in 1:12) {
      pdiffs_perm_rep <- pdiffs_perm %>%
        dplyr::select(ends_with(paste("_", exprep, sep = "")))
      
      if (ncol(pdiffs_perm_rep) <= 1) {
        #print("Need at least two increments. Skipping to next replicate...")
        next
      }
      
      # un-corrected covariance matrices
      cov_pdiffs_perm <- cov(pdiffs_perm_rep)
      
      # TO-DO: correct covariance matrices
      
      # calculate covariance rolling sums
      gt_perm <- gt_from_covmat(cov_pdiffs_perm)
      
      # save result
      gt_perm$rep = exprep
      gt_perms <- rbind(gt_perms, gt_perm)
    }
    # average across replicates to get one permutation sample
    if(is.null(gt_perms)){
      next
    }
    
    mean_gt_perm = gt_perms %>%
      group_by(gen) %>%
      summarize(
        gt = mean(sum_cov) / mean(sum_cov + sum_var),
        gt_abs = mean(sum_abs_cov) / mean(sum_abs_cov + sum_var),
        sum_cov = mean(sum_cov),
        sum_abs_cov = mean(sum_abs_cov),
        sum_pos_cov = mean(sum_pos_cov),
        sum_neg_cov = mean(sum_neg_cov),
        sum_var = mean(sum_var),
        ratio = mean(ratio, na.rm = T),
        n = n()
      )
    
    mean_gt_perm$perm = p
    mean_gt_perm$site = site
    
    write.table(mean_gt_perm, output_gt_perm, col.names=!file.exists(output_gt_perm), append = T, row.names = F, sep = ",", quote = F)
    #mean_gt_perms = rbind(mean_gt_perms, mean_gt_perm)
  }
}

```

### calculate permutation null distributions

```{r}

mean_gt_perms <- fread(paste(results_path, "gt_abs_permutations.txt", sep = ""), header = T)

# remove extra permutations
# sometimes happened, if R needed to be restarted from checkpoint
mean_gt_perms <- mean_gt_perms %>% 
  distinct(site, gen, perm, .keep_all = TRUE)

# calculate percentiles of permutation distribution
perm_int <- mean_gt_perms %>% 
  group_by(gen, site) %>% 
  summarize(lwr = quantile(gt_abs, 0.025), med = median(gt_abs), upr = quantile(gt_abs, 0.975)) %>%
  arrange(site)

write.table(perm_int, paste(results_path, "gt_abs_permutation_intervals_200.txt", sep = ""), row.names = F, quote = F, sep = "\t")
```

# Estimate linked selection parameters

## load, format, and filter recombination map

```{r}
#
library(MareyMap)
data("Arabidopsis_thaliana")

recom_map_1 <- data.frame(
  chrom = 1,
  pos = Arabidopsis_thaliana@maps$`Chromosome 1`@physicalPositions,
  cm = Arabidopsis_thaliana@maps$`Chromosome 1`@geneticDistances
)

recom_map_1 <- recom_map_1 %>% filter(cm == cummax(cm))

recom_map_2 <- data.frame(
  chrom = 2,
  pos = Arabidopsis_thaliana@maps$`Chromosome 2`@physicalPositions,
  cm = Arabidopsis_thaliana@maps$`Chromosome 2`@geneticDistances
)

recom_map_2$cm <- recom_map_2$cm - min(recom_map_2$cm)

recom_map_2 <- recom_map_2 %>% filter(cm == cummax(cm))

recom_map_3 <- data.frame(
  chrom = 3,
  pos = Arabidopsis_thaliana@maps$`Chromosome 3`@physicalPositions,
  cm = Arabidopsis_thaliana@maps$`Chromosome 3`@geneticDistances
)

recom_map_3$cm <- recom_map_3$cm - min(recom_map_3$cm)

recom_map_3 <- recom_map_3 %>% filter(cm == cummax(cm))

recom_map_4 <- data.frame(
  chrom = 4,
  pos = Arabidopsis_thaliana@maps$`Chromosome 4`@physicalPositions,
  cm = Arabidopsis_thaliana@maps$`Chromosome 4`@geneticDistances
)

recom_map_4$cm <- recom_map_4$cm - min(recom_map_4$cm)

recom_map_4 <- recom_map_4 %>% filter(cm == cummax(cm))

# 5
recom_map_5 <- data.frame(
  chrom = 5,
  pos = Arabidopsis_thaliana@maps$`Chromosome 5`@physicalPositions,
  cm = Arabidopsis_thaliana@maps$`Chromosome 5`@geneticDistances
)

recom_map_5$cm <- recom_map_5$cm - min(recom_map_5$cm)

recom_map_5 <- recom_map_5 %>% filter(cm == cummax(cm))

recom_map <- rbind(recom_map_1, recom_map_2, recom_map_3, recom_map_4, recom_map_5)

ggplot(recom_map, aes(x = pos, y = cm)) +
  geom_point() +
  facet_wrap(chrom~., scales = "free") +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(x = "Position", y = "cM")
```

## calculate average ld or average ld decay by block
```{r}

# output file
#output_mean_ld <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/", today, "/mean_ld_by_block.tsv", sep = "")
output_mean_ld <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-05-05/mean_ld_by_block.tsv", sep = "")

# read ld_matrix in blocks and calculate average initial R2
ld_matrix <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/plink.ld.tsv")

# get correlations that are encompassed within each block
blocks <- create_blocks(gen0, X1, X2, 15000, "_")

blocks <- blocks %>% 
  group_by(window) %>% 
  summarize(chr = unique(X1), start = min(X2), end = max(X2))

for(i in 99:nrow(blocks)){
  print(i)
  print(Sys.time())
  
  b = blocks[i,]
  start = b$start
  end = b$end
  chr = b$chr
  w = b$window
  
  result <- data.frame(window = w,
                       mean_ld = ld_matrix %>% 
                         filter((CHR_A == chr) & (CHR_B == chr) & (BP_A >= start) & (BP_A <= end) & (BP_B >= start) & (BP_B <= end)) %>% summarize(mean(R2))
  )
  
  write.table(result, output_mean_ld, col.names=!file.exists(output_mean_ld), append = T, row.names = F, sep = ",", quote = F)
    
}

# 
# ld_matrix %>% summarize(mean(R2))

# process_ld <- function(output){
#   function(x, pos) {
#     print(pos)
#     print(Sys.time())
#     y <- x %>% summarize(chr = unique(CHR_A),
#                     start = min(c(BP_A, BP_B)),
#                     end = max(c(BP_A, BP_B)),
#                     avg_r2 = mean(R2),
#                     n = length(unique(c(SNP_A, SNP_B))))
#     
#     write.table(y, output, col.names=!file.exists(output), append = T, row.names = F, sep = "\t", quote = F)
#   }
# }
# 
# mean_ld <- read_tsv_chunked(
#   "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/plink.ld.tsv",
#   callback = SideEffectChunkCallback$new(process_ld("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/data/plink.ld.mean.tsv")),
#   chunk_size = 99*15000)
```

## use least squares to estimate Ne and Va(1)

Need:
* allele frequencies
* sample sizes
* mean LD decay in windows through time, or just LD in initial generation if assuming no decay through time

```{r}

# interpolation of cM position of SNP based on recombination map
# cm_interpolation <- function(chr, pos, ref){
#   
#   ref <- ref %>% filter(CHR == chr)
#   
#   # find two closest reference points to query point
#   
#   # if query falls before first reference point or last reference point, we can't interpolate
#   
#   # interpolate query point position
#   
# }

#estim_linked_selection_params(cbind(gen0$X3,p), c(1000,292,416,48,279))

# load mean ld
mean_ld_blocks <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-05-05/mean_ld_by_block.tsv")

# assign blocks to allele frequencies
# number of snps per window
window_size <- 15000

# create windows, should not bleed between chromosomes
blocks <- gen0 %>%
  group_by(X1) %>%
  mutate(block = (row_number() - 1) %/% window_size + 1,
         window = paste(X1, block, sep = "_"))

length(unique(blocks$window))

hapfire$window <- blocks$window

# make list of sites
sites <- unique(gsub("_.*", "", colnames(hapfire)))
sites <- sites[c(-30, -32)]

linked_params <- NULL

output_linked_params <- paste("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/", today, "/va_ne_points_for_estimation.txt", sep = "")

# loop over genomic blocks
num_blocks <- length(unique(hapfire$window))

for(b in unique(hapfire$window)){
  print(paste("Window:", b))
  hap_block <- hapfire %>% filter(window == b)
  for(site in unique(sites)){
    print(paste("Site:", site))
    hap_site <- hap_block %>% dplyr::select(starts_with(paste(site, "_", sep = "")))
    gen0_block <- blocks %>% filter(window == b)
    
    mean_ld_block <- mean_ld_blocks %>% filter(window == b)
    
    for(exprep in 1:12){
      hap_rep <- hap_site %>% dplyr::select(ends_with(paste("_", exprep, sep = "")))
      if(ncol(hap_rep) <= 1){
        print("Need at least two increments. Skipping to next replicate...")
        next
      }
      
      # get sample sizes
      rep_n <- sample_sizes[(sample_sizes$site_gen_plot %in% colnames(hap_rep)),]
      rep_n <- rbind(c(0, 1e3), rep_n)
      rep_n <- rep_n %>% arrange(site_gen_plot)
      
      # sort names to be in order of generation
      hap_rep <- hap_rep %>% relocate(all_of(sort(colnames(hap_rep))))
      
      hap_rep <- cbind(gen0_block$X3, hap_rep)
      
      names(hap_rep)[1] = 0
      
      # create table of linked selection covariates for linear model
      linked_param_block <- estim_linked_selection_params(hap_rep, rep_n$n, mean_ld_block$mean.R2., 1/num_blocks)
      
      
      linked_param_block$site = site
      linked_param_block$rep = exprep
      linked_param_block$block = b
      linked_param_block$S = nrow(hap_rep)
      
      write.table(linked_param_block, output_linked_params, col.names=!file.exists(output_linked_params), append = T, row.names = F, sep = ",", quote = F)
      
      #linked_params <- rbind(linked_params, linked_param_block)
    }
  }
}





```

## fit linear model by location

```{r}

va_ne_blocks <- fread("/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/results/2026-05-06/va_ne_points_for_estimation.txt")

# bootstrap estimates
B = 500 # number of bootstrap samples per site
W = 200 # number of windows in the genome
alpha = 0.05 # ci width

cis <- NULL

for(focal_site in unique(va_ne_blocks$site)){
  print(focal_site)
  boot_samples = NULL
  for(b in 1:B){
    boot_sample <- va_ne_blocks %>%
      filter(site == focal_site) %>%
      nest_by(rep, block) %>% # Group and nest into a list-column
      ungroup() %>%
      slice_sample(n = W, replace = TRUE) %>%  # Sample groups with replacement
      #unnest(data) %>% 
      # group_by(site) %>%
      #nest() %>%
      mutate(
        # Fit a linear model to each nested group
        fit = map(data, ~ lm(abs(value) ~ a + b - 1, data = .x)),
        # Extract tidy coefficients (intercept and slope)
        tidied = map(fit, tidy)
      ) %>%
      # Expand the tidied results into a standard data frame
      unnest(tidied) %>% 
      dplyr::select(c(-data, -fit))
    
    boot_samples <- rbind(boot_samples, boot_sample)
  }
  
  ci <- boot_samples %>% 
    group_by(term) %>% 
    summarize(lwr = quantile(estimate, alpha/2), 
              med = median(estimate), 
              upr = quantile(estimate, 1-alpha/2))
  
  ci$site <- focal_site
  
  cis <- rbind(cis, ci)
}

# convert to Ne
avg_boot_ne <- cis %>%
  subset(term == "b") %>%
  mutate(Ne_upr=1/(2*lwr),
         Ne_med=1/(2*med),
         Ne_lwr=1/(2*upr))

ggplot(avg_boot_ne, aes(x = reorder(site, -Ne_med), y = Ne_med, ymin = Ne_lwr, ymax = Ne_upr)) +
  geom_pointrange() +
  scale_y_log10() +
  labs(y = "Ne", x = "Site")

ggsave(paste(results_path, "Ne_estimates_500_perms_abs_cov.pdf"), width = 8, height = 4)

ggplot(subset(cis, term == "a"), aes(x = reorder(site, -med/15000), y = med, ymin = lwr/15000, ymax = upr/15000)) +
  geom_pointrange() +
  labs(y = "Ne", x = "Site")

#### POINT ESTIMATES ####
# fit linear model for each site and replicate
# can look at only absolute value of covariances
va_ne_models <- va_ne_blocks %>%
  group_by(site, rep) %>%
  nest() %>%
  mutate(
    # Fit a linear model to each nested group
    fit = map(data, ~ lm(value ~ a + b - 1, data = .x)),
    # Extract tidy coefficients (intercept and slope)
    tidied = map(fit, tidy)
  ) %>%
  # Expand the tidied results into a standard data frame
  unnest(tidied) %>% 
  dplyr::select(c(-data, -fit))

va_ne_site_means <- va_ne_models %>% 
  group_by(site, term) %>% 
  summarize(mean_est = mean(estimate), sd_est = sd(estimate), n = n(), se_est = sd_est/n)

ggplot(va_ne_models, aes(x = estimate)) +
  geom_histogram() +
  facet_grid(.~term, scales = "free")

avg_ne <- va_ne_models %>% 
  subset(term == "b") %>% 
  mutate(Ne=1/(2*estimate)) %>% 
  group_by(site) %>%
  summarize(sd = sd(Ne), n = n(), se = sd/n, Ne=mean(Ne)) %>%
  arrange(Ne)


```

# Final figures

## linked selection

```{r}
# permutation-based null distributions
# gabs_null_1 <- fread("../results/2026-05-02/gt_abs_permutation_intervals.txt")
# gabs_null_2 <- fread("../results/2026-05-02/gt_abs_permutation_intervals_2.txt")
# gabs_null <- rbind(gabs_null_1, gabs_null_2)

gabs_null <- fread("../results/2026-05-02/gt_abs_permutation_intervals_200.txt")

# bootstrapping-based confidence intervals
gabs_cis <- read.table("../results/2026-04-15/gt_abs_confidence_intervals.txt")

gabs_cis <- merge(gabs_cis, worldclim_sitesdata, by = "site")
  
ggplot(gabs_cis, aes(x = gen, y = med, group = site, color = bio1)) +
  geom_point() +
  geom_line() +
  geom_errorbar(aes(ymin=lwr, ymax=upr), width=.2) +
  geom_ribbon(data = gabs_null, mapping = aes(x=gen, ymin=lwr, ymax=upr), alpha = 0.2, color = "grey", size = 0.01) + 
  scale_color_scico(palette = "berlin") +
  scale_x_discrete(name ="Generation", limits=c(1:5)) +
  labs(x = "Generation", y = expression(G[abs]), color = "Mean\nAnnual\nTemp.\n(\u00B0C)")

ggsave(paste(results_path, "linked_selection_abs_with_null_200_perms.pdf"), width = 6, height = 4)
```

## shared selection between replicates





# Fst 

## historical vs modern pairings

```{r}
pairings <- read.csv("../data/comb_hist+1001g.prs_geo+gen.csv")

historical <- pairings[, 1:2]

historical$type <- "historical"

colnames(historical) <- gsub("historical_", "", colnames(historical))

modern <- pairings[, 5:6]

modern$type <- "modern"

colnames(modern) <- gsub("modern_", "", colnames(modern))

slim_pairings <- rbind(historical, modern)

slim_pairings$time <- slim_pairings$year - min(slim_pairings$year) + 1

slim_pairings <- slim_pairings %>% group_by(type, time) %>% summarize(n = n()) %>% arrange(time, .by_group = T)

slim_pairings$ID = 1:nrow(slim_pairings)

write.table(slim_pairings, "../config/sampling_times.tsv", quote = F, row.names = F, sep = "\t")

sum(slim_pairings$n)

```


## Fst - historical vs modern

```{r}
cutoff = 0.05

msprime <- vroom(
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/msprime_fst.txt",
  delim = "\t",
  col_names = F
)

msprime$ID <- as.numeric(gsub(".*/|.txt", "", fst$X1))

names(msprime)[2] = "msprime_fst"

gcta <- vroom(
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/workflow/gcta_fst.txt",
  delim = "\t",
  col_names = F
)

gcta$ID <- as.numeric(gsub(".*/|.fst:1", "", gcta$X1))

gcta <- gcta %>% group_by(ID) %>% summarize(gcta_fst = mean(X7), S = n())

gcta_msprime <- merge(msprime, gcta, by = "ID")

# compare gcta and msprime
nei_weir_fst <- ggplot(gcta_msprime, aes(x = msprime_fst, y = gcta_fst)) +
  geom_point(shape = 16) +
  #geom_smooth(method = "lm") +
  geom_abline(slope = 1, intercept = 0, lwd = 1.2, color = "green") +
  labs(x = "Nei 1973 Fst (msprime)", y = "Weir 1996 Fst (gcta)")

# merge with window sizes and calculate p-values
params <- vroom(
  "/global/scratch/users/milesroberts/moi_lab_projects/grenenet-phase2/config/fst_params.tsv",
  delim = "\t"
)

params_fst <- merge(params, gcta_msprime, by = "ID")

params_fst %>% 
  #filter(gcta_fst > 0) %>% 
  group_by(L) %>% 
  summarize(upb = quantile(gcta_fst, probs = 1 - cutoff/2, na.rm = T),
            lwb = quantile(gcta_fst, probs = cutoff / 2, na.rm = T),
            ost1_filter_test = 1 - ecdf(gcta_fst)(0.003892181),
            erecta_filter_test = 1 - ecdf(gcta_fst)(0.032451811),
            ost_unfilter_test = 1 - ecdf(gcta_fst)(0.011700622),
            erecta_unfilter_test = 1 - ecdf(gcta_fst)(0.014078437))

# tact distribution of fst from gcta
fst_density <- ggplot(params_fst, aes(x = gcta_fst)) +
  geom_density() +
  facet_wrap(L~., ncol = 1, labeller = "label_both") +
  geom_vline(data = filter(params_fst, L == 3081), aes(xintercept = 0.003892181), lty = 1) +
  geom_vline(data = filter(params_fst, L == 3081), aes(xintercept = 0.011700622), lty = 2) +
  geom_vline(data = filter(params_fst, L == 6583), aes(xintercept = 0.032451811), lty = 1) +
  geom_vline(data = filter(params_fst, L == 6583), aes(xintercept = 0.014078437), lty = 2) +
  labs(x = "Neutral Weir 1996 Fst\n(Sampling desing, Ne = 20,000)")


ggarrange(fst_density, nei_weir_fst, ncol = 2, labels = c("A", "B"), font.label = list(size = 16, family = "Helvetica"))

ggsave(paste("../results/", today, "/baseline_fst_historical_modern.pdf", sep = ""), height = 4, width = 8, device = cairo_pdf)
```


## load scRNA-seq data

```{r}
global <- readRDS("/global/scratch/users/milesroberts/tmp/GSE226097_global_integration_221009.rds")

meta <- global[[]]

#flower <- readRDS("../data/GSE226097_flower_250805.rds")

#flower[["data"]]
```

## check overlap between GEA and AEGIS

```{r}
lfmm_files <- list.files("lfmm.*", path = "../data", full.names = T)

lfmm <- lapply(lfmm_files, read.csv, header = T)

names(lfmm) <- lfmm_files

lfmm <- lfmm %>% 
  bind_rows(.id = "groups")

lfmm$chrom <- as.numeric(gsub("_.*", "", lfmm$snp_id))
lfmm$pos <- as.numeric(gsub(".*_", "", lfmm$snp_id))

lfmm$clim <- gsub(".*_", "", gsub("_k.*", "", lfmm$groups))

# Chr1:1185512..1188696 - CRY2
# Chr1:24331373..24333999 - FT
# Chr2:11058944..11063324 - ELF3
# Chr2:11531967..11534358 - CAM5
# Chr2:18807505..18811125 - SOC1
# Chr4:9195247..9199840 - PHYD
# Chr4:11000771..11003010 - TSF
# Chr5:3173382..3179448 - FLC
# Chr5:5170162..5173085 - CO
# Chr5:18589438..18591506 - DOG1
buffer = 10000
chroms = c(1,1,2,2,2,4,4,5,5,5)
starts = c(1185512,24331373,11058944,11531967,18807505,9195247,11000771,5170162,3173382,18589438)
ends = c(1188696,24333999,11063324,11534358,18811125,9199840,11003010,5173085,3179448,18591506)
starts = starts - buffer
ends = ends + buffer

lfmm_ft_only <- NULL
for(i in seq_along(chroms)){
  print(i)
  lfmm_ft <- lfmm %>% subset((chrom == chroms[i]) & (pos >= starts[i]) & (pos <= ends[i]) & (significant == "True"))
  lfmm_ft_only <- rbind(lfmm_ft_only, lfmm_ft)
}


head(lfmm_ft_only)
```
